### 2017-02-16 proposal.tex
Added proposal.tex

\documentclass{article}

\usepackage[margin=1in]{geometry}

\begin{document}

It's a journey, not a destination.

\end{document}

### 2017-02-17 proposal.tex
Draft proposal

@@
-\documentclass{article}
+\documentclass[twocolumn,11pt]{article}
 
 \usepackage[margin=1in]{geometry}
+\usepackage{caption}
+
+\title{Journey: A Study of Studying Large Memory Systems}
+\author{Mark Mansi, Suhas Pai, Hasnain Ali Pirzada}
+\date{}
 
@@
 
-It's a journey, not a destination.
+\maketitle
+
+\section{Motivation}
+
+Recent advances in memory technologies have seen the emergence of systems with
+terabytes of memory. Many current systems already have dozens or hundreds of
+gigabytes of memory. Moreover, non-volatile memory technologies, which have
+denser physical structures have made large memories more accessable. Thus, we
+conclude that systems with terabytes of memory will become more prevelent in the
+future.
+
+Among recent advances, HP has been working on a system which they call "The
+Machine", in which hundreds of cores have access to a multi-petabyte shared
+memory pool (https://www.labs.hpe.com/the-machine/the-machine-distribution). Meanwhile, Intel and Micro announced their 3D Xpoint
+non-volatile memory technology, which brings dense, fast, persistent memory into
+the consumer and industrial realms (http://www.intel.com/content/www/us/en/architecture-and-technology/3d-xpoint-unveiled-video.html).
+
+Much research is currently being done to study how non-volatile and large
+memory systems would change system design. However, these studies encounter the
+simple difficulty that most researchers do not have the funds to access these
+sorts of systems for experimentation.
+
+We note that there was a similar problem when large storage systems first became
+available \cite{david, exalt}. Typically, there are two alternatives in
+such a situation: trick the system into believing such hardware is available or
+use a simulation (e.g. via a virtual machine or specialized tool).
+
+Currently, though, there are few good tools avaiable to systems researchers for
+understanding how systems interact with large memories. Running virtual machines
+with large memories is extremely slow and limits the scale of these experiments.
+Tricking the system into believing there is a large memory is more difficult
+than tricking the same system into believing there is a large disk because user
+programs have direct access to memory and tend to use more ad hoc in-memory
+structures.
+
+The ideal end-goal of this research is to find a general lightweight technique
+or tool for studying and designing large-memory systems.
+
+\section{Proposal}
+
+This research will proceed in two phases. First, we will analyze and understand
+the existing memory management mechanisms in Linux. In particular, we will look
+at the data structures in use and the overhead of the various memory management
+mechanisms. This understanding will help us to understand where current memory
+management techniques are unscalable, which in turn will help us understand how
+to build either an emulation tool or simulation tool for studying large memory
+systems using more affordable common-sized-memory systems, such as a typical
+research would have access to.
+
+In exploring the design space, we will explore the limits of virtualization
+techniques with regards to memory AND the ways in which the Linux kernel keeps
+discovers and manages memory. Exploring the scalability of Linux memory
+management allows us to see how well we can cause the Linux kernel to trick
+itself into believing there is a large amount of memory. This is important
+because we may find that this is possible but not efficient.
+
+The second phase of the research will be to develop this general tool. We will
+likely start by implementing a less-general tool which targets low-hanging
+fruit. Then, we will make it more general as time permits. This second phase is
+heavily influenced by what we find in the first phase, so it present our notions
+of what we will do are rather shadowy.
+
+\section{Evaluation of Solution}
+
+In the course of our analysis phase, we will identify a number of workloads
+targetted at large-memory systems. We can use these workloads again to evaluate
+our solution.
+
+\bibliography{references}{}
+\bibliographystyle{plain}
 

### 2017-02-26 proposal.tex
Pushed final proposal

--- 2017-02-16 proposal.tex
@@
 
-It's a journey, not a destination.
+\maketitle
+
+\section{Motivation}
+
+Recent advances in memory technologies are making it possible for machines with
+large amounts of memory to become commonplace. Many current systems already
+have dozens or hundreds of gigabytes of memory. These technology advances can
+enable systems with terabytes of memory in the not very distant future. For
+example, non-volatile memory technologies, which have denser physical
+structures than traditional DRAM are making such large memories available and
+more accessible \cite{xpoint}.
+
+Much research is currently being done to study how non-volatile and large
+memory systems would change system design. HP has been working on a system
+which it calls ``The Machine'', in which hundreds of cores have access to a
+multi-petabyte shared memory pool \cite{hp_machine}. Intel and
+Micron have announced their 3D Xpoint non-volatile memory technology, making
+dense, fast, persistent memory comercially available \cite{xpoint}.
+
+However, studies of large-memory systems are often forced to use simulation or
+emulation to evaluate solutions because most researchers do not have access to
+these sorts of systems for experimentation \cite{quartz}. This often
+forces researchers to simplify models, incurring inaccuracy, or to use smaller
+benchmarks, which are less realistic. We note that there was a similar problem
+when large storage systems first became available \cite{david, exalt}.
+
+Currently, there are few good tools available to systems researchers for
+understanding how systems interact with large memories. The ideal end-goal of
+this research is to identify bottlenecks in managing large memories and find a
+general lightweight technique or tool for studying and designing operating
+systems capable of handling large memory while using systems currently
+available to researchers.
+
+\section{Proposal}
+
+This research will proceed in two phases. First, we will analyze and understand
+the existing memory management mechanisms in Linux, particularly the overhead
+of its data structures and operations. We will explore the limits of
+virtualization techniques with large guest memory and the ways in which the
+Linux kernel manages memory, exploring the scalability of Linux memory
+management. Specifically, we want to look at cases where the guest has many
+times more memory than the host. Part of this phase will involve experimentally
+exploring a variety of workloads with large memory footprints.
+
+In the second phase of the research our goal is to build a tool for studying
+large-memory systems using more affordable common-sized-memory systems, such as
+those to which a typical researcher would have access. Specifically, we will
+target operating systems research. We will likely start by implementing a
+less-general tool which targets one simple application or a class of
+applications. Then, we will make it more general as time permits. The design
+choices in our second phase will be heavily influenced by what we find in the
+first phase when we study the overhead of large memory systems, so at present
+our notions of what we will do are rather vague.
+
+\section{Evaluation}
+
+In the course of our analysis phase, we will identify a number of workloads targeted at large-memory systems.
+We can use these workloads again to evaluate our solution.
+Some workloads that we are currently considering are GUPS and memcached.
+
+\section{Required Resources}
+
+We require a machine with enough storage space and memory to develop and test large-memory system emulation techniques.
+
+We will also need access to benchmarks such as GUPS and memcached for our experiments and evaluation.
+
+Finally, we request a large vat of M\&M’s or Snickers, and a supply of free food for three persons for around three months as a catalyst for our research.
+
+\bibliography{references}{}
+\bibliographystyle{plain}
 

### 2017-03-08 related_work.tex
Pushed final proposal
Added related work

--- 2017-02-16 proposal.tex
@@
 
-It's a journey, not a destination.
+\maketitle
+
+The emergence of new technology has in the past often required operating
+systems to change. Traditionally, OS researchers and developers benchmark and
+test their solutions in new environments to demonstrate their efficacy before
+using them in real systems. However, researchers do not always have access to
+new technology for practical reasons; for example, large memory systems are
+still too expensive for most research budgets. This is not the first time that
+researchers have been forced to study systems in environments they do not have
+access to. This section examines past approaches to studying these systems and
+motivates the need for our work.
+
+The emergence of the internet required businesses to buy expensive machines to
+keep up with demand for their services, but systems researchers did not have
+access to these machines. Alameldeen et al. describe how they simulate a
+multi-million dollar server using a \$2000 workstation. To do this, they go
+through several rounds of scaling down, optimizing, changing, and tuning both
+the benchmarks and systems they test. For example, they scale down a workload
+to fit in the 1GB memory of the machine they use and increase the number of
+threads to improve parallelism (TODO: cite). This methodology may be accurate,
+but it requires modifying the benchmarks, which is error prone and time
+consuming; it can be extremely difficult to validate changes to a benchmark. In
+our work, we seek to avoid changes to the experimental workloads including
+changes to code or scale.
+
+Likewise, the Quartz emulator studies the behaviour of applications on
+Non-volatile Memory (NVM) while actually running on top of DRAM. It uses
+existing hardware capabilities to emulate the higher latency and lower
+bandwidth of most NVM technologies. While Quartz does not address the fact that
+the actual memory available to the applications on NVM might be orders of
+magnitude higher than on DRAM, it demonstrates that new technologies may be
+emulated using existing technology (TODO: cite [2]).
+
+The Simics simulator uses demand paging for simulated environments, which means
+that memory is only allocated when it is used. As long as the working set of
+the target system fits in the host memory, performance will be tolerable.
+However, in our work, we wish to study OS behavior under workloads that
+actually use all available memory. Thus, the usefulness of demand paging is
+greatly reduced. For these workloads, the performance of simulators like Simics
+tends to be impractical (TODO: cite [3]). We do not explore simulation further
+because it can be orders of magnitude slower than native execution, making it
+infeasible for studying large workloads or systems (TODO cite: \$2M paper).
+
+A similar problem was encountered by researchers studying both large storage
+drives and large distributed storage systems. David is a system that allows
+storage and big data researchers to run large benchmarks requiring terabytes of
+storage using off the shelf storage devices (which at that time were too
+small). David creates a compressed version of the file system by physically
+storing only metadata and discarding the contents of files. Reads to the disk
+causes David to generate data on the fly. This decision choice is based on the
+observation that most benchmarking frameworks do not care about the actual
+content of the files, and that most of the storage capacity of a drive tends to
+be data rather than metadata (TODO: cite [1]). The Exalt system uses a similar
+methodology for large-scale distributed storage systems (TODO: cite [4]). This
+methodology provides a promising direction for large-memory system studies. One
+may consider ignoring the contents of a process’s heap and only storing kernel
+data structures and a process’s code and stack segments. However, generating
+heap contents on the fly is more difficult than generating disk contents
+because of the common use of custom data structures. Moreover, on Linux, the
+kernel data structures for 1TB of memory may also exceed the size of physical
+memory on present systems[3].
+
+A virtualized environment can be used to provide a guest system with more
+memory than is physically available to the host. A study of the limits of the
+KVM hypervisor found that there is no fundamental limit to the size of guest
+physical memory other than the hardware address width. However, currently, a
+Linux host running KVM will require the guest memory to be backed by host
+physical memory or host swap space (TODO: cite IBM KVM).  This means that when
+the virtual machine uses the whole amount of memory allocated to it, the host
+can swap pages to disk. The resulting poor performance can cause inaccurate
+performance measurements when running benchmarks.
+
+Prefetching pages from swap space can offer a way to mitigate the overhead of
+memory overcommitment. When there is significant memory pressure, even pages
+which are likely to be accessed soon are swapped out to disk. They are faulted
+back into memory when accessed, resulting in significant performance cost.
+Charm++ uses a programming model where computations can be scheduled by the
+language runtime. A designated thread can prefetch pages required by a
+computation, averting page faults. Although this approach is effective in
+avoiding overhead, it requires applications to be written with a particular
+programming language and so is not applicable to most benchmarks without
+changing them (TODO: cite Charm++ paper).
+
+Gupta et al. built an emulator for high speed networks using a technique called
+time dilation, which slows down the OS clock to make it appear that external
+events are occurring faster. This allows the system to emulate network links
+with speeds that are currently not available. The implementation is based on
+the VMs and Xen hypervisor; Xen delivers the timer interrupts to the guest at a
+lower rate than hardware hence  slowing down the guest’s clock (TODO: cite). We
+may use a similar approach to slow down system time  while paging in and out
+large portions of memory. This will allow us to make the system believe that it
+is reading data from the memory while actually most of that data is being read
+from the disk.
+
+Finally, there have been studies which look at the performance overheads
+associated with current implementations of virtual memory and suggest
+mechanisms to mitigate them. RadixVM tries to overcome performance issues in
+highly concurrent workloads due to serialization of memory management
+operations on kernel data structures (TODO: cite RadixVM). This work
+demonstrates that many parts of existing memory management schemes are not
+scalable to larger systems. Similarly, in our work, we wish to examine
+scalability limitations of memory management in the Linux kernel. Other studies
+have suggested that \texttt{struct page}, \texttt{struct vm\_area\_struct}, and page tables
+tend to comprise a large portion of memory management overhead (TODO: cite
+Simics article).
+
+\bibliography{references}{}
+\bibliographystyle{plain}
 

### 2017-03-14 related_work.tex
Added some code

@@
 
-The emergence of new technology has in the past often required operating
-systems to change. Traditionally, OS researchers and developers benchmark and
-test their solutions in new environments to demonstrate their efficacy before
-using them in real systems. However, researchers do not always have access to
-new technology for practical reasons; for example, large memory systems are
-still too expensive for most research budgets. This is not the first time that
-researchers have been forced to study systems in environments they do not have
-access to. This section examines past approaches to studying these systems and
-motivates the need for our work.
+\section{Related Work}
+
+The emergence of new technology has in the past often required operating systems
+to change. Traditionally, OS researchers and developers benchmark and test their
+solutions in new environments to demonstrate their efficacy before using them in
+real systems. However, researchers do not always have access to new technology
+for practical reasons; for example, large memory systems are still too expensive
+for most research budgets. This is not the first time that researchers have been
+forced to study systems in environments they do not have access to. This section
+examines past approaches to studying these systems and motivates the need for
+our work.
 
@@
 through several rounds of scaling down, optimizing, changing, and tuning both
-the benchmarks and systems they test. For example, they scale down a workload
-to fit in the 1GB memory of the machine they use and increase the number of
-threads to improve parallelism (TODO: cite). This methodology may be accurate,
-but it requires modifying the benchmarks, which is error prone and time
-consuming; it can be extremely difficult to validate changes to a benchmark. In
-our work, we seek to avoid changes to the experimental workloads including
-changes to code or scale.
+the benchmarks and systems they test. For example, they scale down a workload to
+fit in the 1GB memory of the machine they use and increase the number of threads
+to improve parallelism \cite{2kmachine}. This methodology may be accurate, but
+it requires modifying the benchmarks, which is error prone and time consuming;
+it can be extremely difficult to validate changes to a benchmark. In our work,
+we seek to avoid changes to the experimental workloads including changes to code
+or scale.
 
@@
 Non-volatile Memory (NVM) while actually running on top of DRAM. It uses
-existing hardware capabilities to emulate the higher latency and lower
-bandwidth of most NVM technologies. While Quartz does not address the fact that
-the actual memory available to the applications on NVM might be orders of
-magnitude higher than on DRAM, it demonstrates that new technologies may be
-emulated using existing technology (TODO: cite [2]).
+existing hardware capabilities to emulate the higher latency and lower bandwidth
+of most NVM technologies. While Quartz does not address the fact that the actual
+memory available to the applications on NVM might be orders of magnitude higher
+than on DRAM, it demonstrates that new technologies may be emulated using
+existing technology \cite{quartz}.
 
 The Simics simulator uses demand paging for simulated environments, which means
-that memory is only allocated when it is used. As long as the working set of
-the target system fits in the host memory, performance will be tolerable.
-However, in our work, we wish to study OS behavior under workloads that
-actually use all available memory. Thus, the usefulness of demand paging is
-greatly reduced. For these workloads, the performance of simulators like Simics
-tends to be impractical (TODO: cite [3]). We do not explore simulation further
-because it can be orders of magnitude slower than native execution, making it
-infeasible for studying large workloads or systems (TODO cite: \$2M paper).
+that memory is only allocated when it is used. As long as the working set of the
+target system fits in the host memory, performance will be tolerable.  However,
+in our work, we wish to study OS behavior under workloads that actually use all
+available memory. Thus, the usefulness of demand paging is greatly reduced. For
+these workloads, the performance of simulators like Simics tends to be
+impractical \cite{simics}. We do not explore simulation further because it can
+be orders of magnitude slower than native execution, making it infeasible for
+studying large workloads or systems \cite{2kmachine}.
 
@@
 storage and big data researchers to run large benchmarks requiring terabytes of
-storage using off the shelf storage devices (which at that time were too
-small). David creates a compressed version of the file system by physically
-storing only metadata and discarding the contents of files. Reads to the disk
-causes David to generate data on the fly. This decision choice is based on the
-observation that most benchmarking frameworks do not care about the actual
-content of the files, and that most of the storage capacity of a drive tends to
-be data rather than metadata (TODO: cite [1]). The Exalt system uses a similar
-methodology for large-scale distributed storage systems (TODO: cite [4]). This
-methodology provides a promising direction for large-memory system studies. One
-may consider ignoring the contents of a process’s heap and only storing kernel
-data structures and a process’s code and stack segments. However, generating
-heap contents on the fly is more difficult than generating disk contents
-because of the common use of custom data structures. Moreover, on Linux, the
-kernel data structures for 1TB of memory may also exceed the size of physical
-memory on present systems[3].
+storage using off the shelf storage devices (which at that time were too small).
+David creates a compressed version of the file system by physically storing only
+metadata and discarding the contents of files. Reads to the disk causes David to
+generate data on the fly. This decision choice is based on the observation that
+most benchmarking frameworks do not care about the actual content of the files,
+and that most of the storage capacity of a drive tends to be data rather than
+metadata \cite{david}. The Exalt system uses a similar methodology for
+large-scale distributed storage systems \cite{exalt}. This methodology provides
+a promising direction for large-memory system studies. One may consider ignoring
+the contents of a process’s heap and only storing kernel data structures and a
+process’s code and stack segments. However, generating heap contents on the fly
+is more difficult than generating disk contents because of the common use of
+custom data structures. Moreover, on Linux, the kernel data structures for 1TB
+of memory may also exceed the size of physical memory on present systems
+\cite{simics}.
 
-A virtualized environment can be used to provide a guest system with more
-memory than is physically available to the host. A study of the limits of the
-KVM hypervisor found that there is no fundamental limit to the size of guest
+A virtualized environment can be used to provide a guest system with more memory
+than is physically available to the host. A study of the limits of the KVM
+hypervisor found that there is no fundamental limit to the size of guest
 physical memory other than the hardware address width. However, currently, a
 Linux host running KVM will require the guest memory to be backed by host
-physical memory or host swap space (TODO: cite IBM KVM).  This means that when
-the virtual machine uses the whole amount of memory allocated to it, the host
-can swap pages to disk. The resulting poor performance can cause inaccurate
+physical memory or host swap space \cite{ibmkvm}.  This means that when the
+virtual machine uses the whole amount of memory allocated to it, the host can
+swap pages to disk. The resulting poor performance can cause inaccurate
 performance measurements when running benchmarks.
@@
 programming language and so is not applicable to most benchmarks without
-changing them (TODO: cite Charm++ paper).
+changing them \cite{charmpp}.
 
@@
 events are occurring faster. This allows the system to emulate network links
-with speeds that are currently not available. The implementation is based on
-the VMs and Xen hypervisor; Xen delivers the timer interrupts to the guest at a
-lower rate than hardware hence  slowing down the guest’s clock (TODO: cite). We
-may use a similar approach to slow down system time  while paging in and out
+with speeds that are currently not available. The implementation is based on the
+VMs and Xen hypervisor; Xen delivers the timer interrupts to the guest at a
+lower rate than hardware hence  slowing down the guest’s clock \cite{timedil}.
+We may use a similar approach to slow down system time  while paging in and out
 large portions of memory. This will allow us to make the system believe that it
@@
 Finally, there have been studies which look at the performance overheads
-associated with current implementations of virtual memory and suggest
-mechanisms to mitigate them. RadixVM tries to overcome performance issues in
-highly concurrent workloads due to serialization of memory management
-operations on kernel data structures (TODO: cite RadixVM). This work
-demonstrates that many parts of existing memory management schemes are not
-scalable to larger systems. Similarly, in our work, we wish to examine
-scalability limitations of memory management in the Linux kernel. Other studies
-have suggested that \texttt{struct page}, \texttt{struct vm\_area\_struct}, and page tables
-tend to comprise a large portion of memory management overhead (TODO: cite
-Simics article).
+associated with current implementations of virtual memory and suggest mechanisms
+to mitigate them. RadixVM tries to overcome performance issues in highly
+concurrent workloads due to serialization of memory management operations on
+kernel data structures \cite{radixvm}. This work demonstrates that many parts of
+existing memory management schemes are not scalable to larger systems.
+Similarly, in our work, we wish to examine scalability limitations of memory
+management in the Linux kernel. Other studies have suggested that \texttt{struct
+page}, \texttt{struct vm\_area\_struct}, and page tables tend to comprise a
+large portion of memory management overhead \cite{simics}.
 

### 2017-03-14 gen_bar.py

@@ -1,0 +1,46 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib
+
+DATA = 'output/latency_large_mmap.csv'
+
+data = {i: [] for i in range(10,30)}
+
+with open(DATA, 'r') as f:
+    for line in f:
+        if len(line.strip()) == 0:
+            continue
+
+        size = int(line.split(",")[0])
+        lat = float(line.split(",")[1])
+
+        data[size].append(float(lat))
+
+def median(vals):
+    return sorted(vals)[len(vals) / 2]
+
+labels = data.keys()
+xs = [median(data[l]) for l in labels]
+
+print xs
+
+ind = np.arange(len(labels))  # the x locations for the groups
+width = 0.7       # the width of the bars
+
+fig, ax = plt.subplots()
+rects1 = ax.bar(ind + ((1 - width) / 2.0), xs, width, color='k')
+
+# add some text for labels, title and axes ticks
+ax.set_ylabel('Latency (cyc)')
+ax.set_xticks(ind + width / 2 + (1 - width) / 2)
+ax.set_xticklabels(labels)
+
+font = {'family' : 'normal',
+        'weight' : 'bold',
+        'size'   : 22}
+
+matplotlib.rc('font', **font)
+
+plt.grid(True)
+
+plt.show()

### 2017-03-14 latency_line.py

@@ -1,0 +1,36 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib
+
+DATA = 'output/many_mmap.out'
+
+data = []
+
+with open(DATA, 'r') as f:
+    for line in f:
+        if len(line.strip()) == 0:
+            continue
+
+        lat = float(line)
+
+        data.append(lat)
+
+line_pipe, = plt.plot(data)
+
+#plt.xscale("log", basex = 2)
+plt.yscale("log")
+
+#plt.xlim(0, 512+32)
+#plt.ylim(0, 600)
+
+#plt.xlabel("Message Size (bytes)")
+#plt.ylabel("Latency ($\mu$secs)")
+plt.grid(True)
+
+font = {'family' : 'normal',
+        'weight' : 'bold',
+        'size'   : 22}
+
+matplotlib.rc('font', **font)
+
+plt.show()

### 2017-03-14 rdtsc.h

@@ -1,0 +1,11 @@
+#ifndef _RDTSC_H_
+#define _RDTSC_H_
+
+static __inline__ unsigned long long rdtsc(void)
+{
+    unsigned hi, lo;
+    __asm__ __volatile__ ("rdtsc" : "=a"(lo), "=d"(hi));
+    return ( (unsigned long long)lo)|( ((unsigned long long)hi)<<32 );
+}
+
+#endif

### 2017-03-14 run_tests.py

@@ -1,0 +1,60 @@
+#! /usr/bin/python
+
+import csv
+import os
+import subprocess
+import shlex
+
+# Assume that the test executables already exist
+tests = [
+        "large_mmap"
+]
+
+mmap_sizes = [i for i in range(10, 30)]
+
+output_dir = os.path.join(os.getcwd(), "output")
+
+def average(results):
+    return sum(results) * 1./ (len(results))
+
+def run_cmd(cmd):
+    return subprocess.check_output(shlex.split(cmd))
+
+def run_test(test_name, mmap_size, iterations=100):
+    test_script = os.path.join(os.getcwd(), test_name)
+    cmd = "%s %d" % (test_script, mmap_size)
+
+    results = []
+    for _ in xrange(iterations):
+        output = run_cmd(cmd)
+        results += [float(output)]
+    return results
+
+def dump_results(test_name, results):
+    if not os.path.exists(output_dir):
+        os.makedirs(output_dir)
+
+    filename = '%s.csv' % os.path.join(output_dir, test_name)
+    with open(filename, 'wb') as csvfile:
+        writer = csv.writer(csvfile)
+        for row in results:
+            writer.writerow(row)
+
+def run_latency_tests():
+    print "LATENCY RESULTS"
+    for test in tests:
+        total_results = []
+        for mmap_size in mmap_sizes:
+            results = run_test(test, mmap_size)
+
+            print "%20s %7d: %s" % (test, mmap_size, average(results))
+
+            for result in results:
+                total_results += [[mmap_size, result]]
+
+        dump_results("latency_" + test, total_results)
+        print
+    print
+
+if __name__ == "__main__":
+    run_latency_tests()

### 2017-04-01 bench_mem_usage.cc
add mmap_small_remap_large for time, refactor time code

@@ -1,0 +1,51 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <sys/mman.h>
+
+#include "rdmemusg.h"
+
+int main(int argc, char **argv)
+{
+	size_t i;
+	char *addr;
+	unsigned long tmem, fmem, pmem, before, after;
+
+	if (argc < 3) {
+		printf("Usage: ./memusage size_order number_order\n");
+		return -1;
+	}
+
+	size_t size = 1 << atoi(argv[1]);
+	size_t amt = 1 << atoi(argv[2]);
+
+	tmem = meminfo("MemTotal");
+	fmem = meminfo("MemFree");
+	pmem = statusinfo("VmRSS");
+	before = tmem - fmem - pmem;
+	printf("%lu\n", before);
+
+	for (i = 0; i < amt; i++) {
+		addr = (char *)mmap(NULL, size, PROT_READ | PROT_WRITE, MAP_ANON | MAP_PRIVATE, -1, 0);
+		if (addr == MAP_FAILED) {
+			fputs("mmap failed!\n", stderr);
+			return -1;
+		} else {
+	//		addr[0] = 'A';
+		}
+
+	tmem = meminfo("MemTotal");
+	fmem = meminfo("MemFree");
+	pmem = statusinfo("VmRSS");
+	after = tmem - fmem - pmem;
+	printf("%lu\n", after);
+
+	}
+
+	tmem = meminfo("MemTotal");
+	fmem = meminfo("MemFree");
+	pmem = statusinfo("VmRSS");
+	after = tmem - fmem - pmem;
+	printf("%lu\n", after);
+
+	return 0;
+}

### 2017-04-01 bench_ops_per_sec_mmap.cc

@@ -1,0 +1,51 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <sys/mman.h>
+
+#include "pattern.h"
+#include "rdtsc.h"
+#include "pinning.h"
+
+int main(int argc, char* argv[]) {
+    set_cpu(0);
+    PatternGenerator pg = get_mmap_cont(1, 7);
+
+    Next cur;
+    size_t i;
+    char* addr;
+
+    if (argc < 2) {
+        printf("Usage: ./pattern total_time_sec\n");
+        exit(1);
+    }
+
+    int time_to_run = atoi(argv[1]);
+    
+    double ticks_per_nano = get_ticks_per_nano();
+
+
+    unsigned long long total_ticks_todo = (unsigned long long)ticks_per_nano * NANO_SEC_PER_SEC * 
+        (unsigned long long)time_to_run;
+
+
+    unsigned int ops_done = 0;
+    unsigned long long start = rdtsc();
+
+    while (rdtsc() - start < total_ticks_todo) {
+        cur = pg.next();
+        if (cur.operation == OP_MMAP) {  // must be
+            addr = (char *)mmap(cur.address, cur.size, cur.permissions, MAP_ANON | MAP_PRIVATE, -1, 0);
+            if (addr == MAP_FAILED) {
+                printf("mmap failed\n");
+                exit(1);
+            }
+        }
+        ops_done++;
+        addr[0] = 'X'; // touch
+    }
+
+    printf("operations/sec = %d\n", ops_done / time_to_run);
+
+    return 0;
+
+}

### 2017-04-01 bench_sample_test.cc

@@ -1,0 +1,61 @@
+#include <stdio.h>
+#include <stdlib.h>
+
+#include "pattern.h"
+#include "rdmemusg.h"
+
+#include <sys/mman.h>
+
+int main(int argc, char **argv) {
+/*        PatternGenerator p = get_frag_prot_pattern();
+        Next val;
+        size_t i;
+
+	char *addr;
+	unsigned long tmem, fmem, pmem, before, after;
+
+	if (argc < 2) {
+		printf("Usage: ./pattern number_order\n");
+		return -1;
+	}
+
+	size_t amt = 1 << atoi(argv[1]);
+
+	tmem = meminfo("MemTotal");
+	fmem = meminfo("MemFree");
+	pmem = statusinfo("VmRSS");
+	before = tmem - fmem - pmem;
+	printf("%lu\n", before);
+
+        for (i = 0; i < amt; i++) {
+          val = p.next();
+          if (val.operation == OP_MMAP) {
+            addr = (char *)mmap(val.address, val.size, val.permissions, MAP_ANON | MAP_PRIVATE, -1, 0);
+            if (addr == MAP_FAILED) {
+              fputs("mmap failed!\n", stderr);
+              return -1;
+            }
+          } else if (val.operation == OP_MPROT) {
+            int j = mprotect(val.address, val.size, val.permissions);
+            if (j) {
+              fputs("mprotect failed!\n", stderr);
+              return -1;
+            }
+          }
+           		addr[0] = 'A';
+
+          tmem = meminfo("MemTotal");
+          fmem = meminfo("MemFree");
+          pmem = statusinfo("VmRSS");
+          after = tmem - fmem - pmem;
+          printf("%lu\n", after);
+
+        }
+
+	tmem = meminfo("MemTotal");
+	fmem = meminfo("MemFree");
+	pmem = statusinfo("VmRSS");
+	after = tmem - fmem - pmem;
+	printf("%lu\n", after);*/
+  return 0;
+}

### 2017-04-01 bench_time_per_op_mmap.cc
measure time for every operation instead of average

@@ -14,14 +14,15 @@ int main(int argc, char* argv[]) {
     size_t i;
     char* addr;
 
-    if (argc < 2) {
-        printf("Usage: ./pattern total_ops_order\n");
+    if (argc < 3) {
+        printf("Usage: ./pattern total_ops_order touch_0_or_1\n");
         exit(1);
     }
 
     int tot_ops = 1 << atoi(argv[1]);
+    int touch = atoi(argv[2]);
     
-    double ticks_per_nano = get_ticks_per_nano();
+    double ticks_per_nano = get_ticks_per_nano();  // calibrate rdtsc before measuring
 
     unsigned long long start = rdtsc();
     
@@ -29,20 +30,23 @@ int main(int argc, char* argv[]) {
         //  assuming all ops were sucessful
         cur = pg.next();
         if (cur.operation == OP_MMAP) {  // must be
+            unsigned long long cur_start = rdtsc();
             addr = (char *)mmap(cur.address, cur.size, cur.permissions, MAP_ANON | MAP_PRIVATE, -1, 0);
+            unsigned long long cur_end = rdtsc();
             if (addr == MAP_FAILED) {
                 printf("mmap failed\n");
                 exit(1);
             }
+            print_interval(cur_start, cur_end, ticks_per_nano);
         }
-        addr[0] = 'X'; // touch
+        if (touch == 1) addr[0] = 'X'; // touch
     }
 
-    unsigned long long end = rdtsc();
+/*    unsigned long long end = rdtsc();
     unsigned long long interval = end - start;
-    unsigned long long nano_secs = (double) interval / ticks_per_nano;
+    unsigned long long nano_secs = (double) interval / ticks_per_nano;*/
 
-    printf("nsec/op = %llu\n", nano_secs / tot_ops);
+/*    printf("nsec/op = %llu\n", nano_secs / tot_ops);*/
 
     return 0;
 

### 2017-04-01 rdtsc.cc

@@ -1,4 +1,5 @@
 #include "rdtsc.h"
+#include <stdio.h>
 
 
 double get_ticks_per_nano() {
@@ -14,4 +15,12 @@ double get_ticks_per_nano() {
    double ticks_per_nano = (double) total_ticks / (double) interval;
 }
 
+void print_interval(unsigned long long start, unsigned long long end, 
+        double ticks_per_nano) {
+
+    unsigned long long interval = end - start;
+    unsigned long long nano_secs = (double) interval / ticks_per_nano;
+    printf("nsecs/op %llu\n", nano_secs);
+
+}
 

### 2017-04-01 rdtsc.h

@@ -12,6 +12,7 @@ static __inline__ unsigned long long rdtsc(void)
 }
 
 double get_ticks_per_nano();
+void print_interval(unsigned long long, unsigned long long, double);
   
 
 

### 2017-04-02 bench_mem_mmap_4k_rand.cc
Random mmaps

@@ -1,0 +1,70 @@
+#include <iostream>
+#include <stdlib.h>
+#include <sys/mman.h>
+#include <string.h>
+#include <errno.h>
+#include <sys/types.h>
+#include <sys/stat.h>
+#include <fcntl.h>
+#include <unistd.h>
+#include <stdio.h>
+
+#include "rdmemusg.h"
+#include "pattern.h"
+#include "pinning.h"
+
+#define START (0x40000000)
+
+int main(int argc, char **argv) {
+    set_cpu(0);
+
+    if (argc < 2) {
+      std::cout << "Usage: ./memusage number_order" << std::endl;
+      return -1;
+    }
+
+    size_t amt = 1 << atoi(argv[1]);
+
+    // Read randomized offsets from a file
+    FILE *f = fopen("offsets.txt", "r");
+    int *offsets, i;
+
+    if (!f) {
+      fputs("Couldn't open offsets.txt\n", stderr);
+      return -1;
+    }
+
+    offsets = (int *)malloc(sizeof(int) * amt);
+
+    for (i = 0; i < amt; i++) {
+      fscanf(f, "%d", &offsets[i]);
+    }
+
+    if (fclose(f)) {
+      fputs("Couldn't close offsets.txt\n", stderr);
+      free(offsets);
+      return -1;
+    }
+
+
+    std::cout << getkernelmem() << " " 
+              << slabinfo("vm_area_struct") << " " 
+              << meminfo("PageTables") << std::endl;
+
+    for (size_t i = 0; i < amt; i++) {
+        int maddr = START + offsets[i] << 12;
+        char *addr = (char *)mmap((void *)maddr, 4 << 12, 7, MAP_ANON | MAP_PRIVATE, -1, 0);
+        if (addr == MAP_FAILED) {
+            std::cerr << "mmap failed: " << strerror(errno) << std::endl;
+            return -1;
+        }
+
+        std::cout << getkernelmem() << " " 
+                  << slabinfo("vm_area_struct") << " " 
+                  << meminfo("PageTables") << std::endl;
+    }
+
+    free(offsets);
+
+    return 0;
+}

### 2017-04-02 pattern.cc
Fix bug

@@ -238,6 +238,7 @@ PatternGenerator get_mmap_4k_remap_8k_unmap_4k() {
     std::vector<PatternPart> pattern;
     pattern.push_back(first);
     pattern.push_back(second);
+    pattern.push_back(third);
 
     return PatternGenerator(pattern);
 }

### 2017-04-02 pattern.cc
Add frag pattern

@@ -169,6 +169,22 @@ PatternGenerator get_mmap_small_remap_large(int small_size, char permissions, in
     return PatternGenerator(pattern);
 }
 
+// Resize pages to larger size
+PatternGenerator get_mremap_frag(unsigned long offset, int size, char permissions) {
+    PatternPart first;
+    first.size = size; // size * 4KB
+    first.location = size;
+    first.permissions = permissions;
+    first.operation = OP_MREMAP;
+
+    std::vector<PatternPart> pattern;
+    pattern.push_back(first);
+
+    unsigned long start = (unsigned long)STARTING_POINT + (offset << 12);
+
+    return PatternGenerator(pattern, (void *)start);
+}
+
 // mmap a bunch of pages with a certain stride
 // Remap the pages to the unmapped spaces
 PatternGenerator get_mmap_stride_reloc(int size, char permissions) {

### 2017-04-02 rdtsc.cc

@@ -13,6 +13,7 @@ double get_ticks_per_nano() {
    unsigned long long interval = NANO_SEC_PER_SEC * (end.tv_sec - start.tv_sec) 
        + end.tv_nsec - start.tv_nsec;
    double ticks_per_nano = (double) total_ticks / (double) interval;
+   return ticks_per_nano;
 }
 
 void print_interval(unsigned long long start, unsigned long long end, 

### 2017-04-02 bench_many_mmap.cc
Fix errors after makefile update

@@ -22,7 +22,7 @@ int main(int argc, char **argv) {
     size_t number = atol(argv[2]);
 
     // actual
-    for(int i = 0; i < number; i++) {
+    for(size_t i = 0; i < number; i++) {
         unsigned long long start = rdtsc();
 
         char *addr = (char*)mmap(NULL, amt, PROT_READ | PROT_WRITE, MAP_ANON | MAP_PRIVATE, -1, 0);

### 2017-04-02 bench_time_per_op_mmap.cc

@@ -11,7 +11,6 @@ int main(int argc, char* argv[]) {
     PatternGenerator pg = get_mmap_cont(1, 7);
 
     Next cur;
-    size_t i;
     char* addr;
 
     if (argc < 3) {
@@ -24,7 +23,6 @@ int main(int argc, char* argv[]) {
     
     double ticks_per_nano = get_ticks_per_nano();  // calibrate rdtsc before measuring
 
-    unsigned long long start = rdtsc();
     
     for (int i = 0; i < tot_ops; i++) {
         //  assuming all ops were sucessful

### 2017-04-03 bench_mem_control.cc
latency tests for mmap cont and strided

@@ -6,7 +6,7 @@
 
 int main(int argc, char **argv)
 {
-	if (argc < 2) {
+/*	if (argc < 2) {
         std::cout << "Usage: ./memusage number_order" << std::endl;
 		return -1;
 	}
@@ -21,5 +21,5 @@ int main(int argc, char **argv)
         std::cout << getkernelmem() << " " 
                   << slabinfo("vm_area_struct") << " " 
                   << meminfo("PageTables") << std::endl;
-	}
+	}*/
 }

### 2017-04-03 bench_mem_mmap_4k_rand.cc

@@ -27,7 +27,8 @@ int main(int argc, char **argv) {
 
     // Read randomized offsets from a file
     FILE *f = fopen("offsets.txt", "r");
-    int *offsets, i;
+    int *offsets;
+    unsigned int i;
 
     if (!f) {
       fputs("Couldn't open offsets.txt\n", stderr);
@@ -52,7 +53,7 @@ int main(int argc, char **argv) {
               << meminfo("PageTables") << std::endl;
 
     for (size_t i = 0; i < amt; i++) {
-        unsigned long maddr = START + offsets[i] << 12;
+        unsigned long maddr = START + (offsets[i] << 12);
         char *addr = (char *)mmap((void *)maddr, 4 << 12, 7, MAP_ANON | MAP_PRIVATE, -1, 0);
         if (addr == MAP_FAILED) {
             std::cerr << "mmap failed: " << strerror(errno) << std::endl;

### 2017-04-03 bench_ops_per_sec_mmap.cc

@@ -11,7 +11,6 @@ int main(int argc, char* argv[]) {
     PatternGenerator pg = get_mmap_cont(1, 7);
 
     Next cur;
-    size_t i;
     char* addr;
 
     if (argc < 2) {

### 2017-04-03 bench_ops_per_sec_mmap_16k_cont_touch.cc

@@ -13,7 +13,7 @@ int main(int argc, char* argv[]) {
     PatternGenerator pg = get_mmap_cont(4 /* 16KB */, 7 /* RWX */);
 
     Next cur;
-    size_t i;
+    //size_t i;
     char* addr;
 
     if (argc < 2) {

### 2017-04-03 do_latency_measurements.py

@@ -1,0 +1,54 @@
+from subprocess import Popen, PIPE
+import os
+
+def run_mmap_cont():
+    print "cont run 1"
+    os.system("./bench_time_per_op_mmap_cont 20 0 >> time_per_op_mmap_cont_1") 
+    print "cont run 2"
+    os.system("./bench_time_per_op_mmap_cont 20 0 >> time_per_op_mmap_cont_2") 
+    print "cont run 3"
+    os.system("./bench_time_per_op_mmap_cont 20 0 >> time_per_op_mmap_cont_3") 
+    print "cont run 4"
+    os.system("./bench_time_per_op_mmap_cont 20 0 >> time_per_op_mmap_cont_4") 
+    print "cont run 5"
+    os.system("./bench_time_per_op_mmap_cont 20 0 >> time_per_op_mmap_cont_5") 
+
+def run_mmap_strided():
+    print "strided run 1"
+    os.system("./bench_time_per_op_mmap_stride 20 0 >> time_per_op_mmap_stride_1") 
+    print "strided run 2"
+    os.system("./bench_time_per_op_mmap_stride 20 0 >> time_per_op_mmap_stride_2") 
+    print "strided run 3"
+    os.system("./bench_time_per_op_mmap_stride 20 0 >> time_per_op_mmap_stride_3") 
+    print "strided run 4"
+    os.system("./bench_time_per_op_mmap_stride 20 0 >> time_per_op_mmap_stride_4") 
+    print "strided run 5"
+    os.system("./bench_time_per_op_mmap_stride 20 0 >> time_per_op_mmap_stride_5") 
+
+
+
+def main():
+    process = Popen(["make", "clean"], stdout=PIPE)
+    (output, err) = process.communicate()
+    exit_code = process.wait()
+    
+    print output
+
+    print "Compiling Source Files..."
+
+    process = Popen(["make"], stdout=PIPE)
+    (output, err) = process.communicate()
+    exit_code = process.wait()
+
+    if exit_code != 0:
+        print "Compile Failed.."
+        return
+    print "Successfully Compiled..."
+
+    
+    
+
+
+
+if __name__ == "__main__":
+    main()

### 2017-04-03 latency_line.py

@@ -1,0 +1,38 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib
+from sys import argv
+
+DATA = argv[1]
+SCALE = argv[2] if len(argv) == 3 else "log"
+
+data = []
+
+with open(DATA, 'r') as f:
+    for line in f:
+        if len(line.strip()) == 0:
+            continue
+
+        lat = float(line)
+
+        data.append(lat)
+
+line_pipe, = plt.plot(data)
+
+#plt.xscale("log", basex = 2)
+plt.yscale(SCALE)
+
+#plt.xlim(0, 512+32)
+#plt.ylim(0, 600)
+
+#plt.xlabel("Message Size (bytes)")
+#plt.ylabel("Latency ($\mu$secs)")
+plt.grid(True)
+
+font = {'family' : 'normal',
+        'weight' : 'bold',
+        'size'   : 22}
+
+matplotlib.rc('font', **font)
+
+plt.show()

### 2017-04-03 rdtsc.cc

@@ -21,7 +21,7 @@ void print_interval(unsigned long long start, unsigned long long end,
 
     unsigned long long interval = end - start;
     unsigned long long nano_secs = (double) interval / ticks_per_nano;
-    printf("nsecs/op %llu\n", nano_secs);
+    printf("%llu\n", nano_secs);
 
 }
 

### 2017-04-03 bench_large_mmap.cc
Some cleanup

@@ -1,53 +1,0 @@
-#include <stdio.h>
-#include <stdlib.h>
-#include <string.h>
-#include <errno.h>
-#include <stddef.h>
-#include <sys/mman.h>
-
-#include "rdtsc.h"
-
-#define WARMUP 100
-
-int main(int argc, char **argv) {
-    // how much to mmap?
-    if(argc < 2) {
-        printf("Usage: ./large_mmap order\n");
-        return -1;
-    }
-
-    size_t amt = 1L << atoi(argv[1]);
-
-    // TODO: try mprotect
-    // TODO: try with huge pages
-    
-    
-    // warmup
-    for(int i = 0; i < WARMUP; i++) {
-        char *addr = (char*)mmap(NULL, amt, PROT_READ | PROT_WRITE, MAP_ANON | MAP_PRIVATE, -1, 0);
-        if(addr == MAP_FAILED) {
-            printf("mmap failed: %s\n", strerror(errno));
-        }
-    }
-
-    // actual
-    unsigned long long start = rdtsc();
-
-    char *addr = (char*)mmap(NULL, amt, PROT_READ | PROT_WRITE, MAP_ANON | MAP_PRIVATE, -1, 0);
-
-    unsigned long long time = rdtsc() - start;
-
-    if(addr == MAP_FAILED) {
-        printf("mmap failed: %s\n", strerror(errno));
-        return -1;
-    }
-
-    /*
-    // touch all pages
-    for(long i = 0; i < SIZE; i += (4 << 10)) {
-        *(addr + i) = 'A';
-    }
-    */
-
-    printf("%llu\n", time);
-}

### 2017-04-03 bench_many_mmap.cc

@@ -1,46 +1,0 @@
-#include <stdio.h>
-#include <stdlib.h>
-#include <string.h>
-#include <errno.h>
-#include <stddef.h>
-#include <sys/mman.h>
-#include <sys/types.h>
-#include <unistd.h>
-
-#include "rdtsc.h"
-
-int main(int argc, char **argv) {
-    // number and size of each mmap
-    if(argc < 3) {
-        printf("Usage: ./large_mmap order number\n");
-        return -1;
-    }
-
-    printf("%d\n\n", (int)getpid());
-
-    size_t amt = 1L << atoi(argv[1]);
-    size_t number = atol(argv[2]);
-
-    // actual
-    for(size_t i = 0; i < number; i++) {
-        unsigned long long start = rdtsc();
-
-        char *addr = (char*)mmap(NULL, amt, PROT_READ | PROT_WRITE, MAP_ANON | MAP_PRIVATE, -1, 0);
-
-        unsigned long long time = rdtsc() - start;
-
-        if(addr == MAP_FAILED) {
-            printf("mmap failed: %s\n", strerror(errno));
-            return -1;
-        }
-
-        printf("%llu\n", time);
-    }
-
-    /*
-    // touch all pages
-    for(long i = 0; i < SIZE; i += (4 << 10)) {
-        *(addr + i) = 'A';
-    }
-    */
-}

### 2017-04-03 bench_sample_test.cc

@@ -1,61 +1,0 @@
-#include <stdio.h>
-#include <stdlib.h>
-
-#include "pattern.h"
-#include "rdmemusg.h"
-
-#include <sys/mman.h>
-
-int main(int argc, char **argv) {
-/*        PatternGenerator p = get_frag_prot_pattern();
-        Next val;
-        size_t i;
-
-	char *addr;
-	unsigned long tmem, fmem, pmem, before, after;
-
-	if (argc < 2) {
-		printf("Usage: ./pattern number_order\n");
-		return -1;
-	}
-
-	size_t amt = 1 << atoi(argv[1]);
-
-	tmem = meminfo("MemTotal");
-	fmem = meminfo("MemFree");
-	pmem = statusinfo("VmRSS");
-	before = tmem - fmem - pmem;
-	printf("%lu\n", before);
-
-        for (i = 0; i < amt; i++) {
-          val = p.next();
-          if (val.operation == OP_MMAP) {
-            addr = (char *)mmap(val.address, val.size, val.permissions, MAP_ANON | MAP_PRIVATE, -1, 0);
-            if (addr == MAP_FAILED) {
-              fputs("mmap failed!\n", stderr);
-              return -1;
-            }
-          } else if (val.operation == OP_MPROT) {
-            int j = mprotect(val.address, val.size, val.permissions);
-            if (j) {
-              fputs("mprotect failed!\n", stderr);
-              return -1;
-            }
-          }
-           		addr[0] = 'A';
-
-          tmem = meminfo("MemTotal");
-          fmem = meminfo("MemFree");
-          pmem = statusinfo("VmRSS");
-          after = tmem - fmem - pmem;
-          printf("%lu\n", after);
-
-        }
-
-	tmem = meminfo("MemTotal");
-	fmem = meminfo("MemFree");
-	pmem = statusinfo("VmRSS");
-	after = tmem - fmem - pmem;
-	printf("%lu\n", after);*/
-  return 0;
-}

### 2017-04-03 bench_test.cc

@@ -1,16 +1,0 @@
-#include "pattern.h"
-#include <stdio.h>
-
-int main() {
-/*    PatternGenerator p1 = get_frag_prot_pattern();
-
-    p1.next().print();
-    p1.next().print();
-    p1.next().print();
-    p1.next().print();
-    p1.next().print();
-    p1.next().print();
-    p1.next().print();
-    p1.next().print(); */
-    return 0;
-}

### 2017-04-03 latency_line.py

@@ -1,38 +1,0 @@
-import numpy as np
-import matplotlib.pyplot as plt
-import matplotlib
-from sys import argv
-
-DATA = argv[1]
-SCALE = argv[2] if len(argv) == 3 else "log"
-
-data = []
-
-with open(DATA, 'r') as f:
-    for line in f:
-        if len(line.strip()) == 0:
-            continue
-
-        lat = float(line)
-
-        data.append(lat)
-
-line_pipe, = plt.plot(data)
-
-#plt.xscale("log", basex = 2)
-plt.yscale(SCALE)
-
-#plt.xlim(0, 512+32)
-#plt.ylim(0, 600)
-
-#plt.xlabel("Message Size (bytes)")
-#plt.ylabel("Latency ($\mu$secs)")
-plt.grid(True)
-
-font = {'family' : 'normal',
-        'weight' : 'bold',
-        'size'   : 22}
-
-matplotlib.rc('font', **font)
-
-plt.show()

### 2017-04-03 bench_mem_mmap_stride.cc
Minor clean up and added memory stride bmk

@@ -1,0 +1,42 @@
+#include <iostream>
+#include <stdlib.h>
+#include <sys/mman.h>
+#include <string.h>
+#include <errno.h>
+
+#include "rdmemusg.h"
+#include "pattern.h"
+#include "pinning.h"
+
+int main(int argc, char **argv) {
+    set_cpu(0);
+
+	if (argc < 2) {
+        std::cout << "Usage: ./memusage number_order" << std::endl;
+		return -1;
+	}
+
+	size_t amt = 1 << atoi(argv[1]);
+
+    PatternGenerator pg = get_mmap_stride(1 /* 4KB */, 7 /* RWX */, 2 /* 8KB */);
+    Next next;
+
+    std::cout << getkernelmem() << " " 
+              << slabinfo("vm_area_struct") << " " 
+              << meminfo("PageTables") << std::endl;
+
+    for (size_t i = 0; i < amt; i++) {
+        next = pg.next();
+
+        char *addr = (char *)mmap(next.address, next.size, next.permissions, MAP_ANON | MAP_PRIVATE, -1, 0);
+        if (addr == MAP_FAILED) {
+            std::cerr << "mmap failed: " << strerror(errno) << std::endl;
+            return -1;
+        }
+
+        std::cout << getkernelmem() << " " 
+                  << slabinfo("vm_area_struct") << " " 
+                  << meminfo("PageTables") << std::endl;
+    }
+}
+

### 2017-04-03 bench_mem_usage.cc

@@ -1,51 +1,0 @@
-#include <stdio.h>
-#include <stdlib.h>
-#include <sys/mman.h>
-
-#include "rdmemusg.h"
-
-int main(int argc, char **argv)
-{
-	size_t i;
-	char *addr;
-	unsigned long tmem, fmem, pmem, before, after;
-
-	if (argc < 3) {
-		printf("Usage: ./memusage size_order number_order\n");
-		return -1;
-	}
-
-	size_t size = 1 << atoi(argv[1]);
-	size_t amt = 1 << atoi(argv[2]);
-
-	tmem = meminfo("MemTotal");
-	fmem = meminfo("MemFree");
-	pmem = statusinfo("VmRSS");
-	before = tmem - fmem - pmem;
-	printf("%lu\n", before);
-
-	for (i = 0; i < amt; i++) {
-		addr = (char *)mmap(NULL, size, PROT_READ | PROT_WRITE, MAP_ANON | MAP_PRIVATE, -1, 0);
-		if (addr == MAP_FAILED) {
-			fputs("mmap failed!\n", stderr);
-			return -1;
-		} else {
-	//		addr[0] = 'A';
-		}
-
-	tmem = meminfo("MemTotal");
-	fmem = meminfo("MemFree");
-	pmem = statusinfo("VmRSS");
-	after = tmem - fmem - pmem;
-	printf("%lu\n", after);
-
-	}
-
-	tmem = meminfo("MemTotal");
-	fmem = meminfo("MemFree");
-	pmem = statusinfo("VmRSS");
-	after = tmem - fmem - pmem;
-	printf("%lu\n", after);
-
-	return 0;
-}

### 2017-04-03 bench_time_per_op_mmap_4k_cont.cc

@@ -8,7 +8,7 @@
 
 int main(int argc, char* argv[]) {
     set_cpu(0);
-    PatternGenerator pg = get_mmap_cont(1, 7);
+    PatternGenerator pg = get_mmap_cont(1 /* 4KB */, 7 /* RWX */);
 
     Next cur;
     char* addr;
@@ -39,12 +39,6 @@ int main(int argc, char* argv[]) {
         }
         if (touch == 1) addr[0] = 'X'; // touch
     }
-
-/*    unsigned long long end = rdtsc();
-    unsigned long long interval = end - start;
-    unsigned long long nano_secs = (double) interval / ticks_per_nano;*/
-
-/*    printf("nsec/op = %llu\n", nano_secs / tot_ops);*/
 
     return 0;
 

### 2017-04-03 paper.tex

--- 2017-03-14 related_work.tex
@@
 \maketitle
+
+\section{Introduction}
+
+Recent advances in memory technologies are making it possible for machines with
+large amounts of memory to become commonplace. Many current systems already
+have dozens or hundreds of gigabytes of memory. These technology advances can
+enable systems with terabytes of memory in the not very distant future. For
+example, non-volatile memory technologies, which have denser physical
+structures than traditional DRAM are making such large memories available and
+more accessible \cite{xpoint}.
+
+Much research is currently being done to study how non-volatile and large
+memory systems would change system design. HP has been working on a system
+which it calls ``The Machine'', in which hundreds of cores have access to a
+multi-petabyte shared memory pool \cite{hp_machine}. Intel and
+Micron have announced their 3D Xpoint non-volatile memory technology, making
+dense, fast, persistent memory comercially available \cite{xpoint}.
+
+However, studies of large-memory systems are often forced to use simulation or
+emulation to evaluate solutions because most researchers do not have access to
+these sorts of systems for experimentation \cite{quartz}. This often
+forces researchers to simplify models, incurring inaccuracy, or to use smaller
+benchmarks, which are less realistic. We note that there was a similar problem
+when large storage systems first became available \cite{david, exalt}.
+
+Currently, there are few good tools available to systems researchers for
+understanding how systems interact with large memories. The ideal end-goal of
+this research is to identify bottlenecks in managing large memories and find a
+general lightweight technique or tool for studying and designing operating
+systems capable of handling large memory while using systems currently
+available to researchers.
 
@@
 
+\section{Methodology}
+
+Our methodology is broken into two parts. First, we measure the scalability of
+several Linux memory management system calls on a system running natively on
+bare metal. Then, we measure the scalability of Linux virtual machines running
+on the KVM hypervisor.
+
+\subsection{Native Experiments}
+
+\begin{figure}
+\centering
+\begin{tabular}{|l|p{5cm}|} \hline
+OS & Ubuntu 16.04.2 \\ \hline
+Kernel & Linux 4.4.0-70 \\ \hline
+CPU & Intel Xeon E5645, 2.40 GHz, 6 cores/12 threads, 384KB Private L1-I Cache,
+1.536 Unified L2 Cache, 12MB Shared L3 Cache \\ \hline
+Memory & 16GB, 1066MHz \\ \hline
+Storage & 500GB, 7200RPM \\
+\hline
+\end{tabular}
+\caption{Specifications of our test machine.  \label{fig:specs}}
+\end{figure}
+
+We measure the memory overhead and latency of the Linux Memory Management
+system's use of \texttt{vm\_area\_struct}, which represents a region of a
+process's virtual address space. (TODO: what about page tables? struct page?)
+When a program makes a memory management system calls such as \texttt{mmap},
+\texttt{mremap}, \texttt{munmap}, \texttt{mprotect}, \texttt{mlock}, or
+\texttt{madvise}, the kernel is creating, changing, or removing
+\texttt{vm\_area\_struct}s; thus, measuring the scalability of Linux memory
+management requires measuring the overhead of these structs.
+
+We run several different microbenchmarks, each of which has a different memory
+allocation pattern. These benchmarks are designed to stress the memory
+management system in different ways, rather than simulate real workloads. For
+each workload, we measure the latency and increase in memory usage due to each
+operation in the benchmark. Notably, we do not touch the pages that we allocate,
+avoiding both a page fault and the allocation of a back physical page for the
+allocated virtual pages. This allows us to measure just the overhead of memory
+management, as opposed to physical page allocation, swapping, page faults, or
+other overheads.  Each workload contains $2^{20}$ memory management operations.
+
+The specifications of the machine are listed in Figure \ref{fig:specs}. We
+disabled Intel SpeedStep and TurboBoost. All experiments are pinned to the same
+core on our machine. Each benchmark was run 5 times and the results aggregated
+as described below. We reboot the machine when switching to a different
+benchmark, but keep the machine running continually during the runs of the same
+benchmark. Each run is executed in isolation, without any additional processes
+running beyond services that start when the system is booted and a single ssh
+session. We are careful to avoid opening extra ssh sessions, running screen or
+tmux, or executing extra commands while a benchmark is running, as these may
+impact memory usage, and thus our results.
+
+We measured latency using the \texttt{rdtsc} instruction provided by x86\_64
+processors, which gives high resolution cycle-level timestamps. We take a
+timestamp before and after each memory management operation in the workload and
+take the difference to get latency. The median latency of each operation is
+reported across all 5 runs.
+
+To measure memory, we use the procfs's reporting on current memory usage, along
+with the slab allocator's count of active \texttt{vm\_area\_struct}s in the
+system. Measuring kernel memory usage and associating memory usage with
+processes is very difficult. Our methodology assumes that our benchmarks running
+in otherwise idle systems should dominate increases in memory usage.  Thus, we
+compute total memory usage in the system as the total amount of memory less the
+amount of memory used. From this amount, we subtract the resident set size of
+the benchmark process. The result is the memory usage of the rest of the system
+$R$. We denote $R_i$ as the $i$-th measurement of $R$ during a run ($i = 0, 1,
+... 2^{20}$). System memory usage is known to jitter in Linux (TODO: cite). To
+adjust for this jitter, we define $\Delta_i = R_i - R_0$. For each operation, we
+take the median value of $\Delta_i$ across the 5 runs to reduce jitter further.
+This is the value graphed in the figures in this section. $\Delta_i$ represents
+the increase in memory usage due to memory management overheads caused by
+running $i$ memory management operations.
+
+The workloads we measure are
+
+\begin{itemize} \item \textit{Control}. This workload is only run for the memory
+overhead experiments. It simply prints the $\Delta_i$ for each $i$. It's graph
+is a straight line with positive slope. This memory usage results from the
+buffer containing stdout, where our readings are being printed. All of our
+experiments have the same stdout overhead, so this benchmark serves a baseline
+for comparison.
+
+\item \textit{Continuous} (``cont'' for short). This workload allocates $2^{20}$
+adjacent 16KB blocks.
+
+\item \textit{Strided}. This workload allocates $2^{20}$ untouching 16KB regions
+in order of increasing memory address.
+
+\item \textit{Random}. This workload allocates $2^{20}$ randomly placed,
+non-overlaping memory regions.
+
+\item \textit{Fragmented} (``frag'' for short). This workload first allocates
+$2^{19} + 2^{18}$ contiguous 4KB blocks. Then, it resizes every other block to
+40KB. This represents the worst case behavior we could think of. It requires the
+kernel to find a block it has not touched in over $2^{19}$ operations and move
+it somewhere else.  \end{itemize}
+
+\subsection{VM Experiments}
+
+TODO
+
 \bibliography{references}{}

### 2017-04-03 bench_mem_control.cc
uncommented control

@@ -6,7 +6,7 @@
 
 int main(int argc, char **argv)
 {
-/*	if (argc < 2) {
+	if (argc < 2) {
         std::cout << "Usage: ./memusage number_order" << std::endl;
 		return -1;
 	}
@@ -21,5 +21,5 @@ int main(int argc, char **argv)
         std::cout << getkernelmem() << " " 
                   << slabinfo("vm_area_struct") << " " 
                   << meminfo("PageTables") << std::endl;
-	}*/
+	}
 }

### 2017-04-03 line_graph.py
Update line_graph.py

@@ -19,7 +19,7 @@ with open(DATA, 'r') as f:
         data.append(xs)
 
 for i in range(len(data[0])) if WHICH is None else WHICH:
-    ys = [x[i] for x in data]
+    ys = [x[i] - data[0][i] for x in data]
     line, = plt.plot(ys)
 
 #plt.xscale("log", basex = 2)

### 2017-04-03 line_graph.py
Adding scripts for analysis

@@ -1,35 +1,47 @@
 import numpy as np
 import matplotlib.pyplot as plt
 import matplotlib
-from sys import argv
+from sys import argv, exit
 
-DATA = argv[1]
-SCALE = argv[2] if len(argv) >= 3 else "linear"
-WHICH = [int(w) for w in argv[3].split(",")] if len(argv) >= 4 else None
+if len(argv) < 5:
+    print("Usage: ./script WHICH AXIS_X AXIS_Y FILE [FILE, FILE, ...]")
+    print("WHICH = <column #>[,<column #>,...]")
+    print("AXIS_X = <x axis label>")
+    print("AXIS_Y = <y axis label>")
+    exit(-1)
 
-data = []
+WHICH = [int(x) for x in argv[1].split(',')]
+AXIS_X = argv[2]
+AXIS_Y = argv[3]
 
-with open(DATA, 'r') as f:
-    for line in f:
-        if len(line.strip()) == 0:
-            continue
+data = {}
 
-        xs = [float(x) for x in line.split()]
+for data_file in argv[4:]:
+    with open(data_file, 'r') as f:
+        data[data_file] = []
 
-        data.append(xs)
+        for line in f:
+            if len(line.strip()) == 0:
+                continue
 
-for i in range(len(data[0])) if WHICH is None else WHICH:
-    ys = [x[i] - data[0][i] for x in data]
-    line, = plt.plot(ys)
+            xs = [float(x) for x in line.split()]
+
+            data[data_file].append(xs)
+
+for data_file in data:
+    for i in WHICH:
+        if i < len(data[data_file][0]):
+            ys = [x[i] - data[data_file][0][i] for x in data[data_file]]
+            line, = plt.plot(ys)
 
 #plt.xscale("log", basex = 2)
-plt.yscale(SCALE)
+#plt.yscale(SCALE)
 
 #plt.xlim(0, 512+32)
 #plt.ylim(0, 600)
 
-#plt.xlabel("Message Size (bytes)")
-#plt.ylabel("Latency ($\mu$secs)")
+plt.xlabel(AXIS_X)
+plt.ylabel(AXIS_Y)
 plt.grid(True)
 
 font = {'family' : 'normal',

### 2017-04-03 medians.py

@@ -1,0 +1,40 @@
+import numpy as np
+from sys import argv, exit
+
+if len(argv) < 4:
+    print("Usage: ./script WHICH OUTPUT_FILE FILE [FILE, FILE, ...]")
+    print("WHICH = <column #>[,<column #>,...]")
+    exit(-1)
+
+WHICH = [int(x) for x in argv[1].split(',')]
+OUTPUT_FILE = argv[2]
+
+data = {}
+
+medians = []
+
+for data_file in argv[3:]:
+    with open(data_file, 'r') as f:
+        data[data_file] = []
+
+        for line in f:
+            if len(line.strip()) == 0:
+                continue
+
+            xs = [float(x) for x in line.split()]
+
+            data[data_file].append(xs)
+
+for row in range(len(data.values()[0])):
+    medians_row = []
+    for i in WHICH:
+        vals = [data[data_file][row][i] - data[data_file][0][i] for data_file in data]
+        medians_row.append(np.median(vals))
+    medians.append(medians_row)
+
+with open(OUTPUT_FILE, 'w') as f:
+    for median_row in medians:
+        f.write("%s\n" % " ".join([str(x) for x in median_row]))
+
+    f.flush()
+

### 2017-04-03 do_latency_measurements.py
latency results

@@ -13,6 +13,31 @@ def run_mmap_cont():
     print "cont run 5"
     os.system("./bench_time_per_op_mmap_cont 20 0 >> time_per_op_mmap_cont_5") 
 
+    files = ["time_per_op_mmap_cont_1", "time_per_op_mmap_cont_2","time_per_op_mmap_cont_3",
+            "time_per_op_mmap_cont_4","time_per_op_mmap_cont_5"]
+    out_file_name = "time_per_op_mmap_cont.res"
+    merge_output(files, out_file_name)
+
+def merge_output(files, file_name):
+    dic = {}
+    i = 0
+    for f in files:
+        with open(f)  as cur_file:
+            for line in cur_file.readlines():
+                if i not in dic:
+                    dic[i] = []
+                dic[i].append(line)
+                i = i + 1
+        i = 0
+
+    f = open(file_name, "w")
+    for k, v in dic.iteritems():
+        for val in v:
+            f.write(val.rstrip() + " ")
+        f.write("\n")
+    f.close()
+
+
 def run_mmap_strided():
     print "strided run 1"
     os.system("./bench_time_per_op_mmap_stride 20 0 >> time_per_op_mmap_stride_1") 
@@ -25,9 +50,19 @@ def run_mmap_strided():
     print "strided run 5"
     os.system("./bench_time_per_op_mmap_stride 20 0 >> time_per_op_mmap_stride_5") 
 
+    files = ["time_per_op_mmap_stride_1","time_per_op_mmap_stride_2", "time_per_op_mmap_stride_3", 
+             "time_per_op_mmap_stride_4", "time_per_op_mmap_stride_5"]
+    out_file_name = "time_per_op_mmap_stride.res"
+
+    merge_output(files, out_file_name)
 
 
 def main():
+    run_mmap_cont()
+    run_mmap_strided()
+
+    return  ## don't wanna pollute the environment by compiling during the actual run!
+
     process = Popen(["make", "clean"], stdout=PIPE)
     (output, err) = process.communicate()
     exit_code = process.wait()
@@ -44,6 +79,9 @@ def main():
         print "Compile Failed.."
         return
     print "Successfully Compiled..."
+
+    run_mmap_cont()
+    run_mmap_strided()
 
     
     

### 2017-04-03 line_graph.py
Reorganized results

@@ -28,11 +28,14 @@ for data_file in argv[4:]:
 
             data[data_file].append(xs)
 
+handles = []
+
 for data_file in data:
     for i in WHICH:
         if i < len(data[data_file][0]):
             ys = [x[i] - data[data_file][0][i] for x in data[data_file]]
-            line, = plt.plot(ys)
+            line, = plt.plot(ys, label = data_file)
+            handles.append(line)
 
 #plt.xscale("log", basex = 2)
 #plt.yscale(SCALE)
@@ -43,6 +46,8 @@ for data_file in data:
 plt.xlabel(AXIS_X)
 plt.ylabel(AXIS_Y)
 plt.grid(True)
+
+plt.legend(handles = handles)
 
 font = {'family' : 'normal',
         'weight' : 'bold',

### 2017-04-03 do_latency_measurements.py
bug fix

@@ -3,15 +3,15 @@ import os
 
 def run_mmap_cont():
     print "cont run 1"
-    os.system("./bench_time_per_op_mmap_cont 20 0 >> time_per_op_mmap_cont_1") 
+    os.system("./bench_time_per_op_mmap_cont 20 0 > time_per_op_mmap_cont_1") 
     print "cont run 2"
-    os.system("./bench_time_per_op_mmap_cont 20 0 >> time_per_op_mmap_cont_2") 
+    os.system("./bench_time_per_op_mmap_cont 20 0 > time_per_op_mmap_cont_2") 
     print "cont run 3"
-    os.system("./bench_time_per_op_mmap_cont 20 0 >> time_per_op_mmap_cont_3") 
+    os.system("./bench_time_per_op_mmap_cont 20 0 > time_per_op_mmap_cont_3") 
     print "cont run 4"
-    os.system("./bench_time_per_op_mmap_cont 20 0 >> time_per_op_mmap_cont_4") 
+    os.system("./bench_time_per_op_mmap_cont 20 0 > time_per_op_mmap_cont_4") 
     print "cont run 5"
-    os.system("./bench_time_per_op_mmap_cont 20 0 >> time_per_op_mmap_cont_5") 
+    os.system("./bench_time_per_op_mmap_cont 20 0 > time_per_op_mmap_cont_5") 
 
     files = ["time_per_op_mmap_cont_1", "time_per_op_mmap_cont_2","time_per_op_mmap_cont_3",
             "time_per_op_mmap_cont_4","time_per_op_mmap_cont_5"]
@@ -40,15 +40,15 @@ def merge_output(files, file_name):
 
 def run_mmap_strided():
     print "strided run 1"
-    os.system("./bench_time_per_op_mmap_stride 20 0 >> time_per_op_mmap_stride_1") 
+    os.system("./bench_time_per_op_mmap_stride 20 0 > time_per_op_mmap_stride_1") 
     print "strided run 2"
-    os.system("./bench_time_per_op_mmap_stride 20 0 >> time_per_op_mmap_stride_2") 
+    os.system("./bench_time_per_op_mmap_stride 20 0 > time_per_op_mmap_stride_2") 
     print "strided run 3"
-    os.system("./bench_time_per_op_mmap_stride 20 0 >> time_per_op_mmap_stride_3") 
+    os.system("./bench_time_per_op_mmap_stride 20 0 > time_per_op_mmap_stride_3") 
[…]

### 2017-04-03 line_graph_with_control.py
Added version of graph script that takes a control

@@ -1,0 +1,59 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib
+from sys import argv, exit
+
+if len(argv) < 6:
+    print("Usage: ./script WHICH AXIS_X AXIS_Y CONTROL FILE [FILE, FILE, ...]")
+    print("WHICH = <column #>[,<column #>,...]")
+    print("AXIS_X = <x axis label>")
+    print("AXIS_Y = <y axis label>")
+    exit(-1)
+
+WHICH = [int(x) for x in argv[1].split(',')]
+AXIS_X = argv[2]
+AXIS_Y = argv[3]
+CONTROL = argv[4]
+
+data = {}
+
+for data_file in argv[4:]:
+    with open(data_file, 'r') as f:
+        data[data_file] = []
+
+        for line in f:
+            if len(line.strip()) == 0:
+                continue
+
+            xs = [float(x) for x in line.split()]
+
+            data[data_file].append(xs)
+
+handles = []
+
+for data_file in data:
+    for i in WHICH:
+        if i < len(data[data_file][0]):
+            ys = [data[data_file][x][i] - data[data_file][0][i] - data[CONTROL][x][i] for x in range(len(data[data_file]))]
+            line, = plt.plot(ys, label = data_file)
+            handles.append(line)
+
+#plt.xscale("log", basex = 2)
+#plt.yscale(SCALE)
+
+#plt.xlim(0, 512+32)
+#plt.ylim(0, 600)
+
+plt.xlabel(AXIS_X)
+plt.ylabel(AXIS_Y)
+plt.grid(True)
+
+plt.legend(handles = handles)
+
+font = {'family' : 'normal',
+        'weight' : 'bold',
+        'size'   : 22}
+
+matplotlib.rc('font', **font)
+
+plt.show()

### 2017-04-04 bench_time_per_op_mmap_4k_rand.cc
Measure time for rand and frag tests

@@ -1,0 +1,68 @@
+#include <iostream>
+#include <stdlib.h>
+#include <sys/mman.h>
+#include <string.h>
+#include <errno.h>
+#include <sys/types.h>
+#include <sys/stat.h>
+#include <fcntl.h>
+#include <unistd.h>
+#include <stdio.h>
+
+#include "rdmemusg.h"
+#include "pattern.h"
+#include "rdtsc.h"
+#include "pinning.h"
+
+#define START 0x40000000
+
+int main(int argc, char **argv) {
+    set_cpu(0);
+
+    if (argc < 2) {
+      std::cout << "Usage: ./memusage number_order" << std::endl;
+      return -1;
+    }
+
+    size_t amt = 1 << atoi(argv[1]);
+
+    // Read randomized offsets from a file
+    FILE *f = fopen("offsets.txt", "r");
+    int *offsets;
+    unsigned int i;
+
+    if (!f) {
+      fputs("Couldn't open offsets.txt\n", stderr);
+      return -1;
+    }
+
+    offsets = (int *)malloc(sizeof(int) * amt);
+
+    for (i = 0; i < amt; i++) {
+      fscanf(f, "%d", &offsets[i]);
+    }
+
+    if (fclose(f)) {
+      fputs("Couldn't close offsets.txt\n", stderr);
+      free(offsets);
+      return -1;
+    }
+
+    double ticks_per_nano = get_ticks_per_nano();  // calibrate rdtsc before measuring
+
+    for (size_t i = 0; i < amt; i++) {
+        unsigned long maddr = START + (offsets[i] << 12);
+        unsigned long long cur_start = rdtsc();
+        char *addr = (char *)mmap((void *)maddr, 4 << 12, 7, MAP_ANON | MAP_PRIVATE, -1, 0);
+        unsigned long long cur_end = rdtsc();
+        if (addr == MAP_FAILED) {
+            std::cerr << "mmap failed: " << strerror(errno) << std::endl;
+            return -1;
+        }
+        print_interval(cur_start, cur_end, ticks_per_nano);
+    }
+
+    free(offsets);
+
+    return 0;
+}

### 2017-04-04 bench_time_per_op_mmap_frag.cc

@@ -1,0 +1,54 @@
+#include <iostream>
+#include <stdlib.h>
+#include <sys/mman.h>
+#include <string.h>
+#include <errno.h>
+#include <stdio.h>
+
+#include "rdmemusg.h"
+#include "pattern.h"
+#include "rdtsc.h"
+#include "pinning.h"
+
+int main(int argc, char **argv) {
+    set_cpu(0);
+
+    if (argc < 2) {
+      std::cout << "Usage: ./memusage number_order" << std::endl;
+      return -1;
+    }
+
+    size_t amt = 1 << atoi(argv[1]);
+
+    PatternGenerator pg = get_mmap_cont(4 /* 16KB */, 7 /* RWX */);
+    Next next;
+
+    double ticks_per_nano = get_ticks_per_nano();  // calibrate rdtsc before measuring
+
+    for (size_t i = 0; i < amt; i++) {
+        next = pg.next();
+
+        char *addr = (char *)mmap(next.address, next.size, next.permissions, MAP_ANON | MAP_PRIVATE, -1, 0);
+
+        if (addr == MAP_FAILED) {
+            std::cerr << "mmap failed: " << strerror(errno) << std::endl;
+            return -1;
+        }
+    }
+
+    pg = get_mremap_frag(1, 4, 7);
+
+    for (size_t i = 0; i < amt; i++) {
+        next = pg.next();
+
+        unsigned long long cur_start = rdtsc();
+        char *addr = (char *)mremap(next.address, 2 << 12, next.size, MREMAP_MAYMOVE);
+        unsigned long long cur_end = rdtsc();
+
+        if (addr == MAP_FAILED) {
+            std::cerr << "mmap failed: " << strerror(errno) << std::endl;
+            return -1;
+        }
+        print_interval(cur_start, cur_end, ticks_per_nano);
+    }
+}

### 2017-04-04 bench_time_per_op_mmap_4k_cont.cc
new latency results

@@ -1,45 +1,0 @@
-#include <stdio.h>
-#include <stdlib.h>
-#include <sys/mman.h>
-
-#include "pattern.h"
-#include "rdtsc.h"
-#include "pinning.h"
-
-int main(int argc, char* argv[]) {
-    set_cpu(0);
-    PatternGenerator pg = get_mmap_cont(1 /* 4KB */, 7 /* RWX */);
-
-    Next cur;
-    char* addr;
-
-    if (argc < 3) {
-        printf("Usage: ./pattern total_ops_order touch_0_or_1\n");
-        exit(1);
-    }
-
-    int tot_ops = 1 << atoi(argv[1]);
-    int touch = atoi(argv[2]);
-    
-    double ticks_per_nano = get_ticks_per_nano();  // calibrate rdtsc before measuring
-
-    
-    for (int i = 0; i < tot_ops; i++) {
-        //  assuming all ops were sucessful
-        cur = pg.next();
-        if (cur.operation == OP_MMAP) {  // must be
-            unsigned long long cur_start = rdtsc();
-            addr = (char *)mmap(cur.address, cur.size, cur.permissions, MAP_ANON | MAP_PRIVATE, -1, 0);
-            unsigned long long cur_end = rdtsc();
-            if (addr == MAP_FAILED) {
-                printf("mmap failed\n");
-                exit(1);
-            }
-            print_interval(cur_start, cur_end, ticks_per_nano);
-        }
-        if (touch == 1) addr[0] = 'X'; // touch
-    }
-
-    return 0;
-
-}

### 2017-04-04 bench_time_per_op_mmap_stride.cc

@@ -1,45 +1,0 @@
-#include <stdio.h>
-#include <stdlib.h>
-#include <sys/mman.h>
-
-#include "pattern.h"
-#include "rdtsc.h"
-#include "pinning.h"
-
-int main(int argc, char* argv[]) {
-    set_cpu(0);
-    PatternGenerator pg = get_mmap_stride(1 /* 4KB */, 7 /* RWX */, 2 /* 8KB */);
-
-    Next cur;
-    char* addr;
-
-    if (argc < 3) {
-        printf("Usage: ./pattern total_ops_order touch_0_or_1\n");
-        exit(1);
-    }
-
-    int tot_ops = 1 << atoi(argv[1]);
-    int touch = atoi(argv[2]);
-    
-    double ticks_per_nano = get_ticks_per_nano();  // calibrate rdtsc before measuring
-
-    
-    for (int i = 0; i < tot_ops; i++) {
-        //  assuming all ops were sucessful
-        cur = pg.next();
-        if (cur.operation == OP_MMAP) {  // must be
-            unsigned long long cur_start = rdtsc();
-            addr = (char *)mmap(cur.address, cur.size, cur.permissions, MAP_ANON | MAP_PRIVATE, -1, 0);
-            unsigned long long cur_end = rdtsc();
-            if (addr == MAP_FAILED) {
-                printf("mmap failed\n");
-                exit(1);
-            }
-            print_interval(cur_start, cur_end, ticks_per_nano);
-        }
-        if (touch == 1) addr[0] = 'X'; // touch
-    }
-
-    return 0;
-
-}

### 2017-04-04 do_latency_measurements.py

@@ -3,20 +3,20 @@ import os
 
 def run_mmap_cont():
     print "cont run 1"
-    os.system("./bench_time_per_op_mmap_cont 20 0 > time_per_op_mmap_cont_1") 
+    os.system("./bench_time_per_op_mmap_16k_cont 20 0 > time_per_op_mmap_cont_1") 
     print "cont run 2"
-    os.system("./bench_time_per_op_mmap_cont 20 0 > time_per_op_mmap_cont_2") 
+    os.system("./bench_time_per_op_mmap_16k_cont 20 0 > time_per_op_mmap_cont_2") 
     print "cont run 3"
-    os.system("./bench_time_per_op_mmap_cont 20 0 > time_per_op_mmap_cont_3") 
+    os.system("./bench_time_per_op_mmap_16k_cont 20 0 > time_per_op_mmap_cont_3") 
     print "cont run 4"
-    os.system("./bench_time_per_op_mmap_cont 20 0 > time_per_op_mmap_cont_4") 
+    os.system("./bench_time_per_op_mmap_16k_cont 20 0 > time_per_op_mmap_cont_4") 
     print "cont run 5"
-    os.system("./bench_time_per_op_mmap_cont 20 0 > time_per_op_mmap_cont_5") 
+    os.system("./bench_time_per_op_mmap_16k_cont 20 0 > time_per_op_mmap_cont_5") 
 
     files = ["time_per_op_mmap_cont_1", "time_per_op_mmap_cont_2","time_per_op_mmap_cont_3",
             "time_per_op_mmap_cont_4","time_per_op_mmap_cont_5"]
     out_file_name = "time_per_op_mmap_cont.res"
-    merge_output(files, out_file_name)
+    #merge_output(files, out_file_name)
 
 def merge_output(files, file_name):
     dic = {}
@@ -40,25 +40,25 @@ def merge_output(files, file_name):
 
 def run_mmap_strided():
     print "strided run 1"
-    os.system("./bench_time_per_op_mmap_stride 20 0 > time_per_op_mmap_stride_1") 
+    os.system("./bench_time_per_op_mmap_16k_stride 20 0 > time_per_op_mmap_stride_1") 
     print "strided run 2"
-    os.system("./bench_time_per_op_mmap_stride 20 0 > time_per_op_mmap_stride_2") 
+    os.system("./bench_time_per_op_mmap_16k_stride 20 0 > time_per_op_mmap_stride_2") 
     print "strided run 3"
[…]

### 2017-04-12 paper.tex
Added swapping experiments

@@
 available to researchers.
+
+(TODO: we might want to add stuff about using swap instead of larger memory)
 
@@
 Our methodology is broken into two parts. First, we measure the scalability of
-several Linux memory management system calls on a system running natively on
-bare metal. Then, we measure the scalability of Linux virtual machines running
-on the KVM hypervisor.
+Linux's \texttt{vm\_area\_struct}. Then, we measure the minimum amount of memory
+necessary for swapping.
 
-\subsection{Native Experiments}
+\subsection{\texttt{vm\_area\_struct} Experiments}
 
@@
 system's use of \texttt{vm\_area\_struct}, which represents a region of a
-process's virtual address space. (TODO: what about page tables? struct page?)
-When a program makes a memory management system calls such as \texttt{mmap},
+process's virtual address space.  We do not measure the overhead of
+\texttt{struct page} because it is known that there is one struct for each
+physical page. We do not measure the overhead of page tables because this can be
+calculated from the size of the virtual memory space used by the process.  When
+a program makes a memory management system calls such as \texttt{mmap},
 \texttt{mremap}, \texttt{munmap}, \texttt{mprotect}, \texttt{mlock}, or
@@
 
-\item \textit{Random}. This workload allocates $2^{20}$ randomly placed,
-non-overlaping memory regions.
+\item \textit{Random}. This workload allocates randomly the pages in a
+inside of a $2^{20}$ page region.
 
@@
 
-\subsection{VM Experiments}
+\subsection{Swapping Experiments}
+
+These experiments are designed to measure the minimum amount of memory needed
+to using swapping.
+
+In this experiment, a single process touches 32GB of virtual memory, leading to
+the actual allocation of physical memory. Notice that 32GB is twice as large as
+the 16GB of physical memory the test machine. This causes Linux kernel to start
+swapping out pages to make more room.
+
+After each page of virtual memory is touched, the resident set size (RSS; the
+amount of physical memory Linux has allocated to the process) is recorded. This
+value is graphed in our results section.
+
+We repeat this experiment with 5 different swapfile sizes to capture the impact
+of swapfile size on swapping overhead.
+
+(TODO: strided access vs random access?)
+
+\section{Result}
+
+\subsection{\texttt{vm\_area\_struct} Experiments}
+
+TODO
+
+\subsection{Swapping Experiments}
 

### 2017-04-12 paper.tex
Added to methodology again

@@
 the 16GB of physical memory the test machine. This causes Linux kernel to start
-swapping out pages to make more room.
+swapping out pages to make more room.  After each page of virtual memory is
+touched, the resident set size (RSS; the amount of physical memory Linux has
+allocated to the process) is recorded. This value is graphed in our results
+section.
 
-After each page of virtual memory is touched, the resident set size (RSS; the
-amount of physical memory Linux has allocated to the process) is recorded. This
-value is graphed in our results section.
-
-We repeat this experiment with 5 different swapfile sizes to capture the impact
+This experiment with 5 different swapfile sizes to capture the impact
 of swapfile size on swapping overhead.
@@
 (TODO: strided access vs random access?)
+
+During this experiment, the values of RSS are stored in a mlocked array and
+written to a file after the benchmark completes. This prevents the writing of
+the results from interacting with paging.
+
+(TODO: optional: trying pulling out a DIMM??)
 

### 2017-04-12 paper.tex
A bit of clean up

@@
 the actual allocation of physical memory. Notice that 32GB is twice as large as
-the 16GB of physical memory the test machine. This causes Linux kernel to start
-swapping out pages to make more room.  After each page of virtual memory is
-touched, the resident set size (RSS; the amount of physical memory Linux has
-allocated to the process) is recorded. This value is graphed in our results
-section.
+the 16GB of physical memory in the test machine, causing Linux to start paging
+after physical memory is exhausted. Also, notice that the Linux kernel will
+reserve some memory for its own operation.
 
-This experiment with 5 different swapfile sizes to capture the impact
-of swapfile size on swapping overhead.
+After each page of virtual memory is touched, the resident set size (RSS; the
+amount of physical memory Linux has allocated to the process) is recorded.  The
+values of RSS are stored in an mlocked array and written to a file after the
+benchmark completes. This prevents the writing of the results from interacting
+with paging.  These RSS values are graphed in our results section.
+
+This experiment is repeated with 5 different swapfile sizes to capture the
+impact of swapfile size on swapping overhead.
 
 (TODO: strided access vs random access?)
-
-During this experiment, the values of RSS are stored in a mlocked array and
-written to a file after the benchmark completes. This prevents the writing of
-the results from interacting with paging.
 

### 2017-04-12 bench_mmap_anon_rss.cc
add RSS benchmark for swap overhead measurement

@@ -1,0 +1,50 @@
+
+#include <iostream>
+#include <stdlib.h>
+#include <sys/mman.h>
+#include <string.h>
+#include <errno.h>
+#include <stdint.h>
+
+#include "rdmemusg.h"
+#include "pattern.h"
+#include "pinning.h"
+
+int main(int argc, char **argv) {
+    set_cpu(0);
+
+
+	if (argc < 1) {
+        std::cout << "Usage: ./memusage number_order" << std::endl;
+        // whatever!
+		return -1;
+	}
+
+    uint64_t vma_size = (uint64_t)1 << 35; // fixed at 32 GB
+
+    char* start_addr = (char*)mmap(0, vma_size, 7, MAP_ANONYMOUS | MAP_PRIVATE, -1, 0);
+    if (start_addr == MAP_FAILED) {
+       printf("fuck\n");
+       exit(1);
+    }
+
+    unsigned long long num_pages_mapped = vma_size / PAGE_SIZE;
+    unsigned long* res_array = (unsigned long*)malloc(num_pages_mapped * 
+            sizeof(unsigned long));
+    // lock this results array itself so itself is not swapped out
+    mlock((const void*) res_array, num_pages_mapped * sizeof(unsigned long));
+
+    
+    // touch all the pages now
+    unsigned long long num_touched = 0;
+    char* addr = start_addr;
+    while (num_touched < num_pages_mapped) {
+        addr[0] = 'X';
+        res_array[num_touched++] = statusinfo("VmRSS");
+
+        addr += PAGE_SIZE;
+    }
+
+    for (unsigned int i = 0; i < num_pages_mapped; i++) std::cout << res_array[i++] << std::endl;
+
+}

### 2017-04-12 bench_mmap_anon_rss.cc
Check mlock failure

@@ -13,38 +13,35 @@
 int main(int argc, char **argv) {
     set_cpu(0);
 
-
-	if (argc < 1) {
-        std::cout << "Usage: ./memusage number_order" << std::endl;
-        // whatever!
-		return -1;
-	}
-
     uint64_t vma_size = (uint64_t)1 << 35; // fixed at 32 GB
 
-    char* start_addr = (char*)mmap(0, vma_size, 7, MAP_ANONYMOUS | MAP_PRIVATE, -1, 0);
+    char* start_addr = (char*)mmap(NULL, vma_size, PROT_READ | PROT_WRITE, MAP_ANONYMOUS | MAP_PRIVATE, -1, 0);
     if (start_addr == MAP_FAILED) {
-       printf("fuck\n");
-       exit(1);
+        std::cerr << "mmap failed: " << strerror(errno) << std::endl;
+        exit(1);
     }
 
     unsigned long long num_pages_mapped = vma_size / PAGE_SIZE;
     unsigned long* res_array = (unsigned long*)malloc(num_pages_mapped * 
             sizeof(unsigned long));
-    // lock this results array itself so itself is not swapped out
-    mlock((const void*) res_array, num_pages_mapped * sizeof(unsigned long));
 
+    // lock this results array itself so itself is not swapped out
+    if(!mlock((const void*) res_array, num_pages_mapped * sizeof(unsigned long))) {
+        std::cerr << "mlock failed: " << strerror(errno) << std::endl;
+        exit(1);
+    }
     
     // touch all the pages now
     unsigned long long num_touched = 0;
     char* addr = start_addr;
     while (num_touched < num_pages_mapped) {
-        addr[0] = 'X';
+        *addr = 'X';
         res_array[num_touched++] = statusinfo("VmRSS");
 
         addr += PAGE_SIZE;
     }
 
-    for (unsigned int i = 0; i < num_pages_mapped; i++) std::cout << res_array[i++] << std::endl;
-
+    for (unsigned int i = 0; i < num_pages_mapped; i++) {
+        std::cout << res_array[i++] << std::endl;
+    }
 }

### 2017-04-12 bench_mmap_anon_rss_rand.cc
Added random swap bmk

@@ -1,0 +1,53 @@
+
+#include <iostream>
+#include <stdlib.h>
+#include <sys/mman.h>
+#include <string.h>
+#include <errno.h>
+#include <stdint.h>
+
+#include <cstdlib>
+#include <ctime>
+
+#include "rdmemusg.h"
+#include "pattern.h"
+#include "pinning.h"
+
+int main(int argc, char **argv) {
+    set_cpu(0);
+
+    uint64_t vma_size = (uint64_t)1 << 35; // fixed at 32 GB
+
+    char* start_addr = (char*)mmap(NULL, vma_size, PROT_READ | PROT_WRITE, MAP_ANONYMOUS | MAP_PRIVATE, -1, 0);
+    if (start_addr == MAP_FAILED) {
+        std::cerr << "mmap failed: " << strerror(errno) << std::endl;
+        exit(1);
+    }
+
+    unsigned long long num_pages_mapped = vma_size / PAGE_SIZE;
+    unsigned long* res_array = (unsigned long*)malloc(num_pages_mapped * 
+            sizeof(unsigned long));
+
+    // lock this results array itself so itself is not swapped out
+    if(!mlock((const void*) res_array, num_pages_mapped * sizeof(unsigned long))) {
+        std::cerr << "mlock failed: " << strerror(errno) << std::endl;
+        exit(1);
+    }
+
+    // seed the rng
+	srand((int)time(0));
+    
+    // touch all the pages now
+    unsigned long long num_touched = 0;
+    char* addr = start_addr;
+    while (num_touched < num_pages_mapped) {
+        *addr = 'X';
+        res_array[num_touched++] = statusinfo("VmRSS");
+
+        addr = ((char*)start_addr) + (rand() % vma_size);
+    }
+
+    for (unsigned int i = 0; i < num_pages_mapped; i++) {
+        std::cout << res_array[i++] << std::endl;
+    }
+}

### 2017-04-13 bench_mmap_anon_rss.cc
Fixed some stuff

@@ -25,6 +25,11 @@ int main(int argc, char **argv) {
     unsigned long* res_array = (unsigned long*)malloc(num_pages_mapped * 
             sizeof(unsigned long));
 
+    // touch all pages
+    for(unsigned long i = 0; i < num_pages_mapped; i++) {
+        res_array[i] = 0;
+    }
+
     // lock this results array itself so itself is not swapped out
     if(!mlock((const void*) res_array, num_pages_mapped * sizeof(unsigned long))) {
         std::cerr << "mlock failed: " << strerror(errno) << std::endl;

### 2017-04-13 bench_mmap_anon_rss_rand.cc

@@ -28,6 +28,11 @@ int main(int argc, char **argv) {
     unsigned long* res_array = (unsigned long*)malloc(num_pages_mapped * 
             sizeof(unsigned long));
 
+    // touch all pages
+    for(unsigned long i = 0; i < num_pages_mapped; i++) {
+        res_array[i] = 0;
+    }
+
     // lock this results array itself so itself is not swapped out
     if(!mlock((const void*) res_array, num_pages_mapped * sizeof(unsigned long))) {
         std::cerr << "mlock failed: " << strerror(errno) << std::endl;
@@ -44,10 +49,16 @@ int main(int argc, char **argv) {
         *addr = 'X';
         res_array[num_touched++] = statusinfo("VmRSS");
 
-        addr = ((char*)start_addr) + (rand() % vma_size);
+        int rand_hi = rand();
+        int rand_lo = rand();
+        unsigned long rand_30 = ((unsigned long)rand_hi) << 15 | rand_lo;
+        unsigned long rand_22 = rand_30 & 0x3FFFFF;
+
+        addr = ((char*)start_addr) + rand_22;
+        //printf("%lx %lx\n", (unsigned long)start_addr, (unsigned long)addr);
     }
 
     for (unsigned int i = 0; i < num_pages_mapped; i++) {
-        std::cout << res_array[i++] << std::endl;
+        //std::cout << res_array[i++] << std::endl;
     }
 }

### 2017-04-13 bench_mmap_anon_rss_rand.cc
Fixed a bug?

@@ -51,7 +51,7 @@ int main(int argc, char **argv) {
 
         int rand_hi = rand();
         int rand_lo = rand();
-        unsigned long rand_30 = ((unsigned long)rand_hi) << 15 | rand_lo;
+        unsigned long rand_30 = (((unsigned long)rand_hi) << 15) ^ rand_lo;
         unsigned long rand_22 = rand_30 & 0x3FFFFF;
 
         addr = ((char*)start_addr) + rand_22;
@@ -59,6 +59,6 @@ int main(int argc, char **argv) {
     }
 
     for (unsigned int i = 0; i < num_pages_mapped; i++) {
-        //std::cout << res_array[i++] << std::endl;
+        std::cout << res_array[i++] << std::endl;
     }
 }

### 2017-04-13 bench_mmap_anon_rss_rand.cc
Proper amount of randomness

@@ -50,15 +50,16 @@ int main(int argc, char **argv) {
         res_array[num_touched++] = statusinfo("VmRSS");
 
         int rand_hi = rand();
+        int rand_mi = rand();
         int rand_lo = rand();
-        unsigned long rand_30 = (((unsigned long)rand_hi) << 15) ^ rand_lo;
-        unsigned long rand_22 = rand_30 & 0x3FFFFF;
+        unsigned long rand_45 = (((unsigned long)rand_hi) << 30) ^ (((unsigned long)rand_mi) << 15) ^ rand_lo;
+        unsigned long rand_35 = rand_45 & 0x7FFFFFFFF;
 
-        addr = ((char*)start_addr) + rand_22;
-        //printf("%lx %lx\n", (unsigned long)start_addr, (unsigned long)addr);
+        addr = ((char*)start_addr) + rand_35;
+        //printf("%lx\n", (unsigned long)addr);
     }
 
     for (unsigned int i = 0; i < num_pages_mapped; i++) {
-        std::cout << res_array[i++] << std::endl;
+        //std::cout << res_array[i++] << std::endl;
     }
 }

### 2017-04-13 bench_mmap_anon_rss_rand.cc
:weeping:

@@ -60,6 +60,6 @@ int main(int argc, char **argv) {
     }
 
     for (unsigned int i = 0; i < num_pages_mapped; i++) {
-        //std::cout << res_array[i++] << std::endl;
+        std::cout << res_array[i++] << std::endl;
     }
 }

### 2017-04-13 bench_mmap_anon_rss_rand.cc
bug fix

@@ -60,6 +60,6 @@ int main(int argc, char **argv) {
     }
 
     for (unsigned int i = 0; i < num_pages_mapped; i++) {
-        std::cout << res_array[i++] << std::endl;
+        std::cout << res_array[i] << std::endl;
     }
 }

### 2017-04-13 bench_mmap_anon_rss.cc
Added command line args for swapping bmks

@@ -13,7 +13,13 @@
 int main(int argc, char **argv) {
     set_cpu(0);
 
-    uint64_t vma_size = (uint64_t)1 << 35; // fixed at 32 GB
+	if (argc < 2) {
+        std::cout << "Usage: ./memusage num_gb" << std::endl;
+		return -1;
+	}
+
+	size_t gbs = atoi(argv[1]);
+    uint64_t vma_size = (uint64_t)gbs << 30;
 
     char* start_addr = (char*)mmap(NULL, vma_size, PROT_READ | PROT_WRITE, MAP_ANONYMOUS | MAP_PRIVATE, -1, 0);
     if (start_addr == MAP_FAILED) {

### 2017-04-13 bench_mmap_anon_rss_rand.cc

@@ -13,10 +13,26 @@
 #include "pattern.h"
 #include "pinning.h"
 
+// returns a random unsigned long with upto 45-bits of pseudo-randomness
+unsigned long random_number(unsigned long min, unsigned long max) {
+        int rand_hi = rand();
+        int rand_mi = rand();
+        int rand_lo = rand();
+        unsigned long rand_45 = (((unsigned long)rand_hi) << 30) ^ (((unsigned long)rand_mi) << 15) ^ rand_lo;
+
+        return rand_45 % max;
+}
+
 int main(int argc, char **argv) {
     set_cpu(0);
 
-    uint64_t vma_size = (uint64_t)1 << 35; // fixed at 32 GB
+	if (argc < 2) {
+        std::cout << "Usage: ./memusage num_gb" << std::endl;
+		return -1;
+	}
+
+	size_t gbs = atoi(argv[1]);
+    uint64_t vma_size = (uint64_t)gbs << 30;
 
     char* start_addr = (char*)mmap(NULL, vma_size, PROT_READ | PROT_WRITE, MAP_ANONYMOUS | MAP_PRIVATE, -1, 0);
     if (start_addr == MAP_FAILED) {
@@ -49,13 +65,7 @@ int main(int argc, char **argv) {
         *addr = 'X';
         res_array[num_touched++] = statusinfo("VmRSS");
 
-        int rand_hi = rand();
-        int rand_mi = rand();
-        int rand_lo = rand();
-        unsigned long rand_45 = (((unsigned long)rand_hi) << 30) ^ (((unsigned long)rand_mi) << 15) ^ rand_lo;
-        unsigned long rand_35 = rand_45 & 0x7FFFFFFFF;
-
-        addr = ((char*)start_addr) + rand_35;
+        addr = ((char*)start_addr) + random_number(0, vma_size);
         //printf("%lx\n", (unsigned long)addr);
     }
 

### 2017-04-13 bench_mmap_anon_rss.cc
Fixed bug

@@ -53,6 +53,6 @@ int main(int argc, char **argv) {
     }
 
     for (unsigned int i = 0; i < num_pages_mapped; i++) {
-        std::cout << res_array[i++] << std::endl;
+        std::cout << res_array[i] << std::endl;
     }
 }

### 2017-04-13 line_graph.py

@@ -34,6 +34,8 @@ for data_file in data:
     for i in WHICH:
         if i < len(data[data_file][0]):
             ys = [x[i] - data[data_file][0][i] for x in data[data_file]]
+            # uncomment next line to prevent normalizing to X_0
+            #ys = [x[i] for x in data[data_file]]
             line, = plt.plot(ys, label = data_file)
             handles.append(line)
 

### 2017-04-13 medians.py

@@ -29,6 +29,8 @@ for row in range(len(data.values()[0])):
     medians_row = []
     for i in WHICH:
         vals = [data[data_file][row][i] - data[data_file][0][i] for data_file in data]
+        # uncomment next line to prevent normalizing to X_0
+        #vals = [data[data_file][row][i] for data_file in data]
         medians_row.append(np.median(vals))
     medians.append(medians_row)
 

### 2017-04-20 paper.tex
Updated Intro, Related Work; added Backgrnd section

@@
 \usepackage{caption}
+\usepackage{graphicx}
 
-\title{Journey: A Study of Studying Large Memory Systems}
+\title{Journey: A Study of Studying Large Memory Systems (TODO: should we get a new title?)}
 \author{Mark Mansi, Suhas Pai, Hasnain Ali Pirzada}
@@
 Currently, there are few good tools available to systems researchers for
-understanding how systems interact with large memories. The ideal end-goal of
-this research is to identify bottlenecks in managing large memories and find a
+understanding how systems interact with large memories. We see the need for a
 general lightweight technique or tool for studying and designing operating
 systems capable of handling large memory while using systems currently
-available to researchers.
+available to researchers.  An important first step towards building such a tool
+is to identify bottlenecks in managing large memories that make a tool
+difficult to develop.
 
-(TODO: we might want to add stuff about using swap instead of larger memory)
+In particular, we look at the case of a single process using a large amount of
+memory in Linux. We run experiments to identify the costs of
+\texttt{vm\_area\_struct}, a common bookkeeping data structure in the Linux
+kernel, along with the memory and processing overheads of swapping. From these
+data, we extrapolate to systems running with more memory with an eye on
+implications for studying large systems. While these experiments are not
+comprehensive, we believe they can offer insights that are useful both for
+building a research tool and for building systems.
 
 \section{Related Work}
+
+(TODO: do we need to refocus this section? or move parts of it to an appendix?)
 
@@
 it requires modifying the benchmarks, which is error prone and time consuming;
-it can be extremely difficult to validate changes to a benchmark. In our work,
-we seek to avoid changes to the experimental workloads including changes to code
-or scale.
+it can be extremely difficult to validate changes to a benchmark. Ideally,
+researchers can run large experimental workloads without changes, but the overheads
+of current memory management systems prevent this from being practical.
 
@@
 than on DRAM, it demonstrates that new technologies may be emulated using
-existing technology \cite{quartz}.
+existing technology \cite{quartz}. (TODO: is this system still relavent?)
 
 The Simics simulator uses demand paging for simulated environments, which means
-that memory is only allocated when it is used. As long as the working set of the
-target system fits in the host memory, performance will be tolerable.  However,
-in our work, we wish to study OS behavior under workloads that actually use all
-available memory. Thus, the usefulness of demand paging is greatly reduced. For
-these workloads, the performance of simulators like Simics tends to be
-impractical \cite{simics}. We do not explore simulation further because it can
-be orders of magnitude slower than native execution, making it infeasible for
-studying large workloads or systems \cite{2kmachine}.
+that memory is only allocated when it is used. As long as the working set of
+the target system fits in the host memory, performance will be tolerable.
+However, in our work, we wish to study OS behavior under workloads that
+actually use all available memory. Thus, the usefulness of demand paging is
+greatly reduced because swapping causes disk latency to become the main
+performance bottleneck. Moreover, the performance overhead of simulators like
+Simics tends to be impractical because system state needs to be maintained by
+the simulator \cite{simics}. We do not explore simulation further because it
+can be orders of magnitude slower than native execution, making it infeasible
+for studying large workloads or systems \cite{2kmachine}. However, the Simics
+simulator illustrates the usefulness of memory management techniques such as
+demand paging and the potential overheads of swapping.
 
@@
 custom data structures. Moreover, on Linux, the kernel data structures for 1TB
-of memory may also exceed the size of physical memory on present systems
+of memory may also exceed the size of physical memory on current systems
 \cite{simics}.
@@
 swap pages to disk. The resulting poor performance can cause inaccurate
-performance measurements when running benchmarks.
+performance measurements when running benchmarks. It can also make large
+benchmarks impractical.
+
+Interestingly, many of non-volatile memory technologies are making their ways
+into storage drives, such as Intel's recently-announced Optane SSDs. One
+potential use of fast storage devices is for fast swapping space. Some
+literature has proposed that small amounts of DRAM backed by large amounts of
+fast swap space can be a scalable alternative to having large amounts of fast
+RAM (TODO: cite). (TODO: can we have this in our report?). In such a system,
+the overhead of swapping may become a bottleneck.
 
 Prefetching pages from swap space can offer a way to mitigate the overhead of
-memory overcommitment. When there is significant memory pressure, even pages
-which are likely to be accessed soon are swapped out to disk. They are faulted
-back into memory when accessed, resulting in significant performance cost.
-Charm++ uses a programming model where computations can be scheduled by the
-language runtime. A designated thread can prefetch pages required by a
-computation, averting page faults. Although this approach is effective in
-avoiding overhead, it requires applications to be written with a particular
-programming language and so is not applicable to most benchmarks without
-changing them \cite{charmpp}.
+memory overcommitment, possibly making large virtual machines a viable
+mechanism for studying large memory systems. When there is significant memory
+pressure, even pages which are likely to be accessed soon are swapped out to
+disk. They are faulted back into memory when accessed, resulting in significant
+performance cost.  Charm++ uses a programming model where computations can be
+scheduled by the language runtime. A designated thread can prefetch pages
+required by a computation, averting page faults \cite{charmpp}. Rather than
+implement page prefetching in a language runtime, one might implement a
+prefetcher in the swapping subsystem, making it general purpose. A large body
+of work already exists on hardware prefetching for processor caches, on which a
+swap prefetcher might draw for inspiration (TODO: cite). However, this approach
+can only work if page access patterns are predictable. Moreover, the high
+latency of disks implies that prefetchers would have to predict the very
+distant future (e.g. seconds ahead of execution). (TODO: Mike's comment on the
+previous version of this paragraph said that it was ``too distant/off topic''.
+Is it better now? I feel like prefetching was an idea that we considered too
+much to leave out...)
 
@@
 events are occurring faster. This allows the system to emulate network links
-with speeds that are currently not available. The implementation is based on the
-VMs and Xen hypervisor; Xen delivers the timer interrupts to the guest at a
+with speeds that are currently not available. The implementation is based on
+the VMs and Xen hypervisor; Xen delivers the timer interrupts to the guest at a
 lower rate than hardware hence  slowing down the guest’s clock \cite{timedil}.
-We may use a similar approach to slow down system time  while paging in and out
-large portions of memory. This will allow us to make the system believe that it
-is reading data from the memory while actually most of that data is being read
-from the disk.
+A similar approach may be applied for large memory system studies to slow down
+guest time while paging in and out large portions of memory. This will allows
+the system to believe that it is reading data from the memory while actually
+most of that data is being read from the disk.
 
@@
 page}, \texttt{struct vm\_area\_struct}, and page tables tend to comprise a
-large portion of memory management overhead \cite{simics}.
+large portion of memory management overhead \cite{simics} (TODO: cite the LWN
+paper).
+
+\section{Background}
+
+In our methodology, we run several experiments to measure the scalability of
+the Linux memory management system. Each experiment is designed to probe a
+different aspect of the system. This section gives some background knowledge
+necessary to understanding the experiments.
+
+\subsection{Virtual Memory Areas}
+
+A process may use the \texttt{mmap} system call to request that the kernel
+allocate part of its virtual address space. Accesses to memory addresses
+outside of allocated regions result in a segmentation fault. This makes memory
+corruption errors easier to find and debug, and increases security from memory
+corruption attacks.
+
+The Linux kernel keeps track of which portions of a process's address space
+have been allocated using the \texttt{vm\_area\_struct}. Because processes may
+map a number of regions, including shared objects, stack, heap, text, and
+anonymous regions, the scalability of this structure is critical to Linux's
+performance.
+
+When a process maps a region, a \texttt{vm\_area\_struct} is created, but no
+memory is allocated. When a process first touches a page, a page fault occurs
+and the processor traps into a page fault handler, which can allocate memory
+and set up a page table entry. This is called demand paging.
+
+For demand paging to work efficiently, the correct \texttt{vm\_area\_struct}
+needs to be efficient to find. Linux caches the most recently used region and
+organizes all regions into a balanced binary tree for fast lookup.  Moreover,
+adjacent regions with similar permissions and properties are merged in an
+effort to keep bookkeeping costs low.
+
+\subsection{Swapping}
+
+TODO
+
+\subsection{Page Frame Reclamation}
+
+TODO
 
@@
 
-Our methodology is broken into two parts. First, we measure the scalability of
+We conduct a series of experiments to measure the scalability of various parts of the Linux memory management system.
+First, we measure the scalability of
 Linux's \texttt{vm\_area\_struct}. Then, we measure the minimum amount of memory
@@
 
-TODO
+\begin{figure}
+    \includegraphics[width=\columnwidth]{memory_controled}
+    \caption{$\Delta_i$ adjusted with control for benchmarks}
+\end{figure}
+
+\begin{figure}
+    \includegraphics[width=\columnwidth]{vm_area_struct_count}
+    \caption{Number of \texttt{vm\_area\_struct}s for benchmarks}
+\end{figure}
+
+TODO: memory results
+
+TODO: time results
 

### 2017-04-20 paper.tex
Worked on background section

@@
 \section{Related Work}
-
-(TODO: do we need to refocus this section? or move parts of it to an appendix?)
 
@@
 
-TODO
+Under memory pressure, the kernel may choose to write some pages of memory to
+disk and then free those pages for use. Memory mapped files are written back to
+their backing files, but anonymous memory is written to portion of the disk
+called swap space. We focus on anonymous memory in our experiments.
+
+Before swapping a page out to disk, the kernel must reserve a slot for the page
+in the swap space, unmap it from the process's address space, and record which
+swap slot corresponds to the page.
+
+On Linux, the system may have multiple swap spaces in the form of a swap disk
+partition or of swapfiles on a filesystem. Both have a similar format on disk,
+in which the space is broken into a number of swap slots large enough for one
+page. For each of these swap spaces, the kernel keeps a swap map, which records
+the usage of each swap slot so that it can find and allocate them quickly.
+
+Unmapping a page from a process's address space involves simply marking its page
+table entry ``Not Present''. Rather than using a separate data structure, Linux
+uses the unused page table entry to record information about which swap space
+and slot is used to swap out the page.
 
@@
 
-TODO
+To determine which pages to swap out, Linux uses the Page Frame Reclamation
+subsystem. Choosing which pages to swap out is important because if the system
+swaps out pages that are about to be used, thrashing may be induced, in which
+the system wastes lots of time swapping out a page just swap it back in.
+
+The Linux Page Frame Reclamation Algorithm (PFRA) attempts to approximate LRU
+order. It uses a clock-like algorithm, in which a list of pages is scanned
+twice. Roughly speaking, if a page is not used between scans, then it is a
+condidate for reclamation.
+
+The PFRA runs primarily as a kernel thread called kswapd. kswapd runs at regular
+intervals and sleeps in between. Also, it may be invoked if the system detects
+that memory has gone below a threshold. kswapd will continue to run until enough
+pages are freed to bring the free memory pool above some threashold. Then,
+kswapd goes back to sleep.  One concern with large memory systems is that kswapd
+may expend significant CPU time scanning through hundreds of thousands of pages.
 
@@
 
-We conduct a series of experiments to measure the scalability of various parts of the Linux memory management system.
-First, we measure the scalability of
+We conduct a series of experiments to measure the scalability of various parts
+of the Linux memory management system.  First, we measure the scalability of
 Linux's \texttt{vm\_area\_struct}. Then, we measure the minimum amount of memory
 necessary for swapping.
-
-\subsection{\texttt{vm\_area\_struct} Experiments}
 
@@
 \end{figure}
+
+\subsection{\texttt{vm\_area\_struct}}
 
@@
 
-\subsection{Swapping Experiments}
+\subsection{Swapping Memory Overhead}
 
@@
 
-(TODO: optional: trying pulling out a DIMM??)
+\subsection{kswapd}
+
+TODO
 

### 2017-04-21 bench_time_mmap_anon_rss.cc
add logic for kswapd cpu consumtion

@@ -9,16 +9,18 @@
 #include "rdtsc.h"
 #include "pattern.h"
 #include "pinning.h"
-
+#include "rdmemusg.h"
 int main(int argc, char **argv) {
     set_cpu(0);
 
-	if (argc < 2) {
-        std::cout << "Usage: ./memusage num_gb" << std::endl;
+	if (argc < 3) {
+        std::cout << "Usage: ./memusage num_gb pid_of_kswapd0" << std::endl;
 		return -1;
 	}
 
+
 	size_t gbs = atoi(argv[1]);
+    int kswapd_pid = atoi(argv[2]);
     uint64_t vma_size = (uint64_t)gbs << 30;
 
     char* start_addr = (char*)mmap(NULL, vma_size, PROT_READ | PROT_WRITE, MAP_ANONYMOUS | MAP_PRIVATE, -1, 0);
@@ -49,6 +51,8 @@ int main(int argc, char **argv) {
         unsigned long long start = rdtsc();
         *addr = 'X';
         res_array[num_touched++] = rdtsc() - start;
+        if (num_touched % 80000 == 0)  // how to set this number? if this is too short elapsed ~= 0 
+        printf("kswapd cpu util %lu \n", get_CPU_usage(kswapd_pid));
 
         addr += PAGE_SIZE;
     }

### 2017-04-21 rdmemusg.cc

@@ -13,6 +13,8 @@
 
 static char buf[2048];
 static int fd = -1;
+unsigned long kswapd_total_before = 0;
+unsigned long elapsed_before = 0;
 
 unsigned long getinfo(const char *file, const char *name) {
 	int local_n;
@@ -111,3 +113,79 @@ unsigned long getkernelmem() {
 
 	return tmem - fmem - pmem;
 }
+
+unsigned long get_CPU_usage(int pid) {
+    char file [40];
+    sprintf(file, "/proc/%d/stat", pid);
+    char buffer[2048];
+	int f = open(file, O_RDONLY);
+	if (f == -1) {
+		fputs("Couldn't open stat file\n", stderr);
+        exit(1);
+	}
+
+    int bytes_read;
+    if ((bytes_read = read(f, buffer, sizeof(buffer) )) < 0) {
+        printf("failed to read stat file \n"); 
+        exit(1);
+    }
+
+    buffer[bytes_read] = '\0';
+
+    char* token;
+    char** my_tail = NULL;
+    token = strtok(buffer, " ");
+    for (int i = 0; i < 12; i++) {
+        token = strtok(NULL, " ");
+    }
+    token = strtok(NULL, " ");
+    unsigned long u_time = strtoul(token, my_tail, 10);
+    token = strtok(NULL, " ");
+    unsigned long s_time = strtoul(token, my_tail, 10);
+
+
+    close(f);
+    
+    
+    unsigned long elapsed_now = get_uptime();
+    unsigned long elapsed_delta = elapsed_now - elapsed_before;
+
+    
+    unsigned long kswapd_total_now = u_time + s_time;
+    unsigned long kswapd_delta = kswapd_total_now - kswapd_total_before;
+    elapsed_before = elapsed_now;
+    kswapd_total_before = kswapd_total_now;
+
+
+    if (elapsed_delta == 0) return 0;
+    unsigned long kswapd_cpu_util = (kswapd_delta / elapsed_delta) * 100;
+    
+    return kswapd_cpu_util;
+
+}
+
+unsigned long get_uptime() {
+    char buffer[64];
+	int f = open("/proc/uptime", O_RDONLY);
+	if (f == -1) {
+		fputs("Couldn't open uptime file\n", stderr);
+        exit(1);
+	}
+
+    int bytes_read;
+    if ((bytes_read = read(f, buffer, sizeof(buffer) )) < 0) {
[…]

### 2017-04-21 rdmemusg.h

@@ -8,6 +8,8 @@ unsigned long getinfo(const char *file, const char *name);
 unsigned long meminfo(const char *name);
 unsigned long statusinfo(const char *name);
 unsigned long slabinfo(const char *name);
+unsigned long get_CPU_usage(int);
+unsigned long get_uptime();
 
 unsigned long getkernelmem();
 

### 2017-04-21 paper.tex

@@
 
-\title{Journey: A Study of Studying Large Memory Systems (TODO: should we get a new title?)}
+\title{Journey: A Study in Unhelpful Titles (TODO: should we get a new title?)}
 \author{Mark Mansi, Suhas Pai, Hasnain Ali Pirzada}
@@
 
+(TODO: explain struct page?)
+
+We measure the memory overhead and latency of the Linux Memory Management
+system's use of \texttt{vm\_area\_struct}, which represents a region of a
+process's virtual address space.  
+We do not measure the overhead of \texttt{struct page} because it is known that
+there is one struct for each physical page. We do not measure the overhead of
+page tables because this can be calculated from the size of the virtual memory
+space used by the process.  When a program makes a memory management system
+calls such as \texttt{mmap}, \texttt{mremap}, \texttt{munmap},
+\texttt{mprotect}, \texttt{mlock}, or \texttt{madvise}, the kernel is creating,
+changing, or removing \texttt{vm\_area\_struct}s; thus, measuring the
+scalability of Linux memory management requires measuring the overhead of these
+structs.
+
 \subsection{Swapping}
@@
 
-\subsection{\texttt{vm\_area\_struct}}
-
-We measure the memory overhead and latency of the Linux Memory Management
-system's use of \texttt{vm\_area\_struct}, which represents a region of a
-process's virtual address space.  We do not measure the overhead of
-\texttt{struct page} because it is known that there is one struct for each
-physical page. We do not measure the overhead of page tables because this can be
-calculated from the size of the virtual memory space used by the process.  When
-a program makes a memory management system calls such as \texttt{mmap},
-\texttt{mremap}, \texttt{munmap}, \texttt{mprotect}, \texttt{mlock}, or
-\texttt{madvise}, the kernel is creating, changing, or removing
-\texttt{vm\_area\_struct}s; thus, measuring the scalability of Linux memory
-management requires measuring the overhead of these structs.
+The specifications of the machine we used to run our experiments are listed in
+Figure \ref{fig:specs}. We disabled Intel SpeedStep and TurboBoost and ensure
+that the processor frequency stays constant (TODO: did we do this?). All
+experiments are pinned to the same core on our machine.
 
@@
 allocation pattern. These benchmarks are designed to stress the memory
-management system in different ways, rather than simulate real workloads. For
-each workload, we measure the latency and increase in memory usage due to each
-operation in the benchmark. Notably, we do not touch the pages that we allocate,
-avoiding both a page fault and the allocation of a back physical page for the
-allocated virtual pages. This allows us to measure just the overhead of memory
-management, as opposed to physical page allocation, swapping, page faults, or
-other overheads.  Each workload contains $2^{20}$ memory management operations.
-
-The specifications of the machine are listed in Figure \ref{fig:specs}. We
-disabled Intel SpeedStep and TurboBoost. All experiments are pinned to the same
-core on our machine. Each benchmark was run 5 times and the results aggregated
-as described below. We reboot the machine when switching to a different
-benchmark, but keep the machine running continually during the runs of the same
-benchmark. Each run is executed in isolation, without any additional processes
+management system in different ways, rather than simulate real workloads. 
+Each run is executed in isolation, without any additional processes
 running beyond services that start when the system is booted and a single ssh
@@
 timestamp before and after each memory management operation in the workload and
-take the difference to get latency. The median latency of each operation is
-reported across all 5 runs.
+take the difference to get latency. (TODO: is this still true?)
 
-To measure memory, we use the procfs's reporting on current memory usage, along
-with the slab allocator's count of active \texttt{vm\_area\_struct}s in the
-system. Measuring kernel memory usage and associating memory usage with
-processes is very difficult. Our methodology assumes that our benchmarks running
-in otherwise idle systems should dominate increases in memory usage.  Thus, we
-compute total memory usage in the system as the total amount of memory less the
-amount of memory used. From this amount, we subtract the resident set size of
-the benchmark process. The result is the memory usage of the rest of the system
-$R$. We denote $R_i$ as the $i$-th measurement of $R$ during a run ($i = 0, 1,
-... 2^{20}$). System memory usage is known to jitter in Linux (TODO: cite). To
-adjust for this jitter, we define $\Delta_i = R_i - R_0$. For each operation, we
-take the median value of $\Delta_i$ across the 5 runs to reduce jitter further.
-This is the value graphed in the figures in this section. $\Delta_i$ represents
-the increase in memory usage due to memory management overheads caused by
-running $i$ memory management operations.
+To measure memory usage, we use the procfs's reporting on current memory usage.
+Where appropriate, we record the slab allocator's count of active
+\texttt{vm\_area\_struct}s in the system. Measuring kernel memory usage and
+associating memory usage with processes is difficult. Our methodology assumes
+that our benchmarks are running in an otherwise idle system and should dominate
+increases in memory usage.  We compute total memory usage by the rest of the
+system as the total amount of memory less the amount of memory used by the
+benchmark.
+
+Finally, we keep transparent huge pages on throughout our experiments, unless
+otherwise specified.
+
+\subsection{\texttt{vm\_area\_struct}}
+
+For each workload, we measure the latency and increase in memory usage due to
+each operation in the benchmark. Notably, we do not touch the pages that we
+allocate, avoiding both a page fault and the allocation of a back physical page
+for the allocated virtual pages. This allows us to measure just the overhead of
+memory management, as opposed to physical page allocation, swapping, page
+faults, or other overheads.  Each workload contains $2^{20}$ memory management
+operations.
+
+Each benchmark was run 5 times and the results aggregated as described below. We
+reboot the machine when switching to a different benchmark, but keep the machine
+running continually during the runs of the same benchmark.
+
+Assume the memory usage of the rest of the system is $R$. We denote $R_i$ as the
+$i$-th measurement of $R$ during a run ($i = 0, 1, ... 2^{20}$). System memory
+usage is known to jitter in Linux (TODO: cite). To adjust for this jitter, we
+define $\Delta_i = R_i - R_0$. For each operation, we take the median value of
+$\Delta_i$ across the 5 runs to reduce jitter further.  This is the value
+graphed in the figures in this section. $\Delta_i$ represents the increase in
+memory usage due to memory management overheads caused by running $i$ memory
+management operations.
 
@@
 
-In this experiment, a single process touches 32GB of virtual memory, leading to
-the actual allocation of physical memory. Notice that 32GB is twice as large as
-the 16GB of physical memory in the test machine, causing Linux to start paging
-after physical memory is exhausted. Also, notice that the Linux kernel will
-reserve some memory for its own operation.
+A single process allocates 18GB of virtual memory space. It then proceeds to
+write one byte of each 4KB page in the region, leading to the actual allocation
+of physical memory. Notice that 18GB exceeds the 16GB of physical memory in the
+test machine, causing Linux to start paging after physical memory is exhausted.
+Also, notice that the Linux kernel will reserve some memory for its own
+operation.
+
+We run this experiment with both a strided access pattern and a random access
+pattern. The stride access pattern simply touches pages in order of increasing
+address. The random access pattern chooses a random address. Both patterns touch
+the same number of addresses.
 
@@
 
-This experiment is repeated with 5 different swapfile sizes to capture the
-impact of swapfile size on swapping overhead.
+Moreover, we measure the latency of each memory store operation to see the
+impact of swapping and virtual memory usage on memory access times.
 
-(TODO: strided access vs random access?)
+We repeat this with 32GB of swap space and 128GB of swap space to see the
+impact of swap space size on the amount of memory the kernel reserves for
+itself, along with the impact on latency.
 
@@
 
-TODO
+These experiments are designed to measure the processing overhead of kswapd when
+the system is under high memory pressure.
+
+A single process allocates 18GB of virtual memory space. It then proceeds to
+write one byte of each 4KB page in the region, leading to the actual allocation
+of physical memory. The process chooses addresses randomly, as in the previous
+set of experiments, inducing the maximum amount of swapping.
+
+In all these experiments, the system is given 32GB of swap space total. We
+measure the impact of physical memory size and transparent huge pages. We run
+four experiments:
+
+\begin{enumerate}
+    \item 1GB physical memory with huge pages enabled
+    \item 16GB physical memory with huge pages enabled
+    \item 1GB physical memory with huge pages disabled
+    \item 16GB physical memory with huge pages disabled
+\end{enumerate}
+
+To run the 1GB memory benchmarks, we hot unplug 15GB using the sysfs in Linux.
+This makes the unplugged memory unusable by the whole system (kernel and user).
 

### 2017-04-21 bench_time_mmap_anon_rss.cc
fix cpu util calculation

@@ -52,7 +52,7 @@ int main(int argc, char **argv) {
         *addr = 'X';
         res_array[num_touched++] = rdtsc() - start;
         if (num_touched % 80000 == 0)  // how to set this number? if this is too short elapsed ~= 0 
-        printf("kswapd cpu util %lu \n", get_CPU_usage(kswapd_pid));
+            std::cout << "kswapd cpu util " << get_CPU_usage(kswapd_pid) << std::endl;
 
         addr += PAGE_SIZE;
     }

### 2017-04-21 rdmemusg.cc

@@ -147,7 +147,7 @@ unsigned long get_CPU_usage(int pid) {
     close(f);
     
     
-    unsigned long elapsed_now = get_uptime();
+    unsigned long elapsed_now = get_uptime_jiffies();
     unsigned long elapsed_delta = elapsed_now - elapsed_before;
 
     
@@ -162,6 +162,46 @@ unsigned long get_CPU_usage(int pid) {
     
     return kswapd_cpu_util;
 
+}
+
+unsigned long get_uptime_jiffies() {
+    unsigned long total_uptime = 0;
+    char buffer[1024];
+	int f = open("/proc/stat", O_RDONLY);
+	if (f == -1) {
+		fputs("Couldn't open /proc/stat file\n", stderr);
+        exit(1);
+	}
+
+    int bytes_read;
+    if ((bytes_read = read(f, buffer, sizeof(buffer) )) < 0) {
+        printf("failed to read /proc/stat file \n"); 
+        exit(1);
+    }
+    buffer[bytes_read] = '\0';
+
+    // want to get just the first line
+    for (unsigned int i = 0; i < sizeof(buffer); i++) {
+        if (buffer[i] == '\n'){
+            buffer[i] = '\0';
+            break;
+        }
+    }
+    close(f);
+
+    char** my_tail = NULL;
+    char* token = strtok(buffer, " "); // first token is "cpu" 
+
+    //  a list of various kinds of time.. 
+    //  sum all of them to get total
+    while (token) {
+        token = strtok(NULL, " ");
+        if (token != NULL)
+        total_uptime += strtoul(token, my_tail, 10);
+    
+    }
+    return total_uptime;
+    
 }
 
 unsigned long get_uptime() {

### 2017-04-21 rdmemusg.h

@@ -10,6 +10,7 @@ unsigned long statusinfo(const char *name);
 unsigned long slabinfo(const char *name);
 unsigned long get_CPU_usage(int);
 unsigned long get_uptime();
+unsigned long get_uptime_jiffies();
 
 unsigned long getkernelmem();
 

### 2017-04-21 paper.tex

@@
 
+We measure CPU time via the procfs using the same methodology as the
+\texttt{top} utility. The \texttt{/proc/[pid]/stat} file for kswapd is read.
+From this file the \texttt{utime} and \texttt{stime} fields are read and summed
+to get the total time allocated to kswapd. Then, the \texttt{/proc/stat} file is
+read to get the total system time elapsed. By dividing, we can find the portion
+of CPU time spent in kswapd.
+
 \section{Result}

### 2017-04-21 paper.tex
Updated a bunch of benchmarks

@@
 \item \textit{Fragmented} (``frag'' for short). This workload first allocates
-$2^{19} + 2^{18}$ contiguous 4KB blocks. Then, it resizes every other block to
-40KB. This represents the worst case behavior we could think of. It requires the
-kernel to find a block it has not touched in over $2^{19}$ operations and move
-it somewhere else.  \end{itemize}
+$2^{20}$ contiguous 4KB blocks. Then, it resizes every other block to
+16KB. This represents the worst case behavior we could think of. It requires the
+kernel to find a block it has not touched in $2^{20}$ operations and move
+it somewhere else.  
+\end{itemize}
 

### 2017-04-21 bench_mem_mmap_frag.cc

@@ -19,7 +19,7 @@ int main(int argc, char **argv) {
 
     size_t amt = 1 << atoi(argv[1]);
 
-    PatternGenerator pg = get_mmap_cont(4 /* 16KB */, 7 /* RWX */);
+    PatternGenerator pg = get_mmap_cont(1 /* 4KB */, 7 /* RWX */);
     Next next;
 
     std::cout << getkernelmem() << " " 
@@ -36,7 +36,7 @@ int main(int argc, char **argv) {
         }
     }
 
-    pg = get_mremap_frag(1, 4, 7);
+    pg = get_mremap_frag(2, 4, 7);
 
     for (size_t i = 0; i < amt; i++) {
         next = pg.next();

### 2017-04-21 bench_time_per_op_mmap_16k_cont.cc

@@ -1,0 +1,39 @@
+#include <iostream>
+#include <stdlib.h>
+#include <sys/mman.h>
+#include <string.h>
+#include <errno.h>
+
+#include "pattern.h"
+#include "rdtsc.h"
+#include "pinning.h"
+
+int main(int argc, char* argv[]) {
+    set_cpu(0);
+    PatternGenerator pg = get_mmap_cont(4 /* 16KB */, 7 /* RWX */);
+
+    Next cur;
+    char* addr;
+
+    if (argc < 2) {
+        std::cout << "Usage: ./pattern total_ops_order" << std::endl;
+        exit(1);
+    }
+
+    int tot_ops = 1 << atoi(argv[1]);
+    
+    for (int i = 0; i < tot_ops; i++) {
+        cur = pg.next();
+        unsigned long long start = rdtsc();
+        addr = (char *)mmap(cur.address, cur.size, cur.permissions, MAP_ANON | MAP_PRIVATE, -1, 0);
+        unsigned long long end = rdtsc();
+        if (addr == MAP_FAILED) {
+            std::cerr << "mmap failed: " << strerror(errno) << std::endl;
+            exit(1);
+        }
+
+        std::cout << (end - start) << std::endl;
+    }
+
+    return 0;
+}

### 2017-04-21 bench_time_per_op_mmap_16k_cont_touch.cc

@@ -1,51 +1,0 @@
-#include <iostream>
-#include <stdlib.h>
-#include <sys/mman.h>
-#include <string.h>
-#include <errno.h>
-
-#include "pattern.h"
-#include "rdtsc.h"
-#include "pinning.h"
-
-int main(int argc, char* argv[]) {
-    set_cpu(0);
-    PatternGenerator pg = get_mmap_cont(4 /* 16KB */, 7 /* RWX */);
-
-    Next cur;
-    char* addr;
-
-    if (argc < 2) {
-        std::cout << "Usage: ./pattern total_ops_order" << std::endl;
-        exit(1);
-    }
-
-    int tot_ops = 1 << atoi(argv[1]);
-    
-    double ticks_per_nano = get_ticks_per_nano();
-
-    unsigned long long start = rdtsc();
-    
-    for (int i = 0; i < tot_ops; i++) {
-        //  assuming all ops were sucessful
-        cur = pg.next();
-        if (cur.operation == OP_MMAP) {  // must be
-            addr = (char *)mmap(cur.address, cur.size, cur.permissions, MAP_ANON | MAP_PRIVATE, -1, 0);
-            if (addr == MAP_FAILED) {
-                std::cerr << "mmap failed: " << strerror(errno) << std::endl;
-                exit(1);
-            }
-        }
-        addr[0] = 'X'; // touch
-    }
-
-    unsigned long long end = rdtsc();
-    unsigned long long interval = end - start;
-    unsigned long long nano_secs = (double) interval / ticks_per_nano;
-
-    // nsec per operation
-    std::cout <<  nano_secs / tot_ops << std::endl;
-
-    return 0;
-
-}

### 2017-04-21 bench_time_per_op_mmap_frag.cc

@@ -20,10 +20,8 @@ int main(int argc, char **argv) {
 
     size_t amt = 1 << atoi(argv[1]);
 
-    PatternGenerator pg = get_mmap_cont(4 /* 16KB */, 7 /* RWX */);
+    PatternGenerator pg = get_mmap_cont(1 /* 4KB */, 7 /* RWX */);
     Next next;
-
-    double ticks_per_nano = get_ticks_per_nano();  // calibrate rdtsc before measuring
 
     for (size_t i = 0; i < amt; i++) {
         next = pg.next();
@@ -36,7 +34,7 @@ int main(int argc, char **argv) {
         }
     }
 
-    pg = get_mremap_frag(1, 4, 7);
+    pg = get_mremap_frag(2, 4, 7);
 
     for (size_t i = 0; i < amt; i++) {
         next = pg.next();
@@ -49,6 +47,6 @@ int main(int argc, char **argv) {
             std::cerr << "mmap failed: " << strerror(errno) << std::endl;
             return -1;
         }
-        print_interval(cur_start, cur_end, ticks_per_nano);
+        std::cout << (cur_end - cur_start) << std::endl;
     }
 }

### 2017-04-21 line_graph.py

@@ -24,6 +24,7 @@ for data_file in argv[4:]:
             if len(line.strip()) == 0:
                 continue
 
+            #xs = [float(x)/1E6 for x in line.split()]
             xs = [float(x) for x in line.split()]
 
             data[data_file].append(xs)
@@ -36,6 +37,8 @@ for data_file in data:
             #ys = [x[i] - data[data_file][0][i] for x in data[data_file]]
             # uncomment next line to prevent normalizing to X_0
             ys = [x[i] for x in data[data_file]]
+            #xs = [float(i) / 1E6 for i in range(len(ys))]
+            #line, = plt.plot(xs, ys, label = data_file)
             line, = plt.plot(ys, label = data_file)
             handles.append(line)
 
@@ -49,7 +52,7 @@ plt.xlabel(AXIS_X)
 plt.ylabel(AXIS_Y)
 plt.grid(True)
 
-plt.legend(handles = handles)
+plt.legend(handles = handles, loc='best')
 
 font = {'family' : 'normal',
         'weight' : 'bold',

### 2017-04-22 bench_mem_mmap_frag.cc
A few bug fixes?

@@ -41,7 +41,7 @@ int main(int argc, char **argv) {
     for (size_t i = 0; i < amt; i++) {
         next = pg.next();
 
-        char *addr = (char *)mremap(next.address, 2 << 12, next.size, MREMAP_MAYMOVE);
+        char *addr = (char *)mremap(next.address, 1 << 12, next.size, MREMAP_MAYMOVE);
         if (addr == MAP_FAILED) {
             std::cerr << "mmap failed: " << strerror(errno) << std::endl;
             return -1;

### 2017-04-22 bench_time_per_op_mmap_frag.cc

@@ -40,7 +40,7 @@ int main(int argc, char **argv) {
         next = pg.next();
 
         unsigned long long cur_start = rdtsc();
-        char *addr = (char *)mremap(next.address, 2 << 12, next.size, MREMAP_MAYMOVE);
+        char *addr = (char *)mremap(next.address, 1 << 12, next.size, MREMAP_MAYMOVE);
         unsigned long long cur_end = rdtsc();
 
         if (addr == MAP_FAILED) {

### 2017-04-22 pattern.cc

@@ -172,17 +172,15 @@ PatternGenerator get_mmap_small_remap_large(int small_size, char permissions, in
 // Resize pages to larger size
 PatternGenerator get_mremap_frag(unsigned long offset, int size, char permissions) {
     PatternPart first;
-    first.size = size; // size * 4KB
-    first.location = size;
+    first.size = size; // 4KB * size
+    first.location = offset;
     first.permissions = permissions;
     first.operation = OP_MREMAP;
 
     std::vector<PatternPart> pattern;
     pattern.push_back(first);
 
-    unsigned long start = (unsigned long)STARTING_POINT + (offset << 12);
-
-    return PatternGenerator(pattern, (void *)start);
+    return PatternGenerator(pattern);
 }
 
 // mmap a bunch of pages with a certain stride

### 2017-04-22 bench_mem_mmap_frag.cc
One more bug fix?

@@ -26,14 +26,12 @@ int main(int argc, char **argv) {
               << slabinfo("vm_area_struct") << " " 
               << meminfo("PageTables") << std::endl;
 
-    for (size_t i = 0; i < amt; i++) {
-        next = pg.next();
+    next = pg.next();
 
-        char *addr = (char *)mmap(next.address, next.size, next.permissions, MAP_ANON | MAP_PRIVATE, -1, 0);
-        if (addr == MAP_FAILED) {
-            std::cerr << "mmap failed: " << strerror(errno) << std::endl;
-            return -1;
-        }
+    char *addr = (char *)mmap(next.address, amt << 1, next.permissions, MAP_ANON | MAP_PRIVATE, -1, 0);
+    if (addr == MAP_FAILED) {
+        std::cerr << "mmap failed: " << strerror(errno) << std::endl;
+        return -1;
     }
 
     pg = get_mremap_frag(2, 4, 7);

### 2017-04-22 bench_time_per_op_mmap_frag.cc

@@ -23,15 +23,12 @@ int main(int argc, char **argv) {
     PatternGenerator pg = get_mmap_cont(1 /* 4KB */, 7 /* RWX */);
     Next next;
 
-    for (size_t i = 0; i < amt; i++) {
-        next = pg.next();
+    next = pg.next();
+    char *addr = (char *)mmap(next.address, amt << 1, next.permissions, MAP_ANON | MAP_PRIVATE, -1, 0);
 
-        char *addr = (char *)mmap(next.address, next.size, next.permissions, MAP_ANON | MAP_PRIVATE, -1, 0);
-
-        if (addr == MAP_FAILED) {
-            std::cerr << "mmap failed: " << strerror(errno) << std::endl;
-            return -1;
-        }
+    if (addr == MAP_FAILED) {
+        std::cerr << "mmap failed: " << strerror(errno) << std::endl;
+        return -1;
     }
 
     pg = get_mremap_frag(2, 4, 7);

### 2017-04-22 bench_mem_mmap_frag.cc
More bugs

@@ -28,7 +28,7 @@ int main(int argc, char **argv) {
 
     next = pg.next();
 
-    char *addr = (char *)mmap(next.address, amt << 1, next.permissions, MAP_ANON | MAP_PRIVATE, -1, 0);
+    char *addr = (char *)mmap(next.address, amt << 12 << 1, next.permissions, MAP_ANON | MAP_PRIVATE, -1, 0);
     if (addr == MAP_FAILED) {
         std::cerr << "mmap failed: " << strerror(errno) << std::endl;
         return -1;
@@ -42,6 +42,7 @@ int main(int argc, char **argv) {
         char *addr = (char *)mremap(next.address, 1 << 12, next.size, MREMAP_MAYMOVE);
         if (addr == MAP_FAILED) {
             std::cerr << "mmap failed: " << strerror(errno) << std::endl;
+            next.print();
             return -1;
         }
         std::cout << getkernelmem() << " " 

### 2017-04-22 bench_time_per_op_mmap_frag.cc

@@ -24,7 +24,7 @@ int main(int argc, char **argv) {
     Next next;
 
     next = pg.next();
-    char *addr = (char *)mmap(next.address, amt << 1, next.permissions, MAP_ANON | MAP_PRIVATE, -1, 0);
+    char *addr = (char *)mmap(next.address, amt << 12 << 1, next.permissions, MAP_ANON | MAP_PRIVATE, -1, 0);
 
     if (addr == MAP_FAILED) {
         std::cerr << "mmap failed: " << strerror(errno) << std::endl;

### 2017-04-22 bench_measure_cpu_util_cpu.cc
refactor cpu util logic to new process

@@ -1,0 +1,39 @@
+#include <iostream>
+#include <stdlib.h>
+#include <sys/mman.h>
+#include <string.h>
+#include <errno.h>
+#include <stdint.h>
+#include <unistd.h>
+#include <signal.h>
+
+#include "rdtsc.h"
+#include "pattern.h"
+#include "pinning.h"
+#include "rdmemusg.h"
+
+void on_stop(int signum) {
+    //std::cout << "bye!" << std::endl;
+    exit(0);
+}
+
+
+int main(int argc, char **argv) {
+    set_cpu(1);  //  run it on a different cpu from the actual benchmark. no context switches
+
+    signal(SIGINT, on_stop);
+
+	if (argc < 2) {
+        std::cout << "Usage: ./executable pid_of_process" << std::endl;
+		return -1;
+	}
+
+    int kswapd_pid = atoi(argv[1]);
+
+    // touch all the pages now
+    while (1) {
+        std::cout << get_CPU_usage(kswapd_pid) << std::endl;
+        usleep(500);  // I don't know how to tune this time. It makes a difference. 
+    }
+
+}

### 2017-04-22 rdmemusg.cc

@@ -131,31 +131,28 @@ unsigned long get_CPU_usage(int pid) {
     }
 
     buffer[bytes_read] = '\0';
+    close(f);
 
     char* token;
     char** my_tail = NULL;
     token = strtok(buffer, " ");
+
     for (int i = 0; i < 12; i++) {
         token = strtok(NULL, " ");
     }
+    
     token = strtok(NULL, " ");
     unsigned long u_time = strtoul(token, my_tail, 10);
     token = strtok(NULL, " ");
     unsigned long s_time = strtoul(token, my_tail, 10);
 
-
-    close(f);
-    
-    
     unsigned long elapsed_now = get_uptime_jiffies();
     unsigned long elapsed_delta = elapsed_now - elapsed_before;
-
-    
     unsigned long kswapd_total_now = u_time + s_time;
     unsigned long kswapd_delta = kswapd_total_now - kswapd_total_before;
+
     elapsed_before = elapsed_now;
     kswapd_total_before = kswapd_total_now;
-
 
     if (elapsed_delta == 0) return 0;
     unsigned long kswapd_cpu_util = (kswapd_delta / elapsed_delta) * 100;
@@ -179,6 +176,7 @@ unsigned long get_uptime_jiffies() {
         exit(1);
     }
     buffer[bytes_read] = '\0';
+    close(f);
 
     // want to get just the first line
     for (unsigned int i = 0; i < sizeof(buffer); i++) {
@@ -187,7 +185,6 @@ unsigned long get_uptime_jiffies() {
             break;
         }
     }
-    close(f);
 
     char** my_tail = NULL;
     char* token = strtok(buffer, " "); // first token is "cpu" 
@@ -225,7 +222,3 @@ unsigned long get_uptime() {
     unsigned long uptime = strtoul(token, my_tail, 10);
     return uptime; 
 }
-
-
-
-

### 2017-04-22 bench_cpu_kswapd_anon.cc
kswapd measurements code

@@ -1,0 +1,58 @@
+
+#include <iostream>
+#include <stdlib.h>
+#include <sys/mman.h>
+#include <string.h>
+#include <errno.h>
+#include <stdint.h>
+
+#include "rdmemusg.h"
+#include "pattern.h"
+#include "pinning.h"
+
+int main(int argc, char **argv) {
+    set_cpu(0);
+
+	if (argc < 3) {
+        std::cout << "Usage: ./prog num_gb kswapd_pid" << std::endl;
+		return -1;
+	}
+
+	size_t gbs = atoi(argv[1]);
+	size_t kswapd = atoi(argv[2]);
+    uint64_t vma_size = (uint64_t)gbs << 30;
+
+    char* start_addr = (char*)mmap(NULL, vma_size, PROT_READ | PROT_WRITE, MAP_ANONYMOUS | MAP_PRIVATE, -1, 0);
+    if (start_addr == MAP_FAILED) {
+        std::cerr << "mmap failed: " << strerror(errno) << std::endl;
+        exit(1);
+    }
+
+    unsigned long long num_pages_mapped = vma_size / PAGE_SIZE;
+    double* res_array = (double*)malloc(num_pages_mapped * sizeof(double));
+
+    // touch all pages
+    for(unsigned long i = 0; i < num_pages_mapped; i++) {
+        res_array[i] = 0;
+    }
+
+    // lock this results array itself so itself is not swapped out
+    if(mlock((const void*) res_array, num_pages_mapped * sizeof(double))) {
+        std::cerr << "mlock failed: " << strerror(errno) << std::endl;
+        exit(1);
+    }
+    
+    // touch all the pages now
+    unsigned long long num_touched = 0;
+    char* addr = start_addr;
+    while (num_touched < num_pages_mapped) {
+        *addr = 'X';
+        res_array[num_touched++] = get_CPU_usage(kswapd);
+
+        addr += PAGE_SIZE;
+    }
+
+    for (unsigned int i = 0; i < num_pages_mapped; i++) {
+        std::cout << res_array[i] << std::endl;
+    }
+}

### 2017-04-22 bench_cpu_kswapd_anon_rand.cc

@@ -1,0 +1,75 @@
+
+#include <iostream>
+#include <stdlib.h>
+#include <sys/mman.h>
+#include <string.h>
+#include <errno.h>
+#include <stdint.h>
+
+#include <cstdlib>
+#include <ctime>
+
+#include "rdmemusg.h"
+#include "pattern.h"
+#include "pinning.h"
+
+// returns a random unsigned long with upto 45-bits of pseudo-randomness
+unsigned long random_number(unsigned long min, unsigned long max) {
+        int rand_hi = rand();
+        int rand_mi = rand();
+        int rand_lo = rand();
+        unsigned long rand_45 = (((unsigned long)rand_hi) << 30) ^ (((unsigned long)rand_mi) << 15) ^ rand_lo;
+
+        return rand_45 % max;
+}
+
+int main(int argc, char **argv) {
+    set_cpu(0);
+
+	if (argc < 3) {
+        std::cout << "Usage: ./prog num_gb kswapd_pid" << std::endl;
+		return -1;
+	}
+
+	size_t gbs = atoi(argv[1]);
+	size_t kswapd = atoi(argv[2]);
+    uint64_t vma_size = (uint64_t)gbs << 30;
+
+    char* start_addr = (char*)mmap(NULL, vma_size, PROT_READ | PROT_WRITE, MAP_ANONYMOUS | MAP_PRIVATE, -1, 0);
+    if (start_addr == MAP_FAILED) {
+        std::cerr << "mmap failed: " << strerror(errno) << std::endl;
+        exit(1);
+    }
+
+    unsigned long long num_pages_mapped = vma_size / PAGE_SIZE;
+    double* res_array = (double*)malloc(num_pages_mapped * sizeof(double));
+
+    // touch all pages
+    for(unsigned long i = 0; i < num_pages_mapped; i++) {
+        res_array[i] = 0;
+    }
+
+    // lock this results array itself so itself is not swapped out
+    if(mlock((const void*) res_array, num_pages_mapped * sizeof(double))) {
+        std::cerr << "mlock failed: " << strerror(errno) << std::endl;
+        exit(1);
+    }
+
+    // seed the rng
+	srand((int)time(0));
+    
+    // touch all the pages now
+    unsigned long long num_touched = 0;
+    char* addr = start_addr;
+    while (num_touched < num_pages_mapped) {
+        *addr = 'X';
+        res_array[num_touched++] = get_CPU_usage(kswapd);
+
[…]

### 2017-04-22 rdmemusg.cc

@@ -114,10 +114,12 @@ unsigned long getkernelmem() {
 	return tmem - fmem - pmem;
 }
 
-unsigned long get_CPU_usage(int pid) {
+double get_CPU_usage(int pid) {
     char file [40];
     sprintf(file, "/proc/%d/stat", pid);
+
     char buffer[2048];
+
 	int f = open(file, O_RDONLY);
 	if (f == -1) {
 		fputs("Couldn't open stat file\n", stderr);
@@ -143,25 +145,18 @@ unsigned long get_CPU_usage(int pid) {
     token = strtok(NULL, " ");
     unsigned long s_time = strtoul(token, my_tail, 10);
 
-
     close(f);
-    
     
     unsigned long elapsed_now = get_uptime_jiffies();
     unsigned long elapsed_delta = elapsed_now - elapsed_before;
-
     
     unsigned long kswapd_total_now = u_time + s_time;
     unsigned long kswapd_delta = kswapd_total_now - kswapd_total_before;
     elapsed_before = elapsed_now;
     kswapd_total_before = kswapd_total_now;
 
-
     if (elapsed_delta == 0) return 0;
-    unsigned long kswapd_cpu_util = (kswapd_delta / elapsed_delta) * 100;
-    
-    return kswapd_cpu_util;
-
+    return (((double)kswapd_delta) / ((double)elapsed_delta)) * 100.0;
 }
 
 unsigned long get_uptime_jiffies() {

### 2017-04-22 rdmemusg.h

@@ -8,7 +8,7 @@ unsigned long getinfo(const char *file, const char *name);
 unsigned long meminfo(const char *name);
 unsigned long statusinfo(const char *name);
 unsigned long slabinfo(const char *name);
-unsigned long get_CPU_usage(int);
+double get_CPU_usage(int);
 unsigned long get_uptime();
 unsigned long get_uptime_jiffies();
 

### 2017-04-22 paper.tex

@@
 \item \textit{Fragmented} (``frag'' for short). This workload first allocates
-$2^{20}$ contiguous 4KB blocks. Then, it resizes every other block to
-16KB. This represents the worst case behavior we could think of. It requires the
-kernel to find a block it has not touched in $2^{20}$ operations and move
-it somewhere else.  
+    $4^{20}$ contiguous 4KB blocks. Then, it resizes every other block to 16KB.
+        The resizing operation, rather than the initial allocation is measured.
 \end{itemize}

### 2017-04-25 medians.py
Moved to figures dir

@@ -28,9 +28,9 @@ for data_file in argv[3:]:
 for row in range(len(data.values()[0])):
     medians_row = []
     for i in WHICH:
-        vals = [data[data_file][row][i] - data[data_file][0][i] for data_file in data]
+        #vals = [data[data_file][row][i] - data[data_file][0][i] for data_file in data]
         # uncomment next line to prevent normalizing to X_0
-        #vals = [data[data_file][row][i] for data_file in data]
+        vals = [data[data_file][row][i] for data_file in data]
         medians_row.append(np.median(vals))
     medians.append(medians_row)
 

### 2017-04-25 paper.tex
Moved to figures dir
Added email addresses to paper
assigned results section
Updated figures

@@
 \title{Journey: A Study in Unhelpful Titles (TODO: should we get a new title?)}
-\author{Mark Mansi, Suhas Pai, Hasnain Ali Pirzada}
+\author{Mark Mansi, Suhas Pai, Hasnain Ali Pirzada\\\texttt{\{markm, hp, spai\}@cs.wisc.edu}}
 \date{}
@@
 \begin{figure}
-    \includegraphics[width=\columnwidth]{memory_controled}
+    \includegraphics[width=\columnwidth]{figures/mmap_mem_usage}
     \caption{$\Delta_i$ adjusted with control for benchmarks}
@@
 \begin{figure}
-    \includegraphics[width=\columnwidth]{vm_area_struct_count}
+    \includegraphics[width=\columnwidth]{figures/vm_area_struct_count}
     \caption{Number of \texttt{vm\_area\_struct}s for benchmarks}
@@
 
-TODO: memory results
+TODO: memory results: SPAI
 
-TODO: time results
+TODO: time results: HP
 
@@
 
-TODO
+TODO: memory: SPAI
+
+TODO: time: HP/MARKM
+
+\begin{figure}
+    \includegraphics[width=\columnwidth]{figures/swap_touch_time_cont_cdf}
+    \caption{CDF Time to touch a page while swapping for continuous benchmark}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=\columnwidth]{figures/swap_touch_time_rand_cdf}
+    \caption{CDF Time to touch a page while swapping for random benchmark}
+\end{figure}
+
+\subsection{kswapd}
+
+TODO: time: MARKM
 

### 2017-04-25 paper.tex
Added a bunch of figures

@@
 $\Delta_i$ across the 5 runs to reduce jitter further.  This is the value
-graphed in the figures in this section. $\Delta_i$ represents the increase in
-memory usage due to memory management overheads caused by running $i$ memory
-management operations.
+graphed in the figures in the following section. $\Delta_i$ represents the
+increase in memory usage due to memory management overheads caused by running
+$i$ memory management operations.
 
@@
 
+\subsubsection{Time (TODO: HP)}
+
+\begin{figure}
+    \includegraphics[width=\columnwidth]{figures/mmap_latency}
+    \caption{Latency of memory operations}
+\end{figure}
+
+\subsubsection{Memory (TODO: SPAI)}
 \begin{figure}
@@
 
-TODO: memory results: SPAI
-
-TODO: time results: HP
-
 \subsection{Swapping Experiments}
 
-TODO: memory: SPAI
+\subsubsection{Memory (TODO: SPAI)}
 
-TODO: time: HP/MARKM
+\begin{figure}
+    \includegraphics[width=\columnwidth]{figures/swap_rss}
+    \caption{RSS while touching pages while swapping for random benchmark.}
+\end{figure}
+
+\subsubsection{Time (TODO: HP/MARKM)}
 
@@
     \includegraphics[width=\columnwidth]{figures/swap_touch_time_cont_cdf}
-    \caption{CDF Time to touch a page while swapping for continuous benchmark}
+    \caption{CDF of time to touch a page while swapping for continuous
+    benchmark. Notice that the y-axis scale is from 99.75\% to 100\%.}
 \end{figure}
+
 \begin{figure}
     \includegraphics[width=\columnwidth]{figures/swap_touch_time_rand_cdf}
-    \caption{CDF Time to touch a page while swapping for random benchmark}
+    \caption{CDF of time to touch a page while swapping for random benchmark.
+    Notice that the y-axis scale is from 93\% to 100\%.}
 \end{figure}
+
+\begin{figure}
+    \includegraphics[width=\columnwidth]{figures/swap_time_cont}
+    \caption{Time to touch a page while swapping for continuous benchmark.
+    Notice that the y-scale is logarithmic.}
+\end{figure}
+
+\begin{figure}
+    \includegraphics[width=\columnwidth]{figures/swap_time_rand}
+    \caption{Time to touch a page while swapping for random benchmark.}
+\end{figure}
+
+This experiment is designed to measure the impact of swapping on memory
+operation latency.
 

### 2017-04-25 bench_cpu_kswapd_anon.cc
Fixed bench_cpu_kswapd* bmks

@@ -29,15 +29,21 @@ int main(int argc, char **argv) {
     }
 
     unsigned long long num_pages_mapped = vma_size / PAGE_SIZE;
-    double* res_array = (double*)malloc(num_pages_mapped * sizeof(double));
+    unsigned long* total_jiffies = (unsigned long*)malloc(num_pages_mapped * sizeof(unsigned long));
+    unsigned long* pid_jiffies = (unsigned long*)malloc(num_pages_mapped * sizeof(unsigned long));
 
     // touch all pages
     for(unsigned long i = 0; i < num_pages_mapped; i++) {
-        res_array[i] = 0;
+        total_jiffies[i] = 0;
+        pid_jiffies[i] = 0;
     }
 
     // lock this results array itself so itself is not swapped out
-    if(mlock((const void*) res_array, num_pages_mapped * sizeof(double))) {
+    if(mlock((const void*) total_jiffies, num_pages_mapped * sizeof(double))) {
+        std::cerr << "mlock failed: " << strerror(errno) << std::endl;
+        exit(1);
+    }
+    if(mlock((const void*) pid_jiffies, num_pages_mapped * sizeof(double))) {
         std::cerr << "mlock failed: " << strerror(errno) << std::endl;
         exit(1);
     }
@@ -47,12 +53,12 @@ int main(int argc, char **argv) {
     char* addr = start_addr;
     while (num_touched < num_pages_mapped) {
         *addr = 'X';
-        res_array[num_touched++] = get_CPU_usage(kswapd);
-
+        total_jiffies[num_touched++] = get_uptime_jiffies();
+        pid_jiffies[num_touched++] = get_pid_jiffies(kswapd);
         addr += PAGE_SIZE;
     }
 
     for (unsigned int i = 0; i < num_pages_mapped; i++) {
-        std::cout << res_array[i] << std::endl;
+        std::cout << total_jiffies[i] << " " << pid_jiffies[i] << std::endl;
     }
 }

### 2017-04-25 bench_cpu_kswapd_anon_rand.cc

@@ -42,15 +42,21 @@ int main(int argc, char **argv) {
     }
 
     unsigned long long num_pages_mapped = vma_size / PAGE_SIZE;
-    double* res_array = (double*)malloc(num_pages_mapped * sizeof(double));
+    unsigned long* total_jiffies = (unsigned long*)malloc(num_pages_mapped * sizeof(unsigned long));
+    unsigned long* pid_jiffies = (unsigned long*)malloc(num_pages_mapped * sizeof(unsigned long));
 
     // touch all pages
     for(unsigned long i = 0; i < num_pages_mapped; i++) {
-        res_array[i] = 0;
+        total_jiffies[i] = 0;
+        pid_jiffies[i] = 0;
     }
 
     // lock this results array itself so itself is not swapped out
-    if(mlock((const void*) res_array, num_pages_mapped * sizeof(double))) {
+    if(mlock((const void*) total_jiffies, num_pages_mapped * sizeof(double))) {
+        std::cerr << "mlock failed: " << strerror(errno) << std::endl;
+        exit(1);
+    }
+    if(mlock((const void*) pid_jiffies, num_pages_mapped * sizeof(double))) {
         std::cerr << "mlock failed: " << strerror(errno) << std::endl;
         exit(1);
     }
@@ -63,13 +69,13 @@ int main(int argc, char **argv) {
     char* addr = start_addr;
     while (num_touched < num_pages_mapped) {
         *addr = 'X';
-        res_array[num_touched++] = get_CPU_usage(kswapd);
-
+        total_jiffies[num_touched++] = get_uptime_jiffies();
+        pid_jiffies[num_touched++] = get_pid_jiffies(kswapd);
         addr = ((char*)start_addr) + random_number(0, vma_size);
         //printf("%lx\n", (unsigned long)addr);
     }
 
     for (unsigned int i = 0; i < num_pages_mapped; i++) {
-        std::cout << res_array[i] << std::endl;
+        std::cout << total_jiffies[i] << " " << pid_jiffies[i] << std::endl;
     }
 }

### 2017-04-25 rdmemusg.cc

@@ -114,7 +114,7 @@ unsigned long getkernelmem() {
 	return tmem - fmem - pmem;
 }
 
-double get_CPU_usage(int pid) {
+unsigned long get_pid_jiffies(int pid) {
     char file [40];
     sprintf(file, "/proc/%d/stat", pid);
 
@@ -148,18 +148,12 @@ double get_CPU_usage(int pid) {
     unsigned long s_time = strtoul(token, my_tail, 10);
 
     close(f);
-    
-    unsigned long elapsed_now = get_uptime_jiffies();
-    unsigned long elapsed_delta = elapsed_now - elapsed_before;
 
     unsigned long kswapd_total_now = u_time + s_time;
     unsigned long kswapd_delta = kswapd_total_now - kswapd_total_before;
-
-    elapsed_before = elapsed_now;
     kswapd_total_before = kswapd_total_now;
 
-    if (elapsed_delta == 0) return 0;
-    return (((double)kswapd_delta) / ((double)elapsed_delta)) * 100.0;
+    return kswapd_delta;
 }
 
 unsigned long get_uptime_jiffies() {
@@ -198,8 +192,11 @@ unsigned long get_uptime_jiffies() {
         total_uptime += strtoul(token, my_tail, 10);
     
     }
-    return total_uptime;
-    
+
+    unsigned long elapsed_delta = total_uptime - elapsed_before;
+    elapsed_before = total_uptime;
+
+    return elapsed_delta;
 }
 
 unsigned long get_uptime() {

### 2017-04-25 rdmemusg.h

@@ -8,9 +8,9 @@ unsigned long getinfo(const char *file, const char *name);
 unsigned long meminfo(const char *name);
 unsigned long statusinfo(const char *name);
 unsigned long slabinfo(const char *name);
-double get_CPU_usage(int);
-unsigned long get_uptime();
+unsigned long get_pid_jiffies(int);
 unsigned long get_uptime_jiffies();
+unsigned long get_uptime();
 
 unsigned long getkernelmem();
 

### 2017-04-25 rdmemusg.cc
Added back get_cpu_usage

@@ -114,6 +114,51 @@ unsigned long getkernelmem() {
 	return tmem - fmem - pmem;
 }
 
+double get_pid_cpu_usage(int pid) {
+    char file [40];
+    sprintf(file, "/proc/%d/stat", pid);
+
+    char buffer[2048];
+
+	int f = open(file, O_RDONLY);
+	if (f == -1) {
+		fputs("Couldn't open stat file\n", stderr);
+        exit(1);
+	}
+
+    int bytes_read;
+    if ((bytes_read = read(f, buffer, sizeof(buffer) )) < 0) {
+        printf("failed to read stat file \n"); 
+        exit(1);
+    }
+
+    buffer[bytes_read] = '\0';
+
+    char* token;
+    char** my_tail = NULL;
+    token = strtok(buffer, " ");
+
+    for (int i = 0; i < 12; i++) {
+        token = strtok(NULL, " ");
+    }
+    
+    token = strtok(NULL, " ");
+    unsigned long u_time = strtoul(token, my_tail, 10);
+    token = strtok(NULL, " ");
+    unsigned long s_time = strtoul(token, my_tail, 10);
+
+    close(f);
+
+    unsigned long elapsed_delta = get_uptime_jiffies();
+
+    unsigned long kswapd_total_now = u_time + s_time;
+    unsigned long kswapd_delta = kswapd_total_now - kswapd_total_before;
+    kswapd_total_before = kswapd_total_now;
+
+    if (elapsed_delta == 0) return 0;
+    return ((double)kswapd_delta) / ((double) elapsed_delta);
+}
+
 unsigned long get_pid_jiffies(int pid) {
     char file [40];
     sprintf(file, "/proc/%d/stat", pid);

### 2017-04-25 rdmemusg.h

@@ -8,6 +8,7 @@ unsigned long getinfo(const char *file, const char *name);
 unsigned long meminfo(const char *name);
 unsigned long statusinfo(const char *name);
 unsigned long slabinfo(const char *name);
+unsigned long get_pid_cpu_usage(int);
 unsigned long get_pid_jiffies(int);
 unsigned long get_uptime_jiffies();
 unsigned long get_uptime();

### 2017-04-26 bench_measure_cpu_util_cpu.cc
Minor fix

@@ -32,7 +32,7 @@ int main(int argc, char **argv) {
 
     // touch all the pages now
     while (1) {
-        std::cout << get_CPU_usage(kswapd_pid) << std::endl;
+        std::cout << get_pid_cpu_usage(kswapd_pid) << std::endl;
         usleep(500);  // I don't know how to tune this time. It makes a difference. 
     }
 

### 2017-04-26 rdmemusg.h

@@ -8,7 +8,7 @@ unsigned long getinfo(const char *file, const char *name);
 unsigned long meminfo(const char *name);
 unsigned long statusinfo(const char *name);
 unsigned long slabinfo(const char *name);
-unsigned long get_pid_cpu_usage(int);
+double get_pid_cpu_usage(int);
 unsigned long get_pid_jiffies(int);
 unsigned long get_uptime_jiffies();
 unsigned long get_uptime();

### 2017-04-26 paper.tex
Working on results section

@@
 
-\subsubsection{Time (TODO: HP/MARKM)}
+\subsubsection{Time}
 
@@
     \caption{CDF of time to touch a page while swapping for continuous
-    benchmark. Notice that the y-axis scale is from 99.75\% to 100\%.}
+    benchmark. Notice that the y-axis scale is from 99.75\% to 100\%.
+    \label{fig:swap_time_cont_cdf}}
 \end{figure}
@@
     \caption{CDF of time to touch a page while swapping for random benchmark.
-    Notice that the y-axis scale is from 93\% to 100\%.}
+    Notice that the y-axis scale is from 93\% to 100\%.\label{fig:swap_time_rand_cdf}}
 \end{figure}
@@
     \caption{Time to touch a page while swapping for continuous benchmark.
-    Notice that the y-scale is logarithmic.}
+    Notice that the y-scale is logarithmic.\label{fig:swap_time_cont}}
 \end{figure}
@@
     \includegraphics[width=\columnwidth]{figures/swap_time_rand}
-    \caption{Time to touch a page while swapping for random benchmark.}
+    \caption{Time to touch a page while swapping for random benchmark.\label{fig:swap_time_rand}}
 \end{figure}
@@
 This experiment is designed to measure the impact of swapping on memory
-operation latency.
+operation latency. Figure \ref{fig:swap_time_cont} and Figure
+\ref{fig:swap_time_rand} show the results.
+
+The overall behavior is as expected. In the random benchmark, as before, all
+memory is exhausted quickly, and the benchmark induces swapping soon after the
+benchmark begins. Many memory accesses thus take milliseconds to complete
+because they require paging in or out, which in turn depends on disk latency.
+Even so, the majority of benchmark run time comes from a minority of accesses.
+Figure \ref{fig:swap_time_rand_cdf} shows that as much as 95\% of memory
+accesses occur on the micro-second scale. Of the remaining 5\%, almost all
+accesses complete in less than 20ms. Moreover, notice that the size of the
+swapfile makes very little difference in this workload because the disk has to
+do random seeks regardless of swapfile size.
+
+The continuous benchmark has more surprising results. Notice that Figure
+\ref{fig:swap_time_cont} has a logarithmic y-scale. The system only begins to
+swap after about 4M operations, as before. Unlike the random benchmark, we see a
+number of operations that take hundreds of milliseconds. Figure
+\ref{fig:swap_time_cont_cdf} shows that these only form a fraction of a percent
+of all accesses done after swapping starts. However, Figure (TODO: figure) shows
+that these few accesses form a majority of run time; they impact performance
+significantly. (TODO: why do they happen).
+
+We also find that in the continuous benchmark, a larger swapfile actually has
+slight more of these slow memory accesses. We believe this has to do with the
+fact that the large swapfile has less disk locality, and so causes more disk
+seeks when paging out.
+
+~\\ \textbf{Conclusions} TODO
+
+Large swapfiles should be created with disk locality or on SSDs.
 

### 2017-04-26 paper.tex
Added awesome graph

@@
 
+\begin{figure}
+    \includegraphics[width=\columnwidth]{figures/swap_time_cdf}
+    \caption{CDF of amount of time spent on operations with a given
+    latency.\label{fig:swap_time_cdf}}
+\end{figure}
+
 This experiment is designed to measure the impact of swapping on memory
@@
 \ref{fig:swap_time_cont_cdf} shows that these only form a fraction of a percent
-of all accesses done after swapping starts. However, Figure (TODO: figure) shows
-that these few accesses form a majority of run time; they impact performance
-significantly. (TODO: why do they happen).
+of all accesses done after swapping starts. However, Figure
+\ref{fig:swap_time_cdf} shows that these few accesses form a majority of run
+time; they impact performance significantly. (TODO: why do they happen?).
 
@@
 fact that the large swapfile has less disk locality, and so causes more disk
-seeks when paging out.
+seeks when paging out. Figure \ref{fig:swap_time_cdf} shows that with a larger
+swapfile, the continuous benchmark has a significantly higher number of
+high-latency accesses as compared with the smaller swapfile.
 
-~\\ \textbf{Conclusions} TODO
+~\\ \textbf{Conclusions}
 
 Large swapfiles should be created with disk locality or on SSDs.
+
+Looking at 99.99-th percentile latencies is important because they can have a
+lot of impact on overall run time.
+
+TODO: what else?
 

### 2017-04-26 bench_cpu_kswapd_anon.cc
Fixed a couple of bugs

@@ -49,16 +49,15 @@ int main(int argc, char **argv) {
     }
     
     // touch all the pages now
-    unsigned long long num_touched = 0;
     char* addr = start_addr;
-    while (num_touched < num_pages_mapped) {
+    while (unsigned long num_touched = 0; num_touched < num_pages_mapped; num_touched++) {
         *addr = 'X';
-        total_jiffies[num_touched++] = get_uptime_jiffies();
-        pid_jiffies[num_touched++] = get_pid_jiffies(kswapd);
+        total_jiffies[num_touched] = get_uptime_jiffies();
+        pid_jiffies[num_touched] = get_pid_jiffies(kswapd);
         addr += PAGE_SIZE;
     }
 
-    for (unsigned int i = 0; i < num_pages_mapped; i++) {
+    for (unsigned long i = 0; i < num_pages_mapped; i++) {
         std::cout << total_jiffies[i] << " " << pid_jiffies[i] << std::endl;
     }
 }

### 2017-04-26 bench_cpu_kswapd_anon_rand.cc

@@ -65,12 +65,11 @@ int main(int argc, char **argv) {
 	srand((int)time(0));
     
     // touch all the pages now
-    unsigned long long num_touched = 0;
     char* addr = start_addr;
-    while (num_touched < num_pages_mapped) {
+    while (unsigned long num_touched = 0; num_touched < num_pages_mapped; num_touched++) {
         *addr = 'X';
-        total_jiffies[num_touched++] = get_uptime_jiffies();
-        pid_jiffies[num_touched++] = get_pid_jiffies(kswapd);
+        total_jiffies[num_touched] = get_uptime_jiffies();
+        pid_jiffies[num_touched] = get_pid_jiffies(kswapd);
         addr = ((char*)start_addr) + random_number(0, vma_size);
         //printf("%lx\n", (unsigned long)addr);
     }

### 2017-04-26 rdmemusg.cc

@@ -115,45 +115,8 @@ unsigned long getkernelmem() {
 }
 
 double get_pid_cpu_usage(int pid) {
-    char file [40];
-    sprintf(file, "/proc/%d/stat", pid);
-
-    char buffer[2048];
-
-	int f = open(file, O_RDONLY);
-	if (f == -1) {
-		fputs("Couldn't open stat file\n", stderr);
-        exit(1);
-	}
-
-    int bytes_read;
-    if ((bytes_read = read(f, buffer, sizeof(buffer) )) < 0) {
-        printf("failed to read stat file \n"); 
-        exit(1);
-    }
-
-    buffer[bytes_read] = '\0';
-
-    char* token;
-    char** my_tail = NULL;
-    token = strtok(buffer, " ");
-
-    for (int i = 0; i < 12; i++) {
-        token = strtok(NULL, " ");
-    }
-    
-    token = strtok(NULL, " ");
-    unsigned long u_time = strtoul(token, my_tail, 10);
-    token = strtok(NULL, " ");
-    unsigned long s_time = strtoul(token, my_tail, 10);
-
-    close(f);
-
     unsigned long elapsed_delta = get_uptime_jiffies();
-
-    unsigned long kswapd_total_now = u_time + s_time;
-    unsigned long kswapd_delta = kswapd_total_now - kswapd_total_before;
-    kswapd_total_before = kswapd_total_now;
+    unsigned long kswapd_delta = get_pid_jiffies(pid);
 
     if (elapsed_delta == 0) return 0;
     return ((double)kswapd_delta) / ((double) elapsed_delta);

### 2017-04-26 bench_cpu_kswapd_anon.cc
fix

@@ -50,7 +50,7 @@ int main(int argc, char **argv) {
     
     // touch all the pages now
     char* addr = start_addr;
-    while (unsigned long num_touched = 0; num_touched < num_pages_mapped; num_touched++) {
+    for (unsigned long num_touched = 0; num_touched < num_pages_mapped; num_touched++) {
         *addr = 'X';
         total_jiffies[num_touched] = get_uptime_jiffies();
         pid_jiffies[num_touched] = get_pid_jiffies(kswapd);

### 2017-04-26 bench_cpu_kswapd_anon_rand.cc

@@ -66,7 +66,7 @@ int main(int argc, char **argv) {
     
     // touch all the pages now
     char* addr = start_addr;
-    while (unsigned long num_touched = 0; num_touched < num_pages_mapped; num_touched++) {
+    for (unsigned long num_touched = 0; num_touched < num_pages_mapped; num_touched++) {
         *addr = 'X';
         total_jiffies[num_touched] = get_uptime_jiffies();
         pid_jiffies[num_touched] = get_pid_jiffies(kswapd);

### 2017-04-27 paper.tex
add latency explanation

@@
 
+Figure \ref{fig:mmap_latency} shows for various microbenchmarks the time taken per mmap operation 
+as it varies with the total amount memory allocate so far in the system. As described above, we use 
+rdtsc to find the number of ticks taken for the operation and then calibrate rdtsc to conver
+these ticks into microseconds. It can be seen that for continuous mmaps the time is not only the lowest
+but also invariant with the total number of mmaps done so far. This is because of two reasons. First, 
+when a continuous mmap (with the same permissions) happens the kernel memory allocator does not need 
+create a new vm area struct, instead it modifies the adjacent vm area structs to account for the newly 
+accommodate memory. This ensures that there is no need to perform the lookup on the red black trees that 
+store the vm area sturct so a constant line is seen.
+
 \begin{figure}
@@
     \caption{Latency of memory operations}
+    \label{fig:mmap_latency}
 \end{figure}

### 2017-04-27 paper.tex
complete time results for mmaps

@@
 
-\subsubsection{Time (TODO: HP)}
+\subsubsection{Time}
 
@@
 accommodate memory. This ensures that there is no need to perform the lookup on the red black trees that 
-store the vm area sturct so a constant line is seen.
+store the vm area sturct so a horizontal line is seen. The second important thing to note here is that 
+for continuous mmap operations, the time taken is always the lowest compared to other benchmarks. This 
+is also a result of expanding the existing vm area sturcts. Since the number of vm area structs is very 
+low (actually just 1 for the process in question) the required node in the red black tree is always found in
+the cache. For rest of the microbench marks, the kernel has to first search through the red-black tree of vm area
+structs. If it finds a mapping adjacent to the new memory allocation to be done, it will extend the existing 
+mapping otherwise it will create a new mapping and insert it into the red-black tree. The traversal of the 
+tree is an o(long n) operation where n is the number of nodes in the tree which in this case are the vm area
+mappings. This log n shape for the lookup is visible in the plots for all the benchmarks in figure \ref{fig:mmap_
+latency}. The frag mmap benchmark takes the most time because for fragmented mmap, the memory is first 
+allocated contiguously and then all allocated regions are expanded to a higher size. This operation needs creation
+of a new memory area struct. Therefore, the kernel first looks through the red-black tree but it never finds 
+a node which can be expanded to include the current mapping. So it always has to create a new mapping after 
+traversing the tree and then inserting it in. It also has to delete the old mapping from the red-black tree
+and then rebalnce the tree if necessary. All of these operations result in the worst case latency for mmap. 
+The case of strided mmaps is very similar to frag because it will also force the kerenl to always create a new
+mapping and insert it into the tree after traversing the tree first. It however does not need to delete an 
+existing mapping so it is faster than the frag benchmark. Finally, rand will also search through the red-black 
+tree first but it will sometimes find an existing mapping adjacent to the desired mapping which can be expanded
+obviating the need to create a new vm area mapping and inserting it into red-black tree. Hence, most of the times
+it is faster than stride.
 

### 2017-04-27 paper.tex
Explain memory results for mmap

@@
 \subsection{\texttt{vm\_area\_struct}}
+\label{ss_vm_area_struct}
 
@@
 
-Figure \ref{fig:mmap_latency} shows for various microbenchmarks the time taken per mmap operation 
-as it varies with the total amount memory allocate so far in the system. As described above, we use 
-rdtsc to find the number of ticks taken for the operation and then calibrate rdtsc to conver
-these ticks into microseconds. It can be seen that for continuous mmaps the time is not only the lowest
-but also invariant with the total number of mmaps done so far. This is because of two reasons. First, 
-when a continuous mmap (with the same permissions) happens the kernel memory allocator does not need 
-create a new vm area struct, instead it modifies the adjacent vm area structs to account for the newly 
-accommodate memory. This ensures that there is no need to perform the lookup on the red black trees that 
-store the vm area sturct so a horizontal line is seen. The second important thing to note here is that 
-for continuous mmap operations, the time taken is always the lowest compared to other benchmarks. This 
-is also a result of expanding the existing vm area sturcts. Since the number of vm area structs is very 
-low (actually just 1 for the process in question) the required node in the red black tree is always found in
-the cache. For rest of the microbench marks, the kernel has to first search through the red-black tree of vm area
-structs. If it finds a mapping adjacent to the new memory allocation to be done, it will extend the existing 
-mapping otherwise it will create a new mapping and insert it into the red-black tree. The traversal of the 
-tree is an o(long n) operation where n is the number of nodes in the tree which in this case are the vm area
-mappings. This log n shape for the lookup is visible in the plots for all the benchmarks in figure \ref{fig:mmap_
-latency}. The frag mmap benchmark takes the most time because for fragmented mmap, the memory is first 
-allocated contiguously and then all allocated regions are expanded to a higher size. This operation needs creation
-of a new memory area struct. Therefore, the kernel first looks through the red-black tree but it never finds 
-a node which can be expanded to include the current mapping. So it always has to create a new mapping after 
-traversing the tree and then inserting it in. It also has to delete the old mapping from the red-black tree
-and then rebalnce the tree if necessary. All of these operations result in the worst case latency for mmap. 
-The case of strided mmaps is very similar to frag because it will also force the kerenl to always create a new
-mapping and insert it into the tree after traversing the tree first. It however does not need to delete an 
-existing mapping so it is faster than the frag benchmark. Finally, rand will also search through the red-black 
-tree first but it will sometimes find an existing mapping adjacent to the desired mapping which can be expanded
-obviating the need to create a new vm area mapping and inserting it into red-black tree. Hence, most of the times
-it is faster than stride.
+Figure \ref{fig:mmap_latency} shows for various microbenchmarks the time taken
+per mmap operation as it varies with the total amount memory allocate so far in
+the system. As described above, we use rdtsc to find the number of ticks taken
+for the operation and then calibrate rdtsc to conver these ticks into
+microseconds. It can be seen that for continuous mmaps the time is not only the
+lowest but also invariant with the total number of mmaps done so far. This is
+because of two reasons. First, when a continuous mmap (with the same
+permissions) happens the kernel memory allocator does not need create a new vm
+area struct, instead it modifies the adjacent vm area structs to account for the
+newly accommodate memory. This ensures that there is no need to perform the
+lookup on the red black trees that store the vm area sturct so a horizontal line
+is seen. The second important thing to note here is that for continuous mmap
+operations, the time taken is always the lowest compared to other benchmarks.
+This is also a result of expanding the existing vm area sturcts. Since the
+number of vm area structs is very low (actually just 1 for the process in
+question) the required node in the red black tree is always found in the cache.
+For rest of the microbench marks, the kernel has to first search through the
+red-black tree of vm area structs. If it finds a mapping adjacent to the new
+memory allocation to be done, it will extend the existing mapping otherwise it
+will create a new mapping and insert it into the red-black tree. The traversal
+of the tree is an o(long n) operation where n is the number of nodes in the tree
+which in this case are the vm area mappings. This log n shape for the lookup is
+visible in the plots for all the benchmarks in figure \ref{fig:mmap_ latency}.
+The frag mmap benchmark takes the most time because for fragmented mmap, the
+memory is first allocated contiguously and then all allocated regions are
+expanded to a higher size. This operation needs creation of a new memory area
+struct. Therefore, the kernel first looks through the red-black tree but it
+never finds a node which can be expanded to include the current mapping. So it
+always has to create a new mapping after traversing the tree and then inserting
+it in. It also has to delete the old mapping from the red-black tree and then
+rebalnce the tree if necessary. All of these operations result in the worst case
+latency for mmap.  The case of strided mmaps is very similar to frag because it
+will also force the kerenl to always create a new mapping and insert it into the
+tree after traversing the tree first. It however does not need to delete an
+existing mapping so it is faster than the frag benchmark. Finally, rand will
+also search through the red-black tree first but it will sometimes find an
+existing mapping adjacent to the desired mapping which can be expanded obviating
+the need to create a new vm area mapping and inserting it into red-black tree.
+Hence, most of the times it is faster than stride.
 
@@
 
-\subsubsection{Memory (TODO: SPAI)}
+\subsubsection{Memory}
+Figure \ref{fig:mmap\_mem\_usage} shows the growth in memory usage of the system
+for the workloads described in section \ref{ss\_vm\_area\_struct}. The
+\texttt{cont} workload coincides with the \texttt{control} workload as only the
+first mmap operation allocates a \texttt{vm\_area\_struct}. The subsequent
+operations are subsumed by the initial \texttt{vm\_area\_struct} as the mapped
+regions are adjacent, and hence the memory usage does not increase despite the
+large number of mmap operations.  The \texttt{stride} workload allocates a
+\texttt{vm\_area\_struct} for each operation due to the separation of
+allocations in the virtual address space. As a result, the memory usage
+increases significantly as more memory is consumed with each mmap operation.
+This is also the case for the \texttt{rand} workload which breaks up a single
+\texttt{vm\_area\_struct} for a large area in the virtual address space into
+multiple smaller areas, which results in the  allocation of a new structure for
+every resize operation. The \texttt{rand} workload initially creates a new
+\texttt{vm\_area\_struct} struct for each mmap operation. Eventually, these
+allocations form larger contiguous allocations when two or more regions are
+mapped next to each other, resulting in a single \texttt{vm\_area\_struct} being
+used for that range. As a result, the memory usage begins to flatten out and
+even decreases slightly as multiple \texttt{vm\_area\_struct}s are collapsed to
+single structures. This is made clear by Figure \ref{fig:vm_area_struct_count}
+which shows the growth \texttt{vm\_area\_struct} counts for each workload.
 \begin{figure}
@@
     \caption{$\Delta_i$ adjusted with control for benchmarks}
+    \label{fig:mmap_mem_usage}
 \end{figure}
@@
     \caption{Number of \texttt{vm\_area\_struct}s for benchmarks}
+    \label{fig:vm_area_struct_count}
 \end{figure}

### 2017-04-27 paper.tex
Fix label references

@@
 \subsubsection{Memory}
-Figure \ref{fig:mmap\_mem\_usage} shows the growth in memory usage of the system
-for the workloads described in section \ref{ss\_vm\_area\_struct}. The
+Figure \ref{fig:mmap_mem_usage} shows the growth in memory usage of the system
+for the workloads described in section \ref{ss_vm_area_struct}. The
 \texttt{cont} workload coincides with the \texttt{control} workload as only the

### 2017-04-27 paper.tex
minor fixes
Explain swapping memory results

@@
 \subsection{Swapping Memory Overhead}
+\label{swapping_memory_overhead}
 
@@
 
-\subsubsection{Memory (TODO: SPAI)}
+\subsubsection{Memory}
+Figure \ref{fig:swap_rss} shows the Resident Set Size of the system for the
+experiments described in section \ref{swapping_memory_overhead}. The
+\texttt{cont} workloads write to  adjacent 4KB pages in the address space. A
+huge page of size 2MB is allocated (TODO: Is it 2MB?) when a new page is touched
+initially. Subsequent 4KB pages that are touched are subpages of the 2MB page
+allocated initially. As a result, the memory used by the workload is almost
+equal to the amount of memory it has written to. The RSS thus increases linearly
+until the system begins to swap when the number of free pages falls below a
+threshold. The \texttt{rand} workload on the other hand writes to pages that are
+not adjacent, and likely not subpages of a single huge page. As a result, a huge
+page is allocated with almost every write, and the system starts to exhaust its
+memory and begins swapping very quickly. Memory made available by kswapd is
+quickly consumed by further writes and hence the memory used by the workload
+continues to stay close to its peak value. As expected, the behavior is the same
+irrespective of the size of the swap file.
 
@@
     \caption{RSS while touching pages while swapping for random benchmark.}
+    \label{swap_rss}
 \end{figure}

### 2017-04-27 paper.tex
Moved related work
Updated methodology section

@@
 building a research tool and for building systems.
-
-\section{Related Work}
-
-The emergence of new technology has in the past often required operating systems
-to change. Traditionally, OS researchers and developers benchmark and test their
-solutions in new environments to demonstrate their efficacy before using them in
-real systems. However, researchers do not always have access to new technology
-for practical reasons; for example, large memory systems are still too expensive
-for most research budgets. This is not the first time that researchers have been
-forced to study systems in environments they do not have access to. This section
-examines past approaches to studying these systems and motivates the need for
-our work.
-
-The emergence of the internet required businesses to buy expensive machines to
-keep up with demand for their services, but systems researchers did not have
-access to these machines. Alameldeen et al. describe how they simulate a
-multi-million dollar server using a \$2000 workstation. To do this, they go
-through several rounds of scaling down, optimizing, changing, and tuning both
-the benchmarks and systems they test. For example, they scale down a workload to
-fit in the 1GB memory of the machine they use and increase the number of threads
-to improve parallelism \cite{2kmachine}. This methodology may be accurate, but
-it requires modifying the benchmarks, which is error prone and time consuming;
-it can be extremely difficult to validate changes to a benchmark. Ideally,
-researchers can run large experimental workloads without changes, but the overheads
-of current memory management systems prevent this from being practical.
-
-Likewise, the Quartz emulator studies the behaviour of applications on
-Non-volatile Memory (NVM) while actually running on top of DRAM. It uses
-existing hardware capabilities to emulate the higher latency and lower bandwidth
-of most NVM technologies. While Quartz does not address the fact that the actual
-memory available to the applications on NVM might be orders of magnitude higher
-than on DRAM, it demonstrates that new technologies may be emulated using
-existing technology \cite{quartz}. (TODO: is this system still relavent?)
-
-The Simics simulator uses demand paging for simulated environments, which means
-that memory is only allocated when it is used. As long as the working set of
-the target system fits in the host memory, performance will be tolerable.
-However, in our work, we wish to study OS behavior under workloads that
-actually use all available memory. Thus, the usefulness of demand paging is
-greatly reduced because swapping causes disk latency to become the main
-performance bottleneck. Moreover, the performance overhead of simulators like
-Simics tends to be impractical because system state needs to be maintained by
-the simulator \cite{simics}. We do not explore simulation further because it
-can be orders of magnitude slower than native execution, making it infeasible
-for studying large workloads or systems \cite{2kmachine}. However, the Simics
-simulator illustrates the usefulness of memory management techniques such as
-demand paging and the potential overheads of swapping.
-
-A similar problem was encountered by researchers studying both large storage
-drives and large distributed storage systems. David is a system that allows
-storage and big data researchers to run large benchmarks requiring terabytes of
-storage using off the shelf storage devices (which at that time were too small).
-David creates a compressed version of the file system by physically storing only
-metadata and discarding the contents of files. Reads to the disk causes David to
-generate data on the fly. This decision choice is based on the observation that
-most benchmarking frameworks do not care about the actual content of the files,
-and that most of the storage capacity of a drive tends to be data rather than
-metadata \cite{david}. The Exalt system uses a similar methodology for
-large-scale distributed storage systems \cite{exalt}. This methodology provides
-a promising direction for large-memory system studies. One may consider ignoring
-the contents of a process’s heap and only storing kernel data structures and a
-process’s code and stack segments. However, generating heap contents on the fly
-is more difficult than generating disk contents because of the common use of
-custom data structures. Moreover, on Linux, the kernel data structures for 1TB
-of memory may also exceed the size of physical memory on current systems
-\cite{simics}.
-
-A virtualized environment can be used to provide a guest system with more memory
-than is physically available to the host. A study of the limits of the KVM
-hypervisor found that there is no fundamental limit to the size of guest
-physical memory other than the hardware address width. However, currently, a
-Linux host running KVM will require the guest memory to be backed by host
-physical memory or host swap space \cite{ibmkvm}.  This means that when the
-virtual machine uses the whole amount of memory allocated to it, the host can
-swap pages to disk. The resulting poor performance can cause inaccurate
-performance measurements when running benchmarks. It can also make large
-benchmarks impractical.
-
-Interestingly, many of non-volatile memory technologies are making their ways
-into storage drives, such as Intel's recently-announced Optane SSDs. One
-potential use of fast storage devices is for fast swapping space. Some
-literature has proposed that small amounts of DRAM backed by large amounts of
-fast swap space can be a scalable alternative to having large amounts of fast
-RAM (TODO: cite). (TODO: can we have this in our report?). In such a system,
-the overhead of swapping may become a bottleneck.
-
-Prefetching pages from swap space can offer a way to mitigate the overhead of
-memory overcommitment, possibly making large virtual machines a viable
-mechanism for studying large memory systems. When there is significant memory
-pressure, even pages which are likely to be accessed soon are swapped out to
-disk. They are faulted back into memory when accessed, resulting in significant
-performance cost.  Charm++ uses a programming model where computations can be
-scheduled by the language runtime. A designated thread can prefetch pages
-required by a computation, averting page faults \cite{charmpp}. Rather than
-implement page prefetching in a language runtime, one might implement a
-prefetcher in the swapping subsystem, making it general purpose. A large body
-of work already exists on hardware prefetching for processor caches, on which a
-swap prefetcher might draw for inspiration (TODO: cite). However, this approach
-can only work if page access patterns are predictable. Moreover, the high
-latency of disks implies that prefetchers would have to predict the very
-distant future (e.g. seconds ahead of execution). (TODO: Mike's comment on the
-previous version of this paragraph said that it was ``too distant/off topic''.
-Is it better now? I feel like prefetching was an idea that we considered too
-much to leave out...)
-
-Gupta et al. built an emulator for high speed networks using a technique called
-time dilation, which slows down the OS clock to make it appear that external
-events are occurring faster. This allows the system to emulate network links
-with speeds that are currently not available. The implementation is based on
-the VMs and Xen hypervisor; Xen delivers the timer interrupts to the guest at a
-lower rate than hardware hence  slowing down the guest’s clock \cite{timedil}.
-A similar approach may be applied for large memory system studies to slow down
-guest time while paging in and out large portions of memory. This will allows
-the system to believe that it is reading data from the memory while actually
-most of that data is being read from the disk.
-
-Finally, there have been studies which look at the performance overheads
-associated with current implementations of virtual memory and suggest mechanisms
-to mitigate them. RadixVM tries to overcome performance issues in highly
-concurrent workloads due to serialization of memory management operations on
-kernel data structures \cite{radixvm}. This work demonstrates that many parts of
-existing memory management schemes are not scalable to larger systems.
-Similarly, in our work, we wish to examine scalability limitations of memory
-management in the Linux kernel. Other studies have suggested that \texttt{struct
-page}, \texttt{struct vm\_area\_struct}, and page tables tend to comprise a
-large portion of memory management overhead \cite{simics} (TODO: cite the LWN
-paper).
 
@@
 
+(TODO: swapping large pages?)
+
 \subsection{Page Frame Reclamation}
@@
 of the Linux memory management system.  First, we measure the scalability of
-Linux's \texttt{vm\_area\_struct}. Then, we measure the minimum amount of memory
-necessary for swapping.
+Linux's \texttt{vm\_area\_struct}. Then, we measure the scalability of Linux's
+swapping mechanisms.
 
@@
 Kernel & Linux 4.4.0-70 \\ \hline
-CPU & Intel Xeon E5645, 2.40 GHz, 6 cores/12 threads, 384KB Private L1-I Cache,
-1.536 Unified L2 Cache, 12MB Shared L3 Cache \\ \hline
+CPU & Intel Xeon E5645, 2.40 GHz, 6 cores/12 threads, 32kB L1-D, 32kB L1-I,
+    256kB L2, 12MB L3 \\ \hline
 Memory & 16GB, 1066MHz \\ \hline
-Storage & 500GB, 7200RPM \\
+Storage & 500GB HDD, 7200RPM \\
 \hline
@@
 Figure \ref{fig:specs}. We disabled Intel SpeedStep and TurboBoost and ensure
-that the processor frequency stays constant (TODO: did we do this?). All
-experiments are pinned to the same core on our machine.
+that the processor frequency stays constant. All experiments are pinned to the
+same core on our machine.
 
@@
 session. We are careful to avoid opening extra ssh sessions, running screen or
-tmux, or executing extra commands while a benchmark is running, as these may
-impact memory usage, and thus our results.
+tmux, or executing extra commands while a benchmark is running since these may
+impact memory usage and, thus, our results.
 
@@
 timestamp before and after each memory management operation in the workload and
-take the difference to get latency. (TODO: is this still true?)
+take the difference to get latency.
 
@@
 system as the total amount of memory less the amount of memory used by the
-benchmark.
+benchmark. We keep transparent huge pages on throughout our experiments, unless
+otherwise specified.
 
-Finally, we keep transparent huge pages on throughout our experiments, unless
-otherwise specified.
+To measure processor usage, we use the procfs's reporting on number of jiffies.
+\texttt{/proc/stat} reports the number of jiffies of uptime the system has seen
+overall, while \texttt{/proc/[pid]/stat} reports on the specified process.
 
@@
 Assume the memory usage of the rest of the system is $R$. We denote $R_i$ as the
-$i$-th measurement of $R$ during a run ($i = 0, 1, ... 2^{20}$). System memory
+$i$-th measurement of $R$ during a run ($i = 0, 1, ..., 2^{20}$). System memory
 usage is known to jitter in Linux (TODO: cite). To adjust for this jitter, we
@@
 \item \textit{Fragmented} (``frag'' for short). This workload first allocates
-    $4^{20}$ contiguous 4KB blocks. Then, it resizes every other block to 16KB.
+    $2^{21}$ contiguous 4KB blocks. Then, it resizes every other block to 16KB.
         The resizing operation, rather than the initial allocation is measured.
@@
 
-We run this experiment with both a strided access pattern and a random access
-pattern. The stride access pattern simply touches pages in order of increasing
-address. The random access pattern chooses a random address. Both patterns touch
-the same number of addresses.
+We run this experiment with both a continuous access pattern and a random access
+pattern. The continuous access pattern simply touches adjacent pages in order of
+increasing address. The random access pattern chooses a random address. Both
+patterns touch the same number of addresses.
 
@@
 benchmark completes. This prevents the writing of the results from interacting
-with paging. These RSS values are graphed in our results section.
+with paging in the kernel page cache. These RSS values are graphed in our
+results section.
 
@@
 
-A single process allocates 18GB of virtual memory space. It then proceeds to
-write one byte of each 4KB page in the region, leading to the actual allocation
-of physical memory. The process chooses addresses randomly, as in the previous
-set of experiments, inducing the maximum amount of swapping.
+A single process allocates 18GB of virtual memory space and writes one byte of
+each 4KB page in the region. The same two access patterns are tested as in the
+previous set of experiments (continuous and random).
 
 In all these experiments, the system is given 32GB of swap space total. We
-measure the impact of physical memory size and transparent huge pages. We run
-four experiments:
+repeat each experiment with hugepages enabled and disabled to see the impact of
+the transparent hugepage system. The processor usage of kswapd and khugepaged is
+recorded for each experiment and reported in the next section.
 
-\begin{enumerate}
-    \item 1GB physical memory with huge pages enabled
-    \item 16GB physical memory with huge pages enabled
-    \item 1GB physical memory with huge pages disabled
-    \item 16GB physical memory with huge pages disabled
-\end{enumerate}
-
-To run the 1GB memory benchmarks, we hot unplug 15GB using the sysfs in Linux.
-This makes the unplugged memory unusable by the whole system (kernel and user).
-
-We measure CPU time via the procfs using the same methodology as the
-\texttt{top} utility. The \texttt{/proc/[pid]/stat} file for kswapd is read.
-From this file the \texttt{utime} and \texttt{stime} fields are read and summed
-to get the total time allocated to kswapd. Then, the \texttt{/proc/stat} file is
-read to get the total system time elapsed. By dividing, we can find the portion
-of CPU time spent in kswapd.
-
-\section{Result}
+\section{Results}
 
@@
     \includegraphics[width=\columnwidth]{figures/swap_time_rand}
-    \caption{Time to touch a page while swapping for random benchmark.\label{fig:swap_time_rand}}
+    \caption{Time to touch a page while swapping for random
+    benchmark.\label{fig:swap_time_rand}}
 \end{figure}
@@
 
+TODO: khugepaged?
+
+\section{Related Work}
+
+The emergence of new technology has in the past often required operating systems
+to change. Traditionally, OS researchers and developers benchmark and test their
+solutions in new environments to demonstrate their efficacy before using them in
+real systems. However, researchers do not always have access to new technology
+for practical reasons; for example, large memory systems are still too expensive
+for most research budgets. This is not the first time that researchers have been
+forced to study systems in environments they do not have access to. This section
+examines past approaches to studying these systems and motivates the need for
+our work.
+
+The emergence of the internet required businesses to buy expensive machines to
+keep up with demand for their services, but systems researchers did not have
+access to these machines. Alameldeen et al. describe how they simulate a
+multi-million dollar server using a \$2000 workstation. To do this, they go
+through several rounds of scaling down, optimizing, changing, and tuning both
+the benchmarks and systems they test. For example, they scale down a workload to
+fit in the 1GB memory of the machine they use and increase the number of threads
+to improve parallelism \cite{2kmachine}. This methodology may be accurate, but
+it requires modifying the benchmarks, which is error prone and time consuming;
+it can be extremely difficult to validate changes to a benchmark. Ideally,
+researchers can run large experimental workloads without changes, but the overheads
+of current memory management systems prevent this from being practical.
+
+Likewise, the Quartz emulator studies the behaviour of applications on
+Non-volatile Memory (NVM) while actually running on top of DRAM. It uses
+existing hardware capabilities to emulate the higher latency and lower bandwidth
+of most NVM technologies. While Quartz does not address the fact that the actual
+memory available to the applications on NVM might be orders of magnitude higher
+than on DRAM, it demonstrates that new technologies may be emulated using
+existing technology \cite{quartz}. (TODO: is this system still relavent?)
+
+The Simics simulator uses demand paging for simulated environments, which means
+that memory is only allocated when it is used. As long as the working set of
+the target system fits in the host memory, performance will be tolerable.
+However, in our work, we wish to study OS behavior under workloads that
+actually use all available memory. Thus, the usefulness of demand paging is
+greatly reduced because swapping causes disk latency to become the main
+performance bottleneck. Moreover, the performance overhead of simulators like
+Simics tends to be impractical because system state needs to be maintained by
+the simulator \cite{simics}. We do not explore simulation further because it
+can be orders of magnitude slower than native execution, making it infeasible
+for studying large workloads or systems \cite{2kmachine}. However, the Simics
+simulator illustrates the usefulness of memory management techniques such as
+demand paging and the potential overheads of swapping.
+
+A similar problem was encountered by researchers studying both large storage
+drives and large distributed storage systems. David is a system that allows
+storage and big data researchers to run large benchmarks requiring terabytes of
+storage using off the shelf storage devices (which at that time were too small).
+David creates a compressed version of the file system by physically storing only
+metadata and discarding the contents of files. Reads to the disk causes David to
+generate data on the fly. This decision choice is based on the observation that
+most benchmarking frameworks do not care about the actual content of the files,
+and that most of the storage capacity of a drive tends to be data rather than
+metadata \cite{david}. The Exalt system uses a similar methodology for
+large-scale distributed storage systems \cite{exalt}. This methodology provides
+a promising direction for large-memory system studies. One may consider ignoring
+the contents of a process’s heap and only storing kernel data structures and a
+process’s code and stack segments. However, generating heap contents on the fly
+is more difficult than generating disk contents because of the common use of
+custom data structures. Moreover, on Linux, the kernel data structures for 1TB
+of memory may also exceed the size of physical memory on current systems
+\cite{simics}.
+
+A virtualized environment can be used to provide a guest system with more memory
+than is physically available to the host. A study of the limits of the KVM
+hypervisor found that there is no fundamental limit to the size of guest
+physical memory other than the hardware address width. However, currently, a
+Linux host running KVM will require the guest memory to be backed by host
+physical memory or host swap space \cite{ibmkvm}.  This means that when the
+virtual machine uses the whole amount of memory allocated to it, the host can
+swap pages to disk. The resulting poor performance can cause inaccurate
+performance measurements when running benchmarks. It can also make large
+benchmarks impractical.
+
+Interestingly, many of non-volatile memory technologies are making their ways
+into storage drives, such as Intel's recently-announced Optane SSDs. One
+potential use of fast storage devices is for fast swapping space. Some
+literature has proposed that small amounts of DRAM backed by large amounts of
+fast swap space can be a scalable alternative to having large amounts of fast
+RAM (TODO: cite). (TODO: can we have this in our report?). In such a system,
+the overhead of swapping may become a bottleneck.
+
+Prefetching pages from swap space can offer a way to mitigate the overhead of
+memory overcommitment, possibly making large virtual machines a viable
+mechanism for studying large memory systems. When there is significant memory
+pressure, even pages which are likely to be accessed soon are swapped out to
+disk. They are faulted back into memory when accessed, resulting in significant
+performance cost.  Charm++ uses a programming model where computations can be
+scheduled by the language runtime. A designated thread can prefetch pages
+required by a computation, averting page faults \cite{charmpp}. Rather than
+implement page prefetching in a language runtime, one might implement a
+prefetcher in the swapping subsystem, making it general purpose. A large body
+of work already exists on hardware prefetching for processor caches, on which a
+swap prefetcher might draw for inspiration (TODO: cite). However, this approach
+can only work if page access patterns are predictable. Moreover, the high
+latency of disks implies that prefetchers would have to predict the very
+distant future (e.g. seconds ahead of execution). (TODO: Mike's comment on the
+previous version of this paragraph said that it was ``too distant/off topic''.
+Is it better now? I feel like prefetching was an idea that we considered too
+much to leave out...)
+
+Gupta et al. built an emulator for high speed networks using a technique called
+time dilation, which slows down the OS clock to make it appear that external
+events are occurring faster. This allows the system to emulate network links
+with speeds that are currently not available. The implementation is based on
+the VMs and Xen hypervisor; Xen delivers the timer interrupts to the guest at a
+lower rate than hardware hence  slowing down the guest’s clock \cite{timedil}.
+A similar approach may be applied for large memory system studies to slow down
+guest time while paging in and out large portions of memory. This will allows
+the system to believe that it is reading data from the memory while actually
+most of that data is being read from the disk.
+
+Finally, there have been studies which look at the performance overheads
+associated with current implementations of virtual memory and suggest mechanisms
+to mitigate them. RadixVM tries to overcome performance issues in highly
+concurrent workloads due to serialization of memory management operations on
+kernel data structures \cite{radixvm}. This work demonstrates that many parts of
+existing memory management schemes are not scalable to larger systems.
+Similarly, in our work, we wish to examine scalability limitations of memory
+management in the Linux kernel. Other studies have suggested that \texttt{struct
+page}, \texttt{struct vm\_area\_struct}, and page tables tend to comprise a
+large portion of memory management overhead \cite{simics} (TODO: cite the LWN
+paper).
+
 \bibliography{references}{}

### 2017-04-27 paper.tex
Edits in the results section

@@
 Figure \ref{fig:mmap_latency} shows for various microbenchmarks the time taken
-per mmap operation as it varies with the total amount memory allocate so far in
-the system. As described above, we use rdtsc to find the number of ticks taken
-for the operation and then calibrate rdtsc to conver these ticks into
-microseconds. It can be seen that for continuous mmaps the time is not only the
-lowest but also invariant with the total number of mmaps done so far. This is
-because of two reasons. First, when a continuous mmap (with the same
-permissions) happens the kernel memory allocator does not need create a new vm
-area struct, instead it modifies the adjacent vm area structs to account for the
-newly accommodate memory. This ensures that there is no need to perform the
-lookup on the red black trees that store the vm area sturct so a horizontal line
-is seen. The second important thing to note here is that for continuous mmap
-operations, the time taken is always the lowest compared to other benchmarks.
-This is also a result of expanding the existing vm area sturcts. Since the
-number of vm area structs is very low (actually just 1 for the process in
-question) the required node in the red black tree is always found in the cache.
-For rest of the microbench marks, the kernel has to first search through the
-red-black tree of vm area structs. If it finds a mapping adjacent to the new
-memory allocation to be done, it will extend the existing mapping otherwise it
-will create a new mapping and insert it into the red-black tree. The traversal
-of the tree is an o(long n) operation where n is the number of nodes in the tree
-which in this case are the vm area mappings. This log n shape for the lookup is
-visible in the plots for all the benchmarks in figure \ref{fig:mmap_ latency}.
-The frag mmap benchmark takes the most time because for fragmented mmap, the
-memory is first allocated contiguously and then all allocated regions are
-expanded to a higher size. This operation needs creation of a new memory area
-struct. Therefore, the kernel first looks through the red-black tree but it
-never finds a node which can be expanded to include the current mapping. So it
-always has to create a new mapping after traversing the tree and then inserting
-it in. It also has to delete the old mapping from the red-black tree and then
-rebalnce the tree if necessary. All of these operations result in the worst case
-latency for mmap.  The case of strided mmaps is very similar to frag because it
-will also force the kerenl to always create a new mapping and insert it into the
-tree after traversing the tree first. It however does not need to delete an
-existing mapping so it is faster than the frag benchmark. Finally, rand will
-also search through the red-black tree first but it will sometimes find an
-existing mapping adjacent to the desired mapping which can be expanded obviating
-the need to create a new vm area mapping and inserting it into red-black tree.
-Hence, most of the times it is faster than stride.
+per \texttt{mmap} operation as it varies with the total amount memory allocated
+so far in the workload.
+
+It can be seen that for continuous \texttt{mmap}s the time is not only the
+lowest but is also invariant with the total number of \texttt{mmap}s done so
+far. There are two reasons for this.
+
+First, when a continuous \texttt{mmap} (with the same permissions) happens the
+kernel memory allocator does not need to create a new \texttt{vm\_area\_struct},
+instead it modifies the adjacent \texttt{vm\_area\_struct}s to account for the
+newly allocated space. This avoids the need to use the slab allocator to create
+a new struct.
+
+Second, for continuous \texttt{mmap} operations, the time taken is always the
+lowest compared to other benchmarks.  This is because the kernel caches the most
+recently used \texttt{vm\_area\_struct}, avoiding a red-black tree search. In
+the continuous benchmark, this cached struct is always conveniently used.
+
+For the other workloads, a search through the red-black tree is often needed on
+most accesses. This operation takes $O(log n)$ time (where $n$ is the number of
+virtual memory areas). The logarithmic behavior is visible in the plots for the
+benchmarks in Figure \ref{fig:mmap_latency}.
+
+The frag benchmark takes the most time of any benchmark. When a region of the
+address space is being expanded, the kernel is forced to unmap it and map a new
+area somewhere else.  This operation needs creation of a new memory area struct.
+The kernel also must look through the red-black tree, though it never finds a
+node which can be expanded to include the new mapping. Then, an insertion on the
+tree is needed, possibly including rebalancing.
+
+The strided workload is very similar to frag because it will also force the
+kerenl to always create a new mapping and insert it into the tree after
+traversing the tree first. It, however, does not need to delete an existing
+mapping so it is faster than the frag benchmark.
+
+Finally, rand will also search through the red-black tree first but there is a
+reasonably high probability that it will find an existing mapping adjacent to
+the desired mapping which can be expanded, obviating the need to create a new
+struct and insert it into red-black tree.
+
+~\\ \textbf{Conclusions}
+
+TODO?
 
@@
 \texttt{cont} workload coincides with the \texttt{control} workload as only the
-first mmap operation allocates a \texttt{vm\_area\_struct}. The subsequent
+first \texttt{mmap} operation allocates a \texttt{vm\_area\_struct}. The subsequent
 operations are subsumed by the initial \texttt{vm\_area\_struct} as the mapped
 regions are adjacent, and hence the memory usage does not increase despite the
-large number of mmap operations.  The \texttt{stride} workload allocates a
+large number of \texttt{mmap} operations.  The \texttt{stride} workload allocates a
 \texttt{vm\_area\_struct} for each operation due to the separation of
 allocations in the virtual address space. As a result, the memory usage
-increases significantly as more memory is consumed with each mmap operation.
+increases significantly as more memory is consumed with each \texttt{mmap} operation.
 This is also the case for the \texttt{rand} workload which breaks up a single
@@
 every resize operation. The \texttt{rand} workload initially creates a new
-\texttt{vm\_area\_struct} struct for each mmap operation. Eventually, these
+\texttt{vm\_area\_struct} struct for each \texttt{mmap} operation. Eventually, these
 allocations form larger contiguous allocations when two or more regions are

### 2017-04-27 paper.tex
Swapping huge pages
Other edits in results section

@@
 
+(TODO: we need to be consistent with whether we texttt the benchmark names or
+not)
+
 \subsection{\texttt{vm\_area\_struct} Experiments}
@@
 Figure \ref{fig:mmap_mem_usage} shows the growth in memory usage of the system
-for the workloads described in section \ref{ss_vm_area_struct}. The
-\texttt{cont} workload coincides with the \texttt{control} workload as only the
+for the workloads described in section \ref{ss_vm_area_struct}.
+
+The
+\texttt{cont} workload coincides with the \texttt{control} workload, as only the
 first \texttt{mmap} operation allocates a \texttt{vm\_area\_struct}. The subsequent
-operations are subsumed by the initial \texttt{vm\_area\_struct} as the mapped
+operations are subsumed by the initial \texttt{vm\_area\_struct} because the mapped
 regions are adjacent, and hence the memory usage does not increase despite the
-large number of \texttt{mmap} operations.  The \texttt{stride} workload allocates a
+large number of \texttt{mmap} operations.
+
+The \texttt{stride} workload allocates a
 \texttt{vm\_area\_struct} for each operation due to the separation of
@@
 increases significantly as more memory is consumed with each \texttt{mmap} operation.
-This is also the case for the \texttt{rand} workload which breaks up a single
+
+This is also the case for the \texttt{frag} workload which breaks up a single
 \texttt{vm\_area\_struct} for a large area in the virtual address space into
 multiple smaller areas, which results in the allocation of a new structure for
-every resize operation. The \texttt{rand} workload initially creates a new
+every resize operation.
+
+The \texttt{rand} workload initially creates a new
 \texttt{vm\_area\_struct} struct for each \texttt{mmap} operation. Eventually, these
@@
 mapped next to each other, resulting in a single \texttt{vm\_area\_struct} being
-used for that range. As a result, the memory usage begins to flatten out and
+used for that range. As a result, the memory usage begins to taper off and
 even decreases slightly as multiple \texttt{vm\_area\_struct}s are collapsed to
@@
 which shows the growth \texttt{vm\_area\_struct} counts for each workload.
+
+~\\ \textbf{Conclusions}
+
+TODO
+
 \begin{figure}
@@
 \subsubsection{Memory}
-Figure \ref{fig:swap_rss} shows the Resident Set Size of the system for the
-experiments described in section \ref{swapping_memory_overhead}. The
-\texttt{cont} workloads write to  adjacent 4KB pages in the address space. A
-huge page of size 2MB is allocated (TODO: Is it 2MB?) when a new page is touched
+
+Figure \ref{fig:swap_rss} shows the RSS of each workload for the
+experiments described in section \ref{swapping_memory_overhead}.
+
+The
+cont workloads write to  adjacent 4KB pages in the address space. A
+huge page of size 2MB is allocated when a new page is touched
 initially. Subsequent 4KB pages that are touched are subpages of the 2MB page
@@
 until the system begins to swap when the number of free pages falls below a
-threshold. The \texttt{rand} workload on the other hand writes to pages that are
-not adjacent, and likely not subpages of a single huge page. As a result, a huge
-page is allocated with almost every write, and the system starts to exhaust its
-memory and begins swapping very quickly. Memory made available by kswapd is
-quickly consumed by further writes and hence the memory used by the workload
+threshold.
+
+The rand workload, on the other hand, writes to pages that are likely
+not adjacent and likely not subpages of a single huge page. As a result, a huge
+page is allocated with almost every write, and system memory is exhausted and
+begins swapping very quickly. Memory made available by kswapd is
+quickly consumed by further allocations. Hence, the memory used by the workload
 continues to stay close to its peak value. As expected, the behavior is the same
-irrespective of the size of the swap file.
+irrespective of the size of the swapfile.
+
+~\\ \textbf{Conclusions}
+
+TODO
 

### 2017-04-27 paper.tex
Minor edit to swapping hugepages
some changes to time section for mmap

@@
 
-(TODO: swapping large pages?)
+On systems with transparent hugepages enabled, huge pages are not swapped out in
+one piece.  They are broken down to their component small pages, which are then
+swapped to disk via the normal mechanisms. This is done to simplify the swapping
+logic in the kernel, but results in additional overheads
+\cite{corbet_transparent}.
 
@@
 
-Figure \ref{fig:mmap_latency} shows for various microbenchmarks the time taken
+Figure \ref{fig:mmap_latency} shows for various microbenchmarks, the time taken
 per \texttt{mmap} operation as it varies with the total amount memory allocated
-so far in the workload.
+so far by the system.
 
@@
 instead it modifies the adjacent \texttt{vm\_area\_struct}s to account for the
-newly allocated space. This avoids the need to use the slab allocator to create
+newly allocated space. This obviates the need to use the slab allocator to create
 a new struct.
@@
 node which can be expanded to include the new mapping. Then, an insertion on the
-tree is needed, possibly including rebalancing.
+tree is needed, possibly including rebalancing. Moreover, the exisiting texttt{
+vm\_area\_struct} might need to be deleted and removed from the tree. All of these
+operations ensure that frag mmap pattern takes the most amount of time.
 
@@
 the desired mapping which can be expanded, obviating the need to create a new
-struct and insert it into red-black tree.
+struct and insert it into red-black tree. Therefore, most of the times the time 
+taken by rand benchmark is lower than strided.
 

### 2017-04-27 paper.tex
Add some conclusions for results

@@
 
-TODO
+The memory overhead of \texttt{vm\_area\_struct}s increases linearly with the
+number of mapped regions when these regions are separated by unused memory
+addresses. When a workload maps a large number of regions, this overhead can be
+of the order of hundreds of megabytes to gigabytes for large memory systems.  To
+reduce this overhead, it is crucial that regions with similar access permissions
+are placed adjacently in the address space so that they can be identified by a
+single structure. It is important for large applications such as object-oriented
+databases that map a large number of regions (TODO: Cite Linux book. Should we
+mention this?)to be aware of these overheads and mitigate them.
 
@@
 
-The
-cont workloads write to  adjacent 4KB pages in the address space. A
-huge page of size 2MB is allocated when a new page is touched
-initially. Subsequent 4KB pages that are touched are subpages of the 2MB page
-allocated initially. As a result, the memory used by the workload is almost
-equal to the amount of memory it has written to. The RSS thus increases linearly
-until the system begins to swap when the number of free pages falls below a
-threshold.
+The cont workloads write to adjacent 4KB pages in the address space. A huge page
+of size 2MB is allocated when a new page is touched initially. Subsequent 4KB
+pages that are touched are subpages of the 2MB page allocated initially. As a
+result, the memory used by the workload is almost equal to the amount of memory
+it has written to. The RSS thus increases linearly until the system begins to
+swap when the number of free pages falls below a threshold.
 
-The rand workload, on the other hand, writes to pages that are likely
-not adjacent and likely not subpages of a single huge page. As a result, a huge
-page is allocated with almost every write, and system memory is exhausted and
-begins swapping very quickly. Memory made available by kswapd is
-quickly consumed by further allocations. Hence, the memory used by the workload
-continues to stay close to its peak value. As expected, the behavior is the same
-irrespective of the size of the swapfile.
+The rand workload, on the other hand, writes to pages that are likely not
+adjacent and likely not subpages of a single huge page. As a result, a huge page
+is allocated with almost every write, and system memory is exhausted and begins
+swapping very quickly. Memory made available by kswapd is quickly consumed by
+further allocations. Hence, the memory used by the workload continues to stay
+close to its peak value. As expected, the behavior is the same irrespective of
+the size of the swapfile.
 
@@
 
-TODO
+Transparent huge pages can consume a large fraction of memory and lead to early
+onset of memory pressure when the workload accesses many pages that are
+separated in its address space. The overhead of swapping greatly exceeds the
+performance benefits provided by huge pages. As a result, the workload
+performance can fall drastically in such cases. It might be advisable to
+disable transparent huge pages for such applications. The kernel can also be
+cautious and limit the allocation of huge pages to processes in order to
+prevent early onset of memory pressure.
 

### 2017-04-27 paper.tex
Some changes to paper

@@
 
-\title{Journey: A Study in Unhelpful Titles (TODO: should we get a new title?)}
+\title{Journey: A Study of Scalabilty of Memory Management in Linux (TODO: should we get a new title?)}
 \author{Mark Mansi, Suhas Pai, Hasnain Ali Pirzada\\\texttt{\{markm, hp, spai\}@cs.wisc.edu}}
@@
 The Linux kernel keeps track of which portions of a process's address space
-have been allocated using the \texttt{vm\_area\_struct}. Because processes may
-map a number of regions, including shared objects, stack, heap, text, and
-anonymous regions, the scalability of this structure is critical to Linux's
-performance.
+have been allocated using the \texttt{vm\_area\_struct} data structure. Because
+processes may map a number of regions, including shared objects, stack, heap,
+text, and anonymous regions, the scalability of this structure is critical to
+Linux's performance.
 
@@
 memory is allocated. When a process first touches a page, a page fault occurs
-and the processor traps into a page fault handler, which can allocate memory
-and set up a page table entry. This is called demand paging.
+and the processor traps to a page fault handler, which allocates memory and
+sets up a page table entry for a valid access. This is called demand paging.
 
@@
 needs to be efficient to find. Linux caches the most recently used region and
-organizes all regions into a balanced binary tree for fast lookup.  Moreover,
+organizes all regions into a red-black tree for fast lookup.  Moreover,
 adjacent regions with similar permissions and properties are merged in an
@@
 
-(TODO: explain struct page?)
-
 We measure the memory overhead and latency of the Linux Memory Management
 system's use of \texttt{vm\_area\_struct}, which represents a region of a
-process's virtual address space.  
-We do not measure the overhead of \texttt{struct page} because it is known that
-there is one struct for each physical page. We do not measure the overhead of
-page tables because this can be calculated from the size of the virtual memory
-space used by the process.  When a program makes a memory management system
-calls such as \texttt{mmap}, \texttt{mremap}, \texttt{munmap},
-\texttt{mprotect}, \texttt{mlock}, or \texttt{madvise}, the kernel is creating,
-changing, or removing \texttt{vm\_area\_struct}s; thus, measuring the
-scalability of Linux memory management requires measuring the overhead of these
-structs.
+process's virtual address space. All the physical pages in the system are
+managed using the \texttt{struct page} descriptors in the kernel.  We do not
+measure the overhead of \texttt{struct page} because it is known that there is
+one struct for each physical page. We do not measure the overhead of page
+tables because this can be calculated from the size of the virtual memory space
+used by the process.  When a program makes a memory management system calls
+such as \texttt{mmap}, \texttt{mremap}, \texttt{munmap}, \texttt{mprotect},
+\texttt{mlock}, or \texttt{madvise}, the kernel is creating, changing, or
+removing \texttt{vm\_area\_struct}s; thus, measuring the scalability of Linux
+memory management requires measuring the overhead of these structs.
 
@@
 First, when a continuous \texttt{mmap} (with the same permissions) happens the
-kernel memory allocator does not need to create a new \texttt{vm\_area\_struct},
-instead it modifies the adjacent \texttt{vm\_area\_struct}s to account for the
+kernel memory allocator does not need to create a new \texttt{vm\_area\_struct}.
+Instead, it uses the existing \texttt{vm\_area\_struct} to account for the
 newly allocated space. This obviates the need to use the slab allocator to create
@@
 Second, for continuous \texttt{mmap} operations, the time taken is always the
-lowest compared to other benchmarks.  This is because the kernel caches the most
-recently used \texttt{vm\_area\_struct}, avoiding a red-black tree search. In
-the continuous benchmark, this cached struct is always conveniently used.
+lowest compared to other benchmarks.  This is because the kernel caches the
+most recently used \texttt{vm\_area\_struct}, avoiding a search in the
+red-black tree. In the continuous benchmark, this cached struct is always
+conveniently used.
 
@@
 address space is being expanded, the kernel is forced to unmap it and map a new
-area somewhere else.  This operation needs creation of a new memory area struct.
-The kernel also must look through the red-black tree, though it never finds a
-node which can be expanded to include the new mapping. Then, an insertion on the
-tree is needed, possibly including rebalancing. Moreover, the exisiting texttt{
-vm\_area\_struct} might need to be deleted and removed from the tree. All of these
-operations ensure that frag mmap pattern takes the most amount of time.
+area elsewhere in the address space. This operation needs creation of a new
+\texttt{vm\_area\_struct}.  The kernel also must search through the red-black
+tree, though it never finds a node which can be expanded to include the new
+mapping. Then, an insertion on the tree is needed, possibly including
+rebalancing. Moreover, the existing \texttt{ vm\_area\_struct} might need to be
+deleted and removed from the tree. All of these operations ensure that frag
+mmap pattern takes the most amount of time.
 
 The strided workload is very similar to frag because it will also force the
-kerenl to always create a new mapping and insert it into the tree after
-traversing the tree first. It, however, does not need to delete an existing
-mapping so it is faster than the frag benchmark.
+kernel to always create a new structure and insert it into the tree after
+traversing the tree first. However, it does not need to delete an existing
+structure, because of which, it is faster than the frag benchmark.
 
-Finally, rand will also search through the red-black tree first but there is a
-reasonably high probability that it will find an existing mapping adjacent to
-the desired mapping which can be expanded, obviating the need to create a new
-struct and insert it into red-black tree. Therefore, most of the times the time 
-taken by rand benchmark is lower than strided.
+Finally, the rand workload will also search through the red-black tree first,
+but there is a reasonably high probability that it will find an existing
+mapping adjacent to the desired mapping which can be expanded, obviating the
+need to create a new struct and insert it into red-black tree. Therefore, most
+of the times the time taken by rand benchmark is lower than strided.
 
@@
 benchmark begins. Many memory accesses thus take milliseconds to complete
-because they require paging in or out, which in turn depends on disk latency.
+because they require paging in or out, which is slow due to disk access.
 Even so, the majority of benchmark run time comes from a minority of accesses.
@@
 accesses complete in less than 20ms. Moreover, notice that the size of the
-swapfile makes very little difference in this workload because the disk has to
+swap file makes very little difference in this workload because the disk has to
 do random seeks regardless of swapfile size.
@@
 
-We also find that in the continuous benchmark, a larger swapfile actually has
-slight more of these slow memory accesses. We believe this has to do with the
+We also find that in the continuous benchmark, a larger swap file actually has
+slightly more of these slow memory accesses. We believe this has to do with the
 fact that the large swapfile has less disk locality, and so causes more disk
 seeks when paging out. Figure \ref{fig:swap_time_cdf} shows that with a larger
-swapfile, the continuous benchmark has a significantly higher number of
-high-latency accesses as compared with the smaller swapfile.
+swap file, the continuous benchmark has a significantly higher number of
+high-latency accesses as compared with the smaller swap file.
 
@@
 
-Large swapfiles should be created with disk locality or on SSDs.
+Large swap files should be created with disk locality or on SSDs.
 
 Looking at 99.99-th percentile latencies is important because they can have a
-lot of impact on overall run time.
+large impact on overall run time.
 

### 2017-04-28 paper.tex
Minor edits

@@
 tables because this can be calculated from the size of the virtual memory space
-used by the process.  When a program makes a memory management system calls
+used by the process.  When a program makes a memory management system call
 such as \texttt{mmap}, \texttt{mremap}, \texttt{munmap}, \texttt{mprotect},
@@
 
-Figure \ref{fig:mmap_latency} shows for various microbenchmarks, the time taken
+Figure \ref{fig:mmap_latency} shows for various microbenchmarks the time taken
 per \texttt{mmap} operation as it varies with the total amount memory allocated
-so far by the system.
+so far by the workload.
 
@@
 mapping. Then, an insertion on the tree is needed, possibly including
-rebalancing. Moreover, the existing \texttt{ vm\_area\_struct} might need to be
-deleted and removed from the tree. All of these operations ensure that frag
-mmap pattern takes the most amount of time.
+rebalancing. Moreover, the existing \texttt{ vm\_area\_struct} needs to be
+removed from the tree. All of these operations ensure that the frag workload
+takes the most amount of time.
 
@@
 traversing the tree first. However, it does not need to delete an existing
-structure, because of which, it is faster than the frag benchmark.
+structure, which causes it to be faster than the frag benchmark.
 
@@
 mapping adjacent to the desired mapping which can be expanded, obviating the
-need to create a new struct and insert it into red-black tree. Therefore, most
-of the times the time taken by rand benchmark is lower than strided.
+need to create a new struct and insert it into red-black tree. Therefore, usually
+the time taken by rand benchmark is lower than strided.
 
@@
 
-The cont workloads write to adjacent 4KB pages in the address space. A huge page
+The cont workload writes to adjacent 4KB pages in the address space. A huge page
 of size 2MB is allocated when a new page is touched initially. Subsequent 4KB

### 2017-04-28 paper.tex
Added figure positions and worked on conclusions

@@
 
-\begin{figure}
+\begin{figure}[t]
 \centering
@@
 
-\begin{figure}
+\begin{figure}[t]
     \includegraphics[width=\columnwidth]{figures/mmap_latency}
@@
 
-\begin{figure}
+\begin{figure}[t]
     \includegraphics[width=\columnwidth]{figures/mmap_mem_usage}
@@
 
-\begin{figure}
+\begin{figure}[t]
     \includegraphics[width=\columnwidth]{figures/vm_area_struct_count}
@@
 
-\begin{figure}
+\begin{figure}[t]
     \includegraphics[width=\columnwidth]{figures/swap_rss}
@@
 
-\begin{figure}
+\begin{figure}[t]
     \includegraphics[width=\columnwidth]{figures/swap_touch_time_cont_cdf}
@@
 
-\begin{figure}
+\begin{figure}[t]
     \includegraphics[width=\columnwidth]{figures/swap_touch_time_rand_cdf}
@@
 
-\begin{figure}
+\begin{figure}[t]
     \includegraphics[width=\columnwidth]{figures/swap_time_cont}
@@
 
-\begin{figure}
+\begin{figure}[t]
     \includegraphics[width=\columnwidth]{figures/swap_time_rand}
@@
 
-\begin{figure}
+\begin{figure}[t]
     \includegraphics[width=\columnwidth]{figures/swap_time_cdf}
@@
 
-Large swap files should be created with disk locality or on SSDs.
+While large swap files allow greater amount of memory overcommitment, they can
+also reduce the degree of disk locality. Thus, large swap files should be
+created with disk locality in mind or on SSDs. Some Linux distributions, such as
+Ubuntu, have switched to using swap files rather than swap partitions by default
+(TODO: cite). Our results suggest that for some workloads this may have a
+negative impact on performance if the swapfile ends up in non-continuous disk
+locations.
 
-Looking at 99.99-th percentile latencies is important because they can have a
-large impact on overall run time.
+Also, looking at 99.99-th percentile latencies is important because they can
+have a large impact on overall run time. As our results show, even very uncommon
+high-latency operations may have a significant impact on performance.
 
-TODO: what else?
+\subsection{kswapd and khugepaged (TODO: markm)}
 
-\subsection{kswapd}
+TODO
 
-TODO: time: MARKM
+~\\ \textbf{Conclusions}
 
-TODO: khugepaged?
+TODO
 

### 2017-04-28 paper.tex
remove textt TODO
Numerous minor edits

@@
 
-\title{Journey: A Study of Scalabilty of Memory Management in Linux (TODO: should we get a new title?)}
-\author{Mark Mansi, Suhas Pai, Hasnain Ali Pirzada\\\texttt{\{markm, hp, spai\}@cs.wisc.edu}}
+\title{Journey: A Study of Scalabilty of Memory Management in Linux}
+\author{Mark Mansi, Suhas Pai, Hasnain Ali Pirzada\\\texttt{\{markm, spai, hp\}@cs.wisc.edu}}
 \date{}
@@
 On Linux, the system may have multiple swap spaces in the form of a swap disk
-partition or of swapfiles on a filesystem. Both have a similar format on disk,
+partition or of swap files on a filesystem. Both have a similar format on disk,
 in which the space is broken into a number of swap slots large enough for one
@@
 
-\item \textit{Continuous} (``cont'' for short). This workload allocates $2^{20}$
+\item \textit{Continuous} (\texttt{cont} for short). This workload allocates $2^{20}$
 adjacent 16KB blocks.
@@
 
-\item \textit{Fragmented} (``frag'' for short). This workload first allocates
+\item \textit{Fragmented} (\texttt{frag} for short). This workload first allocates
     $2^{21}$ contiguous 4KB blocks. Then, it resizes every other block to 16KB.
@@
 
-(TODO: we need to be consistent with whether we texttt the benchmark names or
-not)
-
 \subsection{\texttt{vm\_area\_struct} Experiments}
@@
 most recently used \texttt{vm\_area\_struct}, avoiding a search in the
-red-black tree. In the continuous benchmark, this cached struct is always
+red-black tree. In the \texttt{cont} benchmark, this cached struct is always
 conveniently used.
@@
 
-The frag benchmark takes the most time of any benchmark. When a region of the
+The \texttt{frag} benchmark takes the most time of any benchmark. When a region of the
 address space is being expanded, the kernel is forced to unmap it and map a new
@@
 rebalancing. Moreover, the existing \texttt{ vm\_area\_struct} needs to be
-removed from the tree. All of these operations ensure that the frag workload
+removed from the tree. All of these operations ensure that the \texttt{frag} workload
 takes the most amount of time.
 
-The strided workload is very similar to frag because it will also force the
+The \texttt{stride} workload is very similar to \texttt{frag} because it will also force the
 kernel to always create a new structure and insert it into the tree after
 traversing the tree first. However, it does not need to delete an existing
-structure, which causes it to be faster than the frag benchmark.
+structure, which causes it to be faster than the \texttt{frag} benchmark.
 
-Finally, the rand workload will also search through the red-black tree first,
+Finally, the \texttt{rand} workload will also search through the red-black tree first,
 but there is a reasonably high probability that it will find an existing
@@
 need to create a new struct and insert it into red-black tree. Therefore, usually
-the time taken by rand benchmark is lower than strided.
+the time taken by \texttt{rand} benchmark is lower than \texttt{stride}.
 
@@
     \includegraphics[width=\columnwidth]{figures/mmap_mem_usage}
-    \caption{$\Delta_i$ adjusted with control for benchmarks}
+    \caption{$\Delta_i$ adjusted with \texttt{control} for benchmarks}
     \label{fig:mmap_mem_usage}
@@
 
-The cont workload writes to adjacent 4KB pages in the address space. A huge page
+The \texttt{cont} workload writes to adjacent 4KB pages in the address space. A huge page
 of size 2MB is allocated when a new page is touched initially. Subsequent 4KB
@@
 
-The rand workload, on the other hand, writes to pages that are likely not
+The \texttt{rand} workload, on the other hand, writes to pages that are likely not
 adjacent and likely not subpages of a single huge page. As a result, a huge page
@@
 close to its peak value. As expected, the behavior is the same irrespective of
-the size of the swapfile.
+the size of the swap file.
 
@@
     \includegraphics[width=\columnwidth]{figures/swap_rss}
-    \caption{RSS while touching pages while swapping for random benchmark.}
+    \caption{RSS while touching pages while swapping for \texttt{rand} benchmark.}
     \label{swap_rss}
@@
     \includegraphics[width=\columnwidth]{figures/swap_touch_time_cont_cdf}
-    \caption{CDF of time to touch a page while swapping for continuous
+    \caption{CDF of time to touch a page while swapping for \texttt{cont}
     benchmark. Notice that the y-axis scale is from 99.75\% to 100\%.
@@
     \includegraphics[width=\columnwidth]{figures/swap_touch_time_rand_cdf}
-    \caption{CDF of time to touch a page while swapping for random benchmark.
+    \caption{CDF of time to touch a page while swapping for \texttt{rand} benchmark.
     Notice that the y-axis scale is from 93\% to 100\%.\label{fig:swap_time_rand_cdf}}
@@
     \includegraphics[width=\columnwidth]{figures/swap_time_cont}
-    \caption{Time to touch a page while swapping for continuous benchmark.
+    \caption{Time to touch a page while swapping for \texttt{cont} benchmark.
     Notice that the y-scale is logarithmic.\label{fig:swap_time_cont}}
@@
     \includegraphics[width=\columnwidth]{figures/swap_time_rand}
-    \caption{Time to touch a page while swapping for random
+    \caption{Time to touch a page while swapping for \texttt{rand}
     benchmark.\label{fig:swap_time_rand}}
@@
 
-The overall behavior is as expected. In the random benchmark, as before, all
+The overall behavior is as expected. In the \texttt{rand} benchmark, as before, all
 memory is exhausted quickly, and the benchmark induces swapping soon after the
@@
 swap file makes very little difference in this workload because the disk has to
-do random seeks regardless of swapfile size.
+do random seeks regardless of swap file size.
 
-The continuous benchmark has more surprising results. Notice that Figure
+The \texttt{cont} benchmark has more surprising results. Notice that Figure
 \ref{fig:swap_time_cont} has a logarithmic y-scale. The system only begins to
-swap after about 4M operations, as before. Unlike the random benchmark, we see a
+swap after about 4M operations, as before. Unlike the \texttt{rand} benchmark, we see a
 number of operations that take hundreds of milliseconds. Figure
@@
 
-We also find that in the continuous benchmark, a larger swap file actually has
+We also find that in the \texttt{cont} benchmark, a larger swap file actually has
 slightly more of these slow memory accesses. We believe this has to do with the
-fact that the large swapfile has less disk locality, and so causes more disk
+fact that the large swap file has less disk locality, and so causes more disk
 seeks when paging out. Figure \ref{fig:swap_time_cdf} shows that with a larger
-swap file, the continuous benchmark has a significantly higher number of
+swap file, the \texttt{cont} benchmark has a significantly higher number of
 high-latency accesses as compared with the smaller swap file.
@@
 (TODO: cite). Our results suggest that for some workloads this may have a
-negative impact on performance if the swapfile ends up in non-continuous disk
+negative impact on performance if the swap file ends up in non-continuous disk
 locations.

### 2017-04-28 paper.tex
add mmap latency conclusion

--- 2017-04-28 paper.tex
@@
 ~\\ \textbf{Conclusions}
+
+The latency of mapping memory region is lowest when the new memory is allocated
+next to an already allocated block. This latency is the worst when an already mapped region
+is to be expanded. We suspect that a lot of real world applications to show the fragmented 
+behavior when they need to expand their existing data structures. This furher highlights
+the imporance of optimistically leaving empty spaces around the allocated regions to 
+speed up memory expension. This would particularly be useful for performance crucial applications
+running atop compute infrastructures where memory is not a scarce resource.
 

### 2017-04-28 paper.tex
make changes to mmap latency conclusion

--- 2017-04-28 paper.tex
@@
 
-TODO?
+The kernel latency of m-mapping a memory region is lowest when the new memory is allocated
+next to an already allocated region. This latency is the worst in the frag microbenchmark i.e., when an already mapped region
+is to be expanded. We suspect a lot of real world applications to show the fragmented 
+behavior since it is common for applications to exapnd their existing data structures. 
+This finding highlights
+the imporance of optimistically leaving empty spaces around the allocated regions to 
+speed up memory expension. This practice would particularly be useful for performance crucial applications
+running atop compute infrastructures where memory is not a scarce resource but kernel is.
 

### 2017-04-28 paper.tex
add a comma in mmap latency conclusion

@@
 speed up memory expension. This practice would particularly be useful for performance crucial applications
-running atop compute infrastructures where memory is not a scarce resource but kernel is.
+running atop compute infrastructures where memory is not a scarce resource, but kernel is.
 

### 2017-04-28 paper.tex
added some citations

--- 2017-04-28 paper.tex
@@
 $i$-th measurement of $R$ during a run ($i = 0, 1, ..., 2^{20}$). System memory
-usage is known to jitter in Linux (TODO: cite). To adjust for this jitter, we
+usage is known to jitter in Linux \cite{jitter}. To adjust for this jitter, we
 define $\Delta_i = R_i - R_0$. For each operation, we take the median value of
@@
 need to create a new struct and insert it into red-black tree. Therefore, usually
-the time taken by rand benchmark is lower than strided.
+the time taken by \texttt{rand} benchmark is lower than \texttt{stride}.
 
 ~\\ \textbf{Conclusions}
-
-The latency of mapping memory region is lowest when the new memory is allocated
-next to an already allocated block. This latency is the worst when an already mapped region
-is to be expanded. We suspect that a lot of real world applications to show the fragmented 
-behavior when they need to expand their existing data structures. This furher highlights
-the imporance of optimistically leaving empty spaces around the allocated regions to 
-speed up memory expension. This would particularly be useful for performance crucial applications
-running atop compute infrastructures where memory is not a scarce resource.
 
@@
 Ubuntu, have switched to using swap files rather than swap partitions by default
-(TODO: cite). Our results suggest that for some workloads this may have a
-negative impact on performance if the swapfile ends up in non-continuous disk
+\cite{swap_ubuntu}. Our results suggest that for some workloads this may have a
+negative impact on performance if the swap file ends up in non-continuous disk
 locations.
@@
 page}, \texttt{struct vm\_area\_struct}, and page tables tend to comprise a
-large portion of memory management overhead \cite{simics} (TODO: cite the LWN
-paper).
+large portion of memory management overhead \cite{simics, struct_page}.
 

### 2017-04-28 paper.tex

@@
 
-TODO?
+The kernel latency of m-mapping a memory region is lowest when the new memory is allocated
+next to an already allocated region. This latency is the worst in the frag microbenchmark i.e., when an already mapped region
+is to be expanded. We suspect a lot of real world applications to show the fragmented 
+behavior since it is common for applications to exapnd their existing data structures. 
+This finding highlights
+the imporance of optimistically leaving empty spaces around the allocated regions to 
+speed up memory expension. This practice would particularly be useful for performance crucial applications
+running atop compute infrastructures where memory is not a scarce resource, but kernel is.
 

### 2017-04-28 paper.tex
Minor typo fixes

@@
 
-The kernel latency of m-mapping a memory region is lowest when the new memory is allocated
-next to an already allocated region. This latency is the worst in the frag microbenchmark i.e., when an already mapped region
-is to be expanded. We suspect a lot of real world applications to show the fragmented 
-behavior since it is common for applications to exapnd their existing data structures. 
-This finding highlights
-the imporance of optimistically leaving empty spaces around the allocated regions to 
-speed up memory expension. This practice would particularly be useful for performance crucial applications
-running atop compute infrastructures where memory is not a scarce resource, but kernel is.
+The kernel latency of creating a memory region is lowest when the new memory is
+allocated next to an already allocated region. This latency is the worst in the
+\texttt{frag} microbenchmark i.e., when an already mapped region is to be expanded. We
+suspect a lot of real world applications to show the fragmented behavior since
+it is common for applications to exapnd their existing data structures.  This
+finding highlights the imporance of optimistically leaving empty spaces around
+the allocated regions to speed up memory expension. This practice would
+particularly be useful for performance crucial applications running atop compute
+infrastructures where memory is not a scarce resource, but kernel is.
+
+(TODO: (markm) I am hesitant about these conclusions...)
 

### 2017-04-28 paper.tex
Fixed a couple of minor typos
Added a citation

@@
 
+\begin{figure}[t]
+    \includegraphics[width=\columnwidth]{figures/swap_rss}
+    \caption{RSS while touching pages while swapping for \texttt{rand} benchmark.
+    \label{fig:swap_rss}}
+\end{figure}
+
 Figure \ref{fig:swap_rss} shows the RSS of each workload for the
@@
 prevent early onset of memory pressure.
-
-\begin{figure}[t]
-    \includegraphics[width=\columnwidth]{figures/swap_rss}
-    \caption{RSS while touching pages while swapping for \texttt{rand} benchmark.}
-    \label{swap_rss}
-\end{figure}
 
@@
 of work already exists on hardware prefetching for processor caches, on which a
-swap prefetcher might draw for inspiration (TODO: cite). However, this approach
+swap prefetcher might draw for inspiration \cite{prefetching}. However, this approach
 can only work if page access patterns are predictable. Moreover, the high

### 2017-04-28 bench_measure_cpu_util_cpu.cc
changed benchmarks once again

@@ -28,11 +28,12 @@ int main(int argc, char **argv) {
 		return -1;
 	}
 
-    int kswapd_pid = atoi(argv[1]);
+    //int kswapd_pid = atoi(argv[1]);
 
     // touch all the pages now
     while (1) {
-        std::cout << get_pid_cpu_usage(kswapd_pid) << std::endl;
+        //TODO: fix get_pid_cpu_usage first!
+        //std::cout << get_pid_cpu_usage(kswapd_pid) << std::endl;
         usleep(500);  // I don't know how to tune this time. It makes a difference. 
     }
 

### 2017-04-28 rdmemusg.cc

@@ -13,8 +13,6 @@
 
 static char buf[2048];
 static int fd = -1;
-unsigned long kswapd_total_before = 0;
-unsigned long elapsed_before = 0;
 
 unsigned long getinfo(const char *file, const char *name) {
 	int local_n;
@@ -114,6 +112,7 @@ unsigned long getkernelmem() {
 	return tmem - fmem - pmem;
 }
 
+// TODO: this is no longer correct!!!
 double get_pid_cpu_usage(int pid) {
     unsigned long elapsed_delta = get_uptime_jiffies();
     unsigned long kswapd_delta = get_pid_jiffies(pid);
@@ -157,11 +156,7 @@ unsigned long get_pid_jiffies(int pid) {
 
     close(f);
 
-    unsigned long kswapd_total_now = u_time + s_time;
-    unsigned long kswapd_delta = kswapd_total_now - kswapd_total_before;
-    kswapd_total_before = kswapd_total_now;
-
-    return kswapd_delta;
+    return u_time + s_time;
 }
 
 unsigned long get_uptime_jiffies() {
@@ -201,10 +196,7 @@ unsigned long get_uptime_jiffies() {
     
     }
 
-    unsigned long elapsed_delta = total_uptime - elapsed_before;
-    elapsed_before = total_uptime;
-
-    return elapsed_delta;
+    return total_uptime;
 }
 
 unsigned long get_uptime() {

### 2017-04-28 rdmemusg.h

@@ -8,7 +8,8 @@ unsigned long getinfo(const char *file, const char *name);
 unsigned long meminfo(const char *name);
 unsigned long statusinfo(const char *name);
 unsigned long slabinfo(const char *name);
-double get_pid_cpu_usage(int);
+// TODO: fix the implementation before using this
+//double get_pid_cpu_usage(int);
 unsigned long get_pid_jiffies(int);
 unsigned long get_uptime_jiffies();
 unsigned long get_uptime();

### 2017-04-28 paper.tex
added jokes

@@
 
+A SQL query goes into a bar, walks up to two tables and asks, ``Can I join you?".
+
 \section{Results}
@@
 
-~\\ \textbf{Conclusions}
+When your hammer is C++, everything begins to look like a thumb.
+
+~\\ \textbf{Insights}
 
@@
 
-~\\ \textbf{Conclusions}
+~\\ \textbf{Insights}
 
@@
 
-~\\ \textbf{Conclusions}
+If you put a million monkeys at a million keyboards, one of them will eventually write a Java program. The rest of them will write Perl programs.
+
+~\\ \textbf{Insights}
 
@@
 
+I would tell you a UDP joke, but you might not get it.
+
 \subsection{kswapd and khugepaged (TODO: markm)}
@@
 existing technology \cite{quartz}. (TODO: is this system still relavent?)
+
+Two threads walk into a bar. The barkeeper looks up and yells, “hey, I want
+don’t any conditions race like time last!“
 
@@
 page}, \texttt{struct vm\_area\_struct}, and page tables tend to comprise a
-large portion of memory management overhead \cite{simics, struct_page}.
+large portion of memory management overhead \cite{simics, struct_page}. In conclusion, did you find all 6 jokes in the paper?
 

### 2017-04-29 paper.tex
Address todos

@@
 \maketitle
+
+\begin{abstract}
+    \small
+
+large memory systems
+
+study scalability of linux kernel as first step
+
+point to some results
+
+\end{abstract}
 
@@
 Currently, there are few good tools available to systems researchers for
-understanding how systems interact with large memories. We see the need for a
-general lightweight technique or tool for studying and designing operating
-systems capable of handling large memory while using systems currently
-available to researchers.  An important first step towards building such a tool
-is to identify bottlenecks in managing large memories that make a tool
-difficult to develop.
+understanding how systems interact with large memories. An important first step
+towards building such a tool is to identify bottlenecks in managing large
+memories that make a tool difficult to develop.
 
 In particular, we look at the case of a single process using a large amount of
-memory in Linux. We run experiments to identify the costs of
-\texttt{vm\_area\_struct}, a common bookkeeping data structure in the Linux
-kernel, along with the memory and processing overheads of swapping. From these
-data, we extrapolate to systems running with more memory with an eye on
-implications for studying large systems. While these experiments are not
-comprehensive, we believe they can offer insights that are useful both for
-building a research tool and for building systems.
+memory in Linux. We run experiments to identify the scalability of the Linux
+memory management system across multiple dimensions. First, we measure the
+overhead of \texttt{vm\_area\_struct}, a common bookkeeping data structure in
+the Linux kernel. Then, we measure the memory and processing overheads of
+swapping. From these data, we extrapolate to systems running with more memory
+with an eye on implications for studying large systems. While these experiments
+are not comprehensive, we believe they can offer insights that are useful both
+for building a research tool and for building systems.
 
@@
 allocate part of its virtual address space. Accesses to memory addresses
-outside of allocated regions result in a segmentation fault. This makes memory
-corruption errors easier to find and debug, and increases security from memory
-corruption attacks.
-
+outside of allocated regions result in a segmentation fault.
 The Linux kernel keeps track of which portions of a process's address space
@@
 
-We measure the memory overhead and latency of the Linux Memory Management
-system's use of \texttt{vm\_area\_struct}, which represents a region of a
-process's virtual address space. All the physical pages in the system are
-managed using the \texttt{struct page} descriptors in the kernel.  We do not
-measure the overhead of \texttt{struct page} because it is known that there is
-one struct for each physical page. We do not measure the overhead of page
-tables because this can be calculated from the size of the virtual memory space
-used by the process.  When a program makes a memory management system call
-such as \texttt{mmap}, \texttt{mremap}, \texttt{munmap}, \texttt{mprotect},
-\texttt{mlock}, or \texttt{madvise}, the kernel is creating, changing, or
-removing \texttt{vm\_area\_struct}s; thus, measuring the scalability of Linux
-memory management requires measuring the overhead of these structs.
+In our first set of experiments, we measure the memory overhead and latency of
+the Linux Memory Management system's use of \texttt{vm\_area\_struct}, which
+represents a region of a process's virtual address space. All the physical
+pages in the system are managed using the \texttt{struct page} descriptors in
+the kernel.  We do not measure the overhead of \texttt{struct page} because it
+is known that there is one struct for each physical page. We do not measure the
+overhead of page tables because this can be calculated from the size of the
+virtual memory space used by the process.  When a program makes a memory
+management system call such as \texttt{mmap}, \texttt{mremap}, \texttt{munmap},
+\texttt{mprotect}, \texttt{mlock}, or \texttt{madvise}, the kernel is creating,
+changing, or removing \texttt{vm\_area\_struct}s; thus, measuring the
+scalability of Linux memory management requires measuring the overhead of these
+structs.
 
@@
 Before swapping a page out to disk, the kernel must reserve a slot for the page
-in the swap space, unmap it from the process's address space, and record which
-swap slot corresponds to the page.
+in the swap space, unmap it from the process's address space by removing its
+page table entries and flushing the TLB, and record which swap slot corresponds
+to the page.
 
@@
 pages are freed to bring the free memory pool above some threashold. Then,
-kswapd goes back to sleep.  One concern with large memory systems is that kswapd
-may expend significant CPU time scanning through hundreds of thousands of pages.
+kswapd goes back to sleep.  One concern with systems that have a large memory
+usage is that kswapd may expend significant CPU time scanning through hundreds
+of thousands of pages.
 
@@
 
+% TODO: should we mention slab allocator?
 To measure memory usage, we use the procfs's reporting on current memory usage.
-Where appropriate, we record the slab allocator's count of active
+Where appropriate, we record the number of active
 \texttt{vm\_area\_struct}s in the system. Measuring kernel memory usage and
-associating memory usage with processes is difficult. Our methodology assumes
+associating memory usage with a particular process is difficult. Our methodology assumes
 that our benchmarks are running in an otherwise idle system and should dominate
@@
 
-First, when a continuous \texttt{mmap} (with the same permissions) happens the
+First, when a continuous \texttt{mmap} (with the same permissions) happens, the
 kernel memory allocator does not need to create a new \texttt{vm\_area\_struct}.
@@
 
-The kernel latency of creating a memory region is lowest when the new memory is
-allocated next to an already allocated region. This latency is the worst in the
-\texttt{frag} microbenchmark i.e., when an already mapped region is to be expanded. We
-suspect a lot of real world applications to show the fragmented behavior since
-it is common for applications to exapnd their existing data structures.  This
-finding highlights the imporance of optimistically leaving empty spaces around
-the allocated regions to speed up memory expension. This practice would
-particularly be useful for performance crucial applications running atop compute
-infrastructures where memory is not a scarce resource, but kernel is.
-
-(TODO: (markm) I am hesitant about these conclusions...)
+Linux's use of \texttt{vm\_area\_struct} tends to be scalable in the general
+case. The kernel latency of creating a memory region is lowest when the new
+memory is allocated next to an already allocated region. This latency is the
+worst in the \texttt{frag} microbenchmark i.e., when an already mapped region
+is to be expanded.  We recommend that applications that manage their own memory
+and expand memory regions should avoid this pattern.
 
@@
     \includegraphics[width=\columnwidth]{figures/mmap_latency}
-    \caption{Latency of memory operations}
+    \caption{Latency of memory operations (TODO: improve the scale to make log behavior more apparent)}
     \label{fig:mmap_latency}
@@
 is allocated with almost every write, and system memory is exhausted and begins
-swapping very quickly. Memory made available by kswapd is quickly consumed by
+swapping very quickly (20,000 operations compared to 4 million with the \texttt{cont} benchmark). Memory made available by kswapd is quickly consumed by
 further allocations. Hence, the memory used by the workload continues to stay
@@
 \ref{fig:swap_time_cdf} shows that these few accesses form a majority of run
-time; they impact performance significantly. (TODO: why do they happen?).
+time; they impact performance significantly.
 
@@
 
+Mark was telling Suhas a TCP joke the other day. He had to keep repeating it slower
+and slower until Suhas got it.
+
 Also, looking at 99.99-th percentile latencies is important because they can
@@
 
-I would tell you a UDP joke, but you might not get it.
+I would tell you a UDP joke, but you might not get it. But if you did get it, you
+wouldn't acknowledge it.
 
@@
 
-TODO
+TODO: in progress... should be ready for final product...
 
@@
 \section{Related Work}
+
+Initially, we had intended to build a tool for studying large memory systems.
+We found that the first step in doing so is studying the scalability of current
+memory management systems themselves. Thus, our related work section, while
+still relevant to our overall goal, is more directed at the second step of
+building the tool.
 
@@
 than on DRAM, it demonstrates that new technologies may be emulated using
-existing technology \cite{quartz}. (TODO: is this system still relavent?)
+existing technology \cite{quartz}.
 
-Two threads walk into a bar. The barkeeper looks up and yells, “hey, I want
-don’t any conditions race like time last!“
+Two threads walk into a bar. Bartender Hasnain looks up and yells, ``Hey, I want
+don’t any conditions race like time last!''
 
@@
 Interestingly, many of non-volatile memory technologies are making their ways
-into storage drives, such as Intel's recently-announced Optane SSDs. One
-potential use of fast storage devices is for fast swapping space. Some
-literature has proposed that small amounts of DRAM backed by large amounts of
-fast swap space can be a scalable alternative to having large amounts of fast
-RAM (TODO: cite). (TODO: can we have this in our report?). In such a system,
-the overhead of swapping may become a bottleneck.
+into storage drives. One potential use of fast storage devices is for fast
+swapping space. In such a system, the overhead of swapping may become a
+bottleneck.
 
@@
 of work already exists on hardware prefetching for processor caches, on which a
-swap prefetcher might draw for inspiration \cite{prefetching}. However, this approach
-can only work if page access patterns are predictable. Moreover, the high
-latency of disks implies that prefetchers would have to predict the very
-distant future (e.g. seconds ahead of execution). (TODO: Mike's comment on the
-previous version of this paragraph said that it was ``too distant/off topic''.
-Is it better now? I feel like prefetching was an idea that we considered too
-much to leave out...)
+swap prefetcher might draw for inspiration \cite{prefetching}. However, this
+approach can only work if page access patterns are predictable. Moreover, the
+high latency of disks currently implies that prefetchers would have to predict
+the very distant future (e.g. seconds ahead of execution); however, with
+non-volatile memory-based swap space, this idea may be more practical.
 
@@
 page}, \texttt{struct vm\_area\_struct}, and page tables tend to comprise a
-large portion of memory management overhead \cite{simics, struct_page}. In conclusion, did you find all 6 jokes in the paper?
+large portion of memory management overhead \cite{simics, struct_page}. In conclusion, did you find all 7 jokes in the paper?
 

### 2017-04-29 paper.tex
Addressed a couple more todos and added abstract

@@
 
-large memory systems
+Recent advances in memory technologies are making it possible for machines with
+    large amounts of memory to become commonplace.  Much research is currently
+    being done to study how non-volatile and large memory systems would change
+    system design.  We see the need for a lightweight tool to study such
+    systems at scale with real workloads.  The first step towards building such
+    a tool is to study the scalability of existing memory management systems.
 
-study scalability of linux kernel as first step
+We observe that in most cases, Linux kernel memory management scales well with
+    large amounts of physical memory. However, we observe that some workloads
+    can trigger poor behavior. In particular, workloads that allocated many
+    virtual memory areas can accumulate significant memory and computational
+    overhead of kernel metadata.
 
-point to some results
+We also observe that transparent hugepages can induce significantly earlier
+    onset of memory pressure in workloads that have random access patterns,
+    which may nullify the benefits from better TLB behavior. This can happen
+    even in workloads that would not otherwise experience memory pressure.
+
+Finally, we observe that the 99.99-th percentile of disk latency during swapping
+can account for over 70\% of runtime for the workloads we tested.
 
@@
 single structure. It is important for large applications such as object-oriented
-databases that map a large number of regions (TODO: Cite Linux book. Should we
-mention this?)to be aware of these overheads and mitigate them.
+databases that map a large number of regions to be aware of these overheads and mitigate them \cite{utlk}.
 
@@
 
+% TODO: we need to discuss implications for large memory in the insights sections more
+
 \section{Related Work}

### 2017-04-30 paper.tex
working no kswapd section

@@
 
-TODO: in progress... should be ready for final product...
+Figure (TODO) shows the total number of jiffies spent running kswapd as the
+number of operations performed increases. Figure (TODO) shows the same for
+khugepaged. Notice that one jiffy corresponds to 10ms of runtime.
+
+We find that although kswapd does not take up a significant
+percentage of runtime compared with the applcation, it does still take
+siginficant amounts of time and use system resources. This is especially true
+during times of high memory pressure.
+
+We also found that kswapd's behavior heavily depends on whether huge pages are
+enabled or not.
 

### 2017-05-01 paper.tex
Working on kswapd section

--- 2017-04-29 paper.tex
@@
 
-TODO: in progress... should be ready for final product...
+Figure (TODO) shows the total number of jiffies spent running kswapd as the
+number of operations performed increases. Figure (TODO) shows the same for
+khugepaged. Notice that one jiffy corresponds to 10ms of runtime.
+
+We find that khugepaged does not take significant computational power. (TODO:
+more) However, our workloads also are not conducive to allowing khugepaged to
+merge pages, and it apparently does not choose to merge 2MB pages into 1GB
+pages.
+
+Likewise, we find that kswapd does not take up a significant percentage of
+runtime compared with the applcation. The behavior of kswapd depends heavily on
+the workload and on whether transparent hugepages are enabled. Workloads that do
+not exhibit memory pressure do not incur kswapd overhead, as shown by the
+\texttt{rand} benchmark with hugepages off.
+
+As expected, the \texttt{rand} benchmark with hugepages on incurs the highest
+total overhead since it begins swapping very quickly. For this benchmark we
+observe an overhead of roughly 0.0002 jiffies per page memory touched. Figure
+also shows that when huge pages are enabled, kswapd runs for longer periods of
+time with longer sleeps in between. We suspect that kswapd swaps out several
+large pages at a time, which frees more memory at once than free normal 4kB
+pages, but due to time constraints, we have not verified this.
+
+In contrast, the \texttt{cont} benchmark with hugepages disabled uses
+computational time at roughly 0.0572 jiffies per page of memory touched.
+Moreover, kswapd runs more frequently and for shorter periods of time. We
+suspect that because pages are smaller, kswapd must run more often to reclaim
+the same amount of memory as with large pages.
+
+Notice also that the \texttt{cont} benchmark with hugepages enabled spends very
+little time in kswapd. This is somewhat surprising given the amount of time
+spent for the \texttt{rand} benchmark. We suspect that kswapd runs once and
+choose a large number of pages to swap out. Subsequently, both the workload and
+kswapd wait for I/O. This seems consistent with our previous results, in which
+some operations take on the order of seconds to complete.
 
@@
 
-TODO
+kswapd and the swapping system seem to have some interesting pathological cases
+that are worth exploring further.
+
+We expect that this number increases with the size of physical memory
+because kswapd will have more pages to scan, but due to time constraints, we
+have not verified this.
+
+TODO: anything else?
 

### 2017-05-02 paper.tex

--- 2017-04-29 paper.tex
@@
 
-I would tell you a UDP joke, but you might not get it. But if you did get it, you
+I would tell you a UDP joke, but you might not get it. Well, even if you got it, you
 wouldn't acknowledge it.
@@
 
+\section{Conclusion}
+Terabytes scale memory systems are poised to become commonplace in not too distant future.
+Developers and researchers need tools to study the behavior of such systems even before the actual
+deployments are avaialable. In this paper, we study the scalability of memory management 
+system of Linux Kernel which we believe is the first step toward building such tools. In particular, 
+we look at the scalability of \texttt{vm\_area\_struct},  
+swapping subsystem, and the effect of hugepages across a wide range of varying parameters
+and workloads. Our findings indicate that save for the fragmented mmmap pattern, 
+\texttt{vm\_area\_struct} tends to be sclable in the general case. We further find that the use of 
+transparent hugepages can lead to early onset of memory pressure for the random access pattern of applications. 
+This can lead to excessive swapping which might outweigh any potential benefits enabled by the use of hugepages. 
+We also quantify the degradation in performance due to the use of large swapping files and find that very large 
+swapping files should be used with caution becuase they can lead to very poor performance due to the loss of disk 
+locality.This aspact is more crucial in large scale memory system compared the the conventional systems.  
+While priliminary, we believe that our study is a first step towards building tools 
+to study large memory systems in a principled and informed manner.
+
 \bibliography{references}{}

### 2017-05-02 paper.tex
Update todo
Fix typo

@@
     \includegraphics[width=\columnwidth]{figures/mmap_latency}
-    \caption{Latency of memory operations (TODO: improve the scale to make log behavior more apparent)}
+    \caption{Latency of memory operations. A running average with window of
+    width 5000 operations is used to reduce jitter.}
     \label{fig:mmap_latency}
@@
 swapping subsystem, and the effect of hugepages across a wide range of varying parameters
-and workloads. Our findings indicate that save for the fragmented mmmap pattern, 
+and workloads. Our findings indicate that save for the fragmented mmap pattern, 
 \texttt{vm\_area\_struct} tends to be sclable in the general case. We further find that the use of 

### 2017-05-02 time_cdf.py
added time cdf scrip

@@ -1,0 +1,80 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib
+from sys import argv, exit
+
+if len(argv) < 5:
+    print("Usage: ./script WHICH AXIS_X AXIS_Y FILE [FILE, FILE, ...]")
+    print("WHICH = <column #>[,<column #>,...]")
+    print("AXIS_X = <x axis label>")
+    print("AXIS_Y = <y axis label>")
+    exit(-1)
+
+WHICH = [int(x) for x in argv[1].split(',')]
+AXIS_X = argv[2]
+AXIS_Y = argv[3]
+
+data = {}
+
+def running_mean(l, N):
+    sum = 0
+    result = list( 0 for x in l)
+
+    for i in range( 0, N ):
+        sum = sum + l[i]
+        result[i] = sum / (i+1)
+
+    for i in range( N, len(l) ):
+        sum = sum - l[i-N] + l[i]
+        result[i] = sum / N
+
+    return result
+
+for data_file in argv[4:]:
+    with open(data_file, 'r') as f:
+        data[data_file] = []
+
+        for line in f:
+            if len(line.strip()) == 0:
+                continue
+
+            xs = [float(x)/2.4E6 for x in line.split()]
+            #xs = [float(x) for x in line.split()]
+
+            data[data_file].append(xs)
+
+handles = []
+
+for data_file in sorted(data.keys()):
+    for i in WHICH:
+        if i < len(data[data_file][0]):
+            #ys = [x[i] - data[data_file][0][i] for x in data[data_file]]
+            # uncomment next line to prevent normalizing to X_0
+            #ys = running_mean([x[i] for x in data[data_file]], 5000)
+            ys = [x[i] for x in data[data_file]]
+            counts, bin_edges = np.histogram(ys, bins = int(max(ys)* 2.4E6/500), weights = ys)
+            cdf = np.cumsum(counts) / sum(ys) * 100.0
+            #xs = [float(i) / 1E3 for i in range(len(ys))]
+            line, = plt.plot(bin_edges[:-1], cdf, label = data_file)
+            #line, = plt.plot(cdf, label = data_file)
+            handles.append(line)
+
+#plt.xscale("log", basex = 2)
+#plt.yscale(SCALE)
+
+#plt.xlim(0, 512+32)
+#plt.ylim(0, 3)
+
+plt.xlabel(AXIS_X)
[…]

### 2017-05-02 paper.tex
Shortened abstract

@@
 
-Recent advances in memory technologies are making it possible for machines with
-    large amounts of memory to become commonplace.  Much research is currently
-    being done to study how non-volatile and large memory systems would change
-    system design.  We see the need for a lightweight tool to study such
-    systems at scale with real workloads.  The first step towards building such
-    a tool is to study the scalability of existing memory management systems.
+Much research is currently being done to study how non-volatile and large memory
+    systems would change system design.  We see the need for a lightweight tool
+    to study such systems at scale with real workloads.  The first step towards
+    building such a tool is to study the scalability of existing memory
+    management systems.
 
 We observe that in most cases, Linux kernel memory management scales well with
-    large amounts of physical memory. However, we observe that some workloads
-    can trigger poor behavior. In particular, workloads that allocated many
+    large amounts of physical memory. However, workloads that allocated many
     virtual memory areas can accumulate significant memory and computational

### 2017-05-02 paper.tex
Incorporated Mitch's comments

@@
     virtual memory areas can accumulate significant memory and computational
-    overhead of kernel metadata.
+    overhead due to kernel metadata.
 
@@
 overhead of page tables because this can be calculated from the size of the
-virtual memory space used by the process.  When a program makes a memory
+virtual memory space used by the process.
+
+When a program makes a memory
 management system call such as \texttt{mmap}, \texttt{mremap}, \texttt{munmap},
@@
 
-\subsection{\texttt{vm\_area\_struct}}
+\subsection{\texttt{vm\_area\_struct} Experiments}
 \label{ss_vm_area_struct}
@@
     $2^{21}$ contiguous 4KB blocks. Then, it resizes every other block to 16KB.
-        The resizing operation, rather than the initial allocation is measured.
+        The resizing operation, rather than the initial allocation, is measured.
 \end{itemize}
@@
 These experiments are designed to measure the minimum amount of memory needed
-to using swapping.
+by the swapping subsystem itself.
 
@@
 
-After each page of virtual memory is touched, the resident set size (RSS; the
-amount of physical memory Linux has allocated to the process) is recorded.  The
+After each page of virtual memory is touched, the resident set size (RSS) is
+recorded. The RSS is the
+amount of physical memory Linux has allocated to the process. The
 values of RSS are stored in an mlocked array and written to a file after the
@@
 
-A SQL query goes into a bar, walks up to two tables and asks, ``Can I join you?".
+%A SQL query goes into a bar, walks up to two tables and asks, ``Can I join you?".
 
@@
 
-Figure \ref{fig:mmap_latency} shows for various microbenchmarks the time taken
+Figure \ref{fig:mmap_latency} shows the time taken
 per \texttt{mmap} operation as it varies with the total amount memory allocated
@@
 
-First, when a continuous \texttt{mmap} (with the same permissions) happens, the
+First, when a continuous \texttt{mmap} happens, the
 kernel memory allocator does not need to create a new \texttt{vm\_area\_struct}.
@@
 
-Second, for continuous \texttt{mmap} operations, the time taken is always the
-lowest compared to other benchmarks.  This is because the kernel caches the
-most recently used \texttt{vm\_area\_struct}, avoiding a search in the
+Second, by caching the
+most recently used \texttt{vm\_area\_struct}, the kernel avoids a search in the
 red-black tree. In the \texttt{cont} benchmark, this cached struct is always
@@
 For the other workloads, a search through the red-black tree is often needed on
-most accesses. This operation takes $O(log n)$ time (where $n$ is the number of
-virtual memory areas). The logarithmic behavior is visible in the plots for the
+most accesses. This operation takes $O(log n)$ time, where $n$ is the number of
+virtual memory areas. The logarithmic behavior is visible in the plots for the
 benchmarks in Figure \ref{fig:mmap_latency}.
@@
 rebalancing. Moreover, the existing \texttt{ vm\_area\_struct} needs to be
-removed from the tree. All of these operations ensure that the \texttt{frag} workload
-takes the most amount of time.
+removed from the tree.
 
@@
 
-When your hammer is C++, everything begins to look like a thumb.
+%When your hammer is C++, everything begins to look like a thumb.
 
@@
 addresses. When a workload maps a large number of regions, this overhead can be
-of the order of hundreds of megabytes to gigabytes for large memory systems.  To
+on the order of hundreds of megabytes to gigabytes for large memory systems.  To
 reduce this overhead, it is crucial that regions with similar access permissions
@@
     \includegraphics[width=\columnwidth]{figures/mmap_mem_usage}
-    \caption{$\Delta_i$ adjusted with \texttt{control} for benchmarks}
+    \caption{$\Delta_i$ adjusted with \texttt{control} for benchmarks. Note that
+    graphs of \texttt{frag} and \texttt{stride} overlap. Also, \texttt{cont} and
+    \texttt{control} overlap.}
     \label{fig:mmap_mem_usage}
@@
     \includegraphics[width=\columnwidth]{figures/swap_rss}
-    \caption{RSS while touching pages while swapping for \texttt{rand} benchmark.
-    \label{fig:swap_rss}}
+    \caption{RSS while touching pages.  \label{fig:swap_rss}}
 \end{figure}
@@
 
-If you put a million monkeys at a million keyboards, one of them will eventually write a Java program. The rest of them will write Perl programs.
+% If you put a million monkeys at a million keyboards, one of them will eventually write a Java program. The rest of them will write Perl programs.
 
@@
 also reduce the degree of disk locality. Thus, large swap files should be
-created with disk locality in mind or on SSDs. Some Linux distributions, such as
+created with disk locality in mind. Some Linux distributions, such as
 Ubuntu, have switched to using swap files rather than swap partitions by default
@@
 
-Mark was telling Suhas a TCP joke the other day. He had to keep repeating it slower
-and slower until Suhas got it.
+% Mark was telling Suhas a TCP joke the other day. He had to keep repeating it slower
+% and slower until Suhas got it.
 
-Also, looking at 99.99-th percentile latencies is important because they can
+Also, looking at tail latencies is important because they can
 have a large impact on overall run time. As our results show, even very uncommon
@@
 
-I would tell you a UDP joke, but you might not get it. Well, even if you got it, you
-wouldn't acknowledge it.
+%I would tell you a UDP joke, but you might not get it. Well, even if you got it, you
+% wouldn't acknowledge it.
 
@@
 page}, \texttt{struct vm\_area\_struct}, and page tables tend to comprise a
-large portion of memory management overhead \cite{simics, struct_page}. In conclusion, did you find all 7 jokes in the paper?
+large portion of memory management overhead \cite{simics, struct_page}.
+
+%In conclusion, did you find all 7 jokes in the paper?
 
 \section{Conclusion}
-Terabytes scale memory systems are poised to become commonplace in not too distant future.
+Terabyte-scale memory systems are poised to become commonplace in not too distant future.
 Developers and researchers need tools to study the behavior of such systems even before the actual
-deployments are avaialable. In this paper, we study the scalability of memory management 
+deployments are avaialable.
+
+In this paper, we study the scalability of memory management 
 system of Linux Kernel which we believe is the first step toward building such tools. In particular, 
@@
 swapping subsystem, and the effect of hugepages across a wide range of varying parameters
-and workloads. Our findings indicate that save for the fragmented mmap pattern, 
-\texttt{vm\_area\_struct} tends to be sclable in the general case. We further find that the use of 
-transparent hugepages can lead to early onset of memory pressure for the random access pattern of applications. 
-This can lead to excessive swapping which might outweigh any potential benefits enabled by the use of hugepages. 
-We also quantify the degradation in performance due to the use of large swapping files and find that very large 
-swapping files should be used with caution becuase they can lead to very poor performance due to the loss of disk 
-locality.This aspact is more crucial in large scale memory system compared the the conventional systems.  
+and workloads.
+
+Our findings indicate that Linux's use of
+\texttt{vm\_area\_struct} tends to be sclable in the general case, but
+fragmented allocation patterns can cause increased computational and memory
+overheads. We further find that the use of 
+transparent hugepages can lead to early onset of memory pressure in applications
+with random access patterns. 
+This can lead to swapping which might outweigh any potential benefits enabled by the use of hugepages. 
+We also quantify the degradation in performance due to the use of large swap
+files. We propose that large swap files should be allocated with disk locality
+in mind to avoid performance degradation.
+
 While priliminary, we believe that our study is a first step towards building tools 

### 2017-05-03 paper.tex
Added acknowledgements section
reordered sections in results
Removed a couple of figures; acknowledgements

@@
 
+\begin{figure}[t]
+    \includegraphics[width=\columnwidth]{figures/mmap_latency}
+    \caption{Latency of memory operations. A running average with window of
+    width 5000 operations is used to reduce jitter.}
+    \label{fig:mmap_latency}
+\end{figure}
+
+\subsubsection{Memory}
+Figure \ref{fig:mmap_mem_usage} shows the growth in memory usage of the system
+for the workloads described in section \ref{ss_vm_area_struct}.
+
+The
+\texttt{cont} workload coincides with the \texttt{control} workload, as only the
+first \texttt{mmap} operation allocates a \texttt{vm\_area\_struct}. The subsequent
+operations are subsumed by the initial \texttt{vm\_area\_struct} because the mapped
+regions are adjacent, and hence the memory usage does not increase despite the
+large number of \texttt{mmap} operations.
+
+The \texttt{stride} workload allocates a
+\texttt{vm\_area\_struct} for each operation due to the separation of
+allocations in the virtual address space. As a result, the memory usage
+increases significantly as more memory is consumed with each \texttt{mmap} operation.
+
+This is also the case for the \texttt{frag} workload which breaks up a single
+\texttt{vm\_area\_struct} for a large area in the virtual address space into
+multiple smaller areas, which results in the allocation of a new structure for
+every resize operation.
+
+The \texttt{rand} workload initially creates a new
+\texttt{vm\_area\_struct} struct for each \texttt{mmap} operation. Eventually, these
+allocations form larger contiguous allocations when two or more regions are
+mapped next to each other, resulting in a single \texttt{vm\_area\_struct} being
+used for that range. As a result, the memory usage begins to taper off and
+even decreases slightly as multiple \texttt{vm\_area\_struct}s are collapsed to
+single structures. This is made clear by Figure \ref{fig:vm_area_struct_count}
+which shows the growth \texttt{vm\_area\_struct} counts for each workload.
+
+~\\ \textbf{Insights}
+
+The memory overhead of \texttt{vm\_area\_struct}s increases linearly with the
+number of mapped regions when these regions are separated by unused memory
+addresses. When a workload maps a large number of regions, this overhead can be
+on the order of hundreds of megabytes to gigabytes for large memory systems.  To
+reduce this overhead, it is crucial that regions with similar access permissions
+are placed adjacently in the address space so that they can be identified by a
+single structure. It is important for large applications such as object-oriented
+databases that map a large number of regions to be aware of these overheads and mitigate them \cite{utlk}.
+
 \subsubsection{Time}
@@
 and expand memory regions should avoid this pattern.
-
-\begin{figure}[t]
-    \includegraphics[width=\columnwidth]{figures/mmap_latency}
-    \caption{Latency of memory operations. A running average with window of
-    width 5000 operations is used to reduce jitter.}
-    \label{fig:mmap_latency}
-\end{figure}
-
-\subsubsection{Memory}
-Figure \ref{fig:mmap_mem_usage} shows the growth in memory usage of the system
-for the workloads described in section \ref{ss_vm_area_struct}.
-
-The
-\texttt{cont} workload coincides with the \texttt{control} workload, as only the
-first \texttt{mmap} operation allocates a \texttt{vm\_area\_struct}. The subsequent
-operations are subsumed by the initial \texttt{vm\_area\_struct} because the mapped
-regions are adjacent, and hence the memory usage does not increase despite the
-large number of \texttt{mmap} operations.
-
-The \texttt{stride} workload allocates a
-\texttt{vm\_area\_struct} for each operation due to the separation of
-allocations in the virtual address space. As a result, the memory usage
-increases significantly as more memory is consumed with each \texttt{mmap} operation.
-
-This is also the case for the \texttt{frag} workload which breaks up a single
-\texttt{vm\_area\_struct} for a large area in the virtual address space into
-multiple smaller areas, which results in the allocation of a new structure for
-every resize operation.
-
-The \texttt{rand} workload initially creates a new
-\texttt{vm\_area\_struct} struct for each \texttt{mmap} operation. Eventually, these
-allocations form larger contiguous allocations when two or more regions are
-mapped next to each other, resulting in a single \texttt{vm\_area\_struct} being
-used for that range. As a result, the memory usage begins to taper off and
-even decreases slightly as multiple \texttt{vm\_area\_struct}s are collapsed to
-single structures. This is made clear by Figure \ref{fig:vm_area_struct_count}
-which shows the growth \texttt{vm\_area\_struct} counts for each workload.
-
-~\\ \textbf{Insights}
-
-The memory overhead of \texttt{vm\_area\_struct}s increases linearly with the
-number of mapped regions when these regions are separated by unused memory
-addresses. When a workload maps a large number of regions, this overhead can be
-on the order of hundreds of megabytes to gigabytes for large memory systems.  To
-reduce this overhead, it is crucial that regions with similar access permissions
-are placed adjacently in the address space so that they can be identified by a
-single structure. It is important for large applications such as object-oriented
-databases that map a large number of regions to be aware of these overheads and mitigate them \cite{utlk}.
 
@@
 
-\begin{figure}[t]
-    \includegraphics[width=\columnwidth]{figures/swap_time_cont}
-    \caption{Time to touch a page while swapping for \texttt{cont} benchmark.
-    Notice that the y-scale is logarithmic.\label{fig:swap_time_cont}}
-\end{figure}
-
-\begin{figure}[t]
-    \includegraphics[width=\columnwidth]{figures/swap_time_rand}
-    \caption{Time to touch a page while swapping for \texttt{rand}
-    benchmark.\label{fig:swap_time_rand}}
-\end{figure}
+%\begin{figure}[t]
+%    \includegraphics[width=\columnwidth]{figures/swap_time_cont}
+%    \caption{Time to touch a page while swapping for \texttt{cont} benchmark.
+%    Notice that the y-scale is logarithmic.\label{fig:swap_time_cont}}
+%\end{figure}
+%
+%\begin{figure}[t]
+%    \includegraphics[width=\columnwidth]{figures/swap_time_rand}
+%    \caption{Time to touch a page while swapping for \texttt{rand}
+%    benchmark.\label{fig:swap_time_rand}}
+%\end{figure}
 
@@
 
+\section{Acknowledgements}
+
+We would like to thank Mitchell Manar, Akhil Guliani, Jing Liu, and Mayur
+Cherukuri for their useful and insightful reviews on our draft. We hope they
+enjoyed our jokes. Also, we would like to thank Mike Swift for his mentorship
+and guidance.
+
 \bibliography{references}{}

### 2017-05-04 paper.tex
Various changes; add spai's paragraph

@@
 In particular, we look at the case of a single process using a large amount of
-memory in Linux. We run experiments to identify the scalability of the Linux
-memory management system across multiple dimensions. First, we measure the
-overhead of \texttt{vm\_area\_struct}, a common bookkeeping data structure in
-the Linux kernel. Then, we measure the memory and processing overheads of
-swapping. From these data, we extrapolate to systems running with more memory
-with an eye on implications for studying large systems. While these experiments
-are not comprehensive, we believe they can offer insights that are useful both
-for building a research tool and for building systems.
+memory in a current Linux system. We run experiments to identify the scalability
+of the Linux memory management system across multiple dimensions. First, we
+measure the overhead of \texttt{vm\_area\_struct}, a common bookkeeping data
+structure in the Linux kernel. Then, we measure the memory and processing
+overheads of swapping. From these data, we extrapolate to systems running with
+more memory; many of the behaviors we identify would be aggravated by having or
+using more memory.
+
+While these experiments are not comprehensive, we believe they can offer
+insights that are useful both for building a research tool and for building
+systems. Also, we focus on large memory systems in general, rather than
+persistent memory system in particular.
 
@@
 pages are freed to bring the free memory pool above some threashold. Then,
-kswapd goes back to sleep.  One concern with systems that have a large memory
+kswapd goes back to sleep \cite{utlk}.  One concern with systems that have a large memory
 usage is that kswapd may expend significant CPU time scanning through hundreds
@@
 each operation in the benchmark. Notably, we do not touch the pages that we
-allocate, avoiding both a page fault and the allocation of a back physical page
+allocate, avoiding both a page fault and the allocation of a backing physical page
 for the allocated virtual pages. This allows us to measure just the overhead of
@@
 
-\item \textit{Random}. This workload allocates randomly the pages in a
-inside of a $2^{20}$ page region.
+\item \textit{Random}. This workload randomly allocates the pages in a
+$2^{20}$ page region.
 
@@
         The resizing operation, rather than the initial allocation, is measured.
+        By resizing every other block, rather than every block, the number of
+        worst-case operations is maximized because the kernel must move each
+        resized block.
 \end{itemize}
@@
 
-A single process allocates 18GB of virtual memory space and writes one byte of
+As in the previous set of experiments, a single process allocates 18GB of virtual memory space and writes one byte of
 each 4KB page in the region. The same two access patterns are tested as in the
@@
 Figure \ref{fig:mmap_latency} shows the time taken
-per \texttt{mmap} operation as it varies with the total amount memory allocated
+per \texttt{mmap} operation as it varies with the total amount of memory allocated
 so far by the workload.
@@
 
+(TODO: where to put this paragraph?)
+A study of recent changes to memory management in Linux has found that there has
+been a lot of focus on improving the scalability of data structures used in the
+Linux memory manager (TODO cite: Huang et. al. ATC'16). Since systems with
+persistent memory are expected to have large amounts of memory, there have been
+efforts to reduce the overheads of struct page. This is because these structures
+can consume several gigabytes of memory for systems with terabytes of persistent
+memory (TODO cite: https://lwn.net/Articles/640113/). These structures need to
+be stored in volatile memory not only because they need to be accessed quickly
+for performance reasons, but also to avoid wearing out of persistent memory due
+to frequent writes to this data structure. Patches proposed to the Linux kernel
+have proposed getting rid of the struct page altogether, but the complexity
+associated with such radical changes have hindered their acceptance by the
+community. Other works such as RadixVM \cite{radixvm} have also approached
+scalability with respect to the number of cores in the system.
+
 \section{Conclusion}

### 2017-05-04 paper.tex
shorten conclusion

@@
 
-We also observe that transparent hugepages can induce significantly earlier
+In addition, transparent hugepages can induce significantly earlier
     onset of memory pressure in workloads that have random access patterns,
@@
     even in workloads that would not otherwise experience memory pressure.
-
-Finally, we observe that the 99.99-th percentile of disk latency during swapping
+Finally, we see that the 99.99-th percentile of disk latency during swapping
     can account for over 70\% of runtime for the workloads we tested.

### 2017-05-04 paper.tex
refactor abstract

@@
 
-We observe that in most cases, Linux kernel memory management scales well with
+We find that in most cases, Linux kernel memory management scales well with
     large amounts of physical memory. However, workloads that allocated many
@@
     overhead due to kernel metadata.     
+    Furthermore, in certain situations, transparent hugepages can nullify the
+    benefits from better TLB reaches by inducing memory pressure earlier, even
+    for workloads that would otherwise not experience it.
 
-In addition, transparent hugepages can induce significantly earlier
-    onset of memory pressure in workloads that have random access patterns,
-    which may nullify the benefits from better TLB behavior. This can happen
-    even in workloads that would not otherwise experience memory pressure.
 Finally, we see that the 99.99-th percentile of disk latency during swapping

### 2017-05-04 paper.tex
Add references

--- 2017-05-04 paper.tex
@@
 (TODO: where to put this paragraph?)
-A study of recent changes to memory management in Linux has found that there has
-been a lot of focus on improving the scalability of data structures used in the
-Linux memory manager (TODO cite: Huang et. al. ATC'16). Since systems with
+A study of recent changes to memory management in Linux has found that there
+has been a lot of focus on improving the scalability of data structures used in
+the Linux memory manager \cite{huang2016evolutionary}. Since systems with
 persistent memory are expected to have large amounts of memory, there have been
-efforts to reduce the overheads of struct page. This is because these structures
-can consume several gigabytes of memory for systems with terabytes of persistent
-memory (TODO cite: https://lwn.net/Articles/640113/). These structures need to
-be stored in volatile memory not only because they need to be accessed quickly
-for performance reasons, but also to avoid wearing out of persistent memory due
-to frequent writes to this data structure. Patches proposed to the Linux kernel
-have proposed getting rid of the struct page altogether, but the complexity
-associated with such radical changes have hindered their acceptance by the
-community. Other works such as RadixVM \cite{radixvm} have also approached
-scalability with respect to the number of cores in the system.
+efforts to reduce the overheads of struct page. This is because these
+structures can consume several gigabytes of memory for systems with terabytes
+of persistent memory \cite{corbet_persistent_progress}. These structures need
+to be stored in volatile memory not only because they need to be accessed
+quickly for performance reasons, but also to avoid wearing out of persistent
+memory due to frequent writes to this data structure. Patches proposed to the
+Linux kernel have proposed getting rid of the struct page altogether, but the
+complexity associated with such radical changes have hindered their acceptance
+by the community. Other works such as RadixVM \cite{radixvm} have also
+approached scalability with respect to the number of cores in the system.
 

### 2017-05-04 paper.tex
one para long abstract

@@
     management systems. 
-
 We find that in most cases, Linux kernel memory management scales well with
@@
     for workloads that would otherwise not experience it.
-
 Finally, we see that the 99.99-th percentile of disk latency during swapping

### 2017-05-04 paper.tex
gq on abstract
Remove a line
gq on a paragraph
Remove a joke

--- 2017-05-04 paper.tex
@@
 
-Much research is currently being done to study how non-volatile and large memory
-    systems would change system design.  We see the need for a lightweight tool
-    to study such systems at scale with real workloads.  The first step towards
-    building such a tool is to study the scalability of existing memory
-    management systems.
-
-We find that in most cases, Linux kernel memory management scales well with
-    large amounts of physical memory. However, workloads that allocated many
-    virtual memory areas can accumulate significant memory and computational
-    overhead due to kernel metadata.     
-    Furthermore, in certain situations, transparent hugepages can nullify the
-    benefits from better TLB reaches by inducing memory pressure earlier, even
-    for workloads that would otherwise not experience it.
-
-Finally, we see that the 99.99-th percentile of disk latency during swapping
-    can account for over 70\% of runtime for the workloads we tested.
+Much research is currently being done to study how non-volatile and large
+memory systems would change system design.  We see the need for a lightweight
+tool to study such systems at scale with real workloads.  The first step
+towards building such a tool is to study the scalability of existing memory
+management systems.  We find that in most cases, Linux kernel memory management
+scales well with large amounts of physical memory. However, workloads that
+allocated many virtual memory areas can accumulate significant memory and
+computational overhead due to kernel metadata.     Furthermore, in certain
+situations, transparent hugepages can nullify the benefits from better TLB
+reaches by inducing memory pressure earlier, even for workloads that would
+otherwise not experience it.  Finally, we see that the 99.99-th percentile of
+disk latency during swapping can account for over 70\% of runtime for the
+workloads we tested.
 
@@
 through several rounds of scaling down, optimizing, changing, and tuning both
-the benchmarks and systems they test. For example, they scale down a workload to
-fit in the 1GB memory of the machine they use and increase the number of threads
-to improve parallelism \cite{2kmachine}. This methodology may be accurate, but
-it requires modifying the benchmarks, which is error prone and time consuming;
-it can be extremely difficult to validate changes to a benchmark. Ideally,
-researchers can run large experimental workloads without changes, but the overheads
-of current memory management systems prevent this from being practical.
+the benchmarks and systems they test. For example, they scale down a workload
+to fit in the 1GB memory of the machine they use and increase the number of
+threads to improve parallelism \cite{2kmachine}. This methodology may be
+accurate, but it requires modifying the benchmarks, which is error prone and
+time consuming.  Ideally, researchers can run large experimental workloads
+without changes, but the overheads of current memory management systems prevent
+this from being practical.
 
@@
 
-Two threads walk into a bar. Bartender Hasnain looks up and yells, ``Hey, I want
-don’t any conditions race like time last!''
+%Two threads walk into a bar. Bartender Hasnain looks up and yells, ``Hey, I want
+%don’t any conditions race like time last!''
 

### 2017-05-04 paper.tex
Remove a line
Remove not so relevant line

@@
 
-The Simics simulator uses demand paging for simulated environments, which means
-that memory is only allocated when it is used. As long as the working set of
+The Simics simulator uses demand paging for simulated environments.
+As long as the working set of
 the target system fits in the host memory, performance will be tolerable.
@@
 can be orders of magnitude slower than native execution, making it infeasible
-for studying large workloads or systems \cite{2kmachine}. However, the Simics
-simulator illustrates the usefulness of memory management techniques such as
-demand paging and the potential overheads of swapping.
+for studying large workloads or systems \cite{2kmachine}.
 

### 2017-05-05 paper.tex
Format struct page

@@
 persistent memory are expected to have large amounts of memory, there have been
-efforts to reduce the overheads of struct page. This is because these
+efforts to reduce the overheads of \texttt{struct page}. This is because these
 structures can consume several gigabytes of memory for systems with terabytes
@@
 memory due to frequent writes to this data structure. Patches proposed to the
-Linux kernel have proposed getting rid of the struct page altogether, but the
+Linux kernel have proposed getting rid of the \texttt{struct page} altogether, but the
 complexity associated with such radical changes have hindered their acceptance

### 2017-05-05 paper.tex
Remove for saving space

@@
 
-Interestingly, many of non-volatile memory technologies are making their ways
-into storage drives. One potential use of fast storage devices is for fast
-swapping space. In such a system, the overhead of swapping may become a
-bottleneck.
-
-Prefetching pages from swap space can offer a way to mitigate the overhead of
-memory overcommitment, possibly making large virtual machines a viable
-mechanism for studying large memory systems. When there is significant memory
-pressure, even pages which are likely to be accessed soon are swapped out to
-disk. They are faulted back into memory when accessed, resulting in significant
-performance cost.  Charm++ uses a programming model where computations can be
-scheduled by the language runtime. A designated thread can prefetch pages
-required by a computation, averting page faults \cite{charmpp}. Rather than
-implement page prefetching in a language runtime, one might implement a
-prefetcher in the swapping subsystem, making it general purpose. A large body
-of work already exists on hardware prefetching for processor caches, on which a
-swap prefetcher might draw for inspiration \cite{prefetching}. However, this
-approach can only work if page access patterns are predictable. Moreover, the
-high latency of disks currently implies that prefetchers would have to predict
-the very distant future (e.g. seconds ahead of execution); however, with
-non-volatile memory-based swap space, this idea may be more practical.
+%Interestingly, many of non-volatile memory technologies are making their ways
+%into storage drives. One potential use of fast storage devices is for fast
+%swapping space. In such a system, the overhead of swapping may become a
+%bottleneck.
+%
+%Prefetching pages from swap space can offer a way to mitigate the overhead of
+%memory overcommitment, possibly making large virtual machines a viable
+%mechanism for studying large memory systems. When there is significant memory
+%pressure, even pages which are likely to be accessed soon are swapped out to
+%disk. They are faulted back into memory when accessed, resulting in significant
+%performance cost.  Charm++ uses a programming model where computations can be
+%scheduled by the language runtime. A designated thread can prefetch pages
+%required by a computation, averting page faults \cite{charmpp}. Rather than
+%implement page prefetching in a language runtime, one might implement a
+%prefetcher in the swapping subsystem, making it general purpose. A large body
+%of work already exists on hardware prefetching for processor caches, on which a
+%swap prefetcher might draw for inspiration \cite{prefetching}. However, this
+%approach can only work if page access patterns are predictable. Moreover, the
+%high latency of disks currently implies that prefetchers would have to predict
+%the very distant future (e.g. seconds ahead of execution); however, with
+%non-volatile memory-based swap space, this idea may be more practical.
 
@@
 complexity associated with such radical changes have hindered their acceptance
-by the community. Other works such as RadixVM \cite{radixvm} have also
-approached scalability with respect to the number of cores in the system.
+by the community.
 

### 2017-05-05 paper.tex
edited kswapd sections

--- 2017-05-04 paper.tex
@@
 
-Much research is currently being done to study how non-volatile and large
-memory systems would change system design.  We see the need for a lightweight
-tool to study such systems at scale with real workloads.  The first step
-towards building such a tool is to study the scalability of existing memory
-management systems.  We find that in most cases, Linux kernel memory management
-scales well with large amounts of physical memory. However, workloads that
-allocated many virtual memory areas can accumulate significant memory and
-computational overhead due to kernel metadata.     Furthermore, in certain
-situations, transparent hugepages can nullify the benefits from better TLB
-reaches by inducing memory pressure earlier, even for workloads that would
-otherwise not experience it.  Finally, we see that the 99.99-th percentile of
-disk latency during swapping can account for over 70\% of runtime for the
-workloads we tested.
+Much research is currently being done to study how non-volatile and large memory
+    systems would change system design.  We see the need for a lightweight tool
+    to study such systems at scale with real workloads.  The first step towards
+    building such a tool is to study the scalability of existing memory
+    management systems. 
+We find that in most cases, Linux kernel memory management scales well with
+    large amounts of physical memory. However, workloads that allocated many
+    virtual memory areas can accumulate significant memory and computational
+    overhead due to kernel metadata.     
+    Furthermore, in certain situations, transparent hugepages can nullify the
+    benefits from better TLB reaches by inducing memory pressure earlier, even
+    for workloads that would otherwise not experience it.
+Finally, we see that the 99.99-th percentile of disk latency during swapping
+    can account for over 70\% of runtime for the workloads we tested.
 
@@
 % TODO: should we mention slab allocator?
+% (MM) I don't thinks so...
 To measure memory usage, we use the procfs's reporting on current memory usage.
@@
 repeat each experiment with hugepages enabled and disabled to see the impact of
-the transparent hugepage system. The processor usage of kswapd and khugepaged is
+the transparent hugepage system. The processor usage of kswapd is
 recorded for each experiment and reported in the next section.
@@
 Transparent huge pages can consume a large fraction of memory and lead to early
-onset of memory pressure when the workload accesses many pages that are
-separated in its address space. The overhead of swapping greatly exceeds the
+onset of memory pressure for workloads with random access pattners. The overhead of swapping greatly exceeds the
 performance benefits provided by huge pages. As a result, the workload
@@
 This experiment is designed to measure the impact of swapping on memory
-operation latency. Figure \ref{fig:swap_time_cont} and Figure
-\ref{fig:swap_time_rand} show the results.
+operation latency.
 
@@
 
+The latency of our disk leads us to expect that swapping out pages should take
+roughly 4ms. However, we see that in both benchmarks, there are operations that
+take dozens or hundreds of milliseconds. We used the \texttt{ftrace} kernel
+profiling mechanism in Linux to find out what causes some operations to be
+extremely slow.
+
+Surprisingly, the majority of this latency comes from kswapd itself. kswapd goes
+through phases of moving pages to the inactive list and reclaiming pages from
+the inactive list. This repeated until the amount of free memory increases
+beyond a threshold. kswapd does not do any particularly expensive operations
+beyond disk I/O. Rather, it executes a bunch of cheaper operations (e.g.
+checking if a page was referenced or freeing a page) millions of times in deeply
+nested loops. Meanwhile, the workload continues to use up any memory reclaimed
+by kswapd. Whenever kswapd is invoked by the page fault handler, it blocks the
+faulting process until enough pages are reclaimed.  Moreover, in some cases,
+during the \texttt{cont} workload kswapd chooses pages to reclaim much faster
+than they can be written to disk.
+
 ~\\ \textbf{Insights}
@@
 
-Also, looking at tail latencies is important because they can
-have a large impact on overall run time. As our results show, even very uncommon
-high-latency operations may have a significant impact on performance.
+Also, looking at tail latencies is important because they can have a large
+impact on overall run time. As our results show, even very uncommon high-latency
+operations may have a significant impact on performance.
+
+Our results also show that in some cases, kswapd induces more overhead to choose
+a good page to reclaim than would be accrued by choosing the wrong page to
+reclaim. Most of this overhead seems to be due to the poor asymptotic complexity
+of PFRA. Moreover, the swapping subsystem should be aware of disk bandwidth and
+latency, since they fundamentally limits how much memory can be reclaimed per
+unit time.
+
+Systems with larger memory may conceivably support more memory overcommitment,
+so improving the performance of swapping may be highly relavent moving forward.
+Additionally, if a system has orders of magnitude more pages, kswapd may spend
+significantly more time scanning the active and inactive lists.
 
@@
 
-\subsection{kswapd and khugepaged (TODO: markm)}
+\subsection{kswapd}
 
-TODO: in progress... should be ready for final product...
+Figure (TODO) shows the total number of jiffies spent running kswapd as the
+number of operations performed increases. Notice that one jiffy corresponds to
+10ms of runtime.
+
+First, we find that number of jiffies is a very imperfect measure of CPU runtime
+for kernel threads because it often misses executions of PFRA in the context of
+page faulting processes. Due to time constraints, we were not able to rerun our
+experiments using a better metric. However, these results do complement our
+findings from the previous set of experiments by showing the amount of time
+kswapd runs in its own context.
+
+kswapd does not take up a significant percentage of foreground runtime compared
+with the applcation. The behavior of kswapd depends heavily on the workload and
+on whether transparent hugepages are enabled. Not surprisingly, workloads that
+do not exhibit memory pressure do not incur kswapd overhead, as shown by the
+\texttt{rand} benchmark with hugepages off.
+
+As expected, the \texttt{rand} benchmark with hugepages on incurs the highest
+total overhead since it begins swapping very early. For this benchmark we
+observe an overhead of roughly 0.0002 jiffies per page memory touched after
+swapping begins. Also, kswapd runs for longer periods of time with longer
+intervals of sleep in between. This is most likely because the benchmark has
+some probability of retouching a page, so free memory is used more slowly, but
+kswapd also has to do more work to find pages that are inactive.
+
+In contrast, the \texttt{cont} benchmark with hugepages disabled uses
+computational time at roughly 0.0572 jiffies per page of memory touched.
+kswapd runs more frequently and for shorter periods of time because new pages
+are touched very quickly, requiring faster reclamation, but because all pages
+are only touched once, kswapd has to do very little work to choose pages to
+reclaim.
+
+Notice also that the \texttt{cont} benchmark with hugepages enabled spends
+surprisingly little time in kswapd. We suspect that kswapd runs once and choose
+a large number of pages to swap out. Subsequently, both the workload and kswapd
+wait for I/O. This seems consistent with our previous results, in which some
+operations take on the order of seconds to complete, but we have not yet
+verified it.
 
@@
 
-TODO
-
-% TODO: we need to discuss implications for large memory in the insights sections more
+The swapping system in Linux has complicated interactions with workloads and
+transparent hugepages. It is not obvious how to optimize a workload for better
+swapping behavior. Moreover, we see that under memory pressure, kswapd tends to
+run more often during page faults than in its own context.
 
@@
 through several rounds of scaling down, optimizing, changing, and tuning both
-the benchmarks and systems they test. For example, they scale down a workload
-to fit in the 1GB memory of the machine they use and increase the number of
-threads to improve parallelism \cite{2kmachine}. This methodology may be
-accurate, but it requires modifying the benchmarks, which is error prone and
-time consuming.  Ideally, researchers can run large experimental workloads
-without changes, but the overheads of current memory management systems prevent
-this from being practical.
+the benchmarks and systems they test. For example, they scale down a workload to
+fit in the 1GB memory of the machine they use and increase the number of threads
+to improve parallelism \cite{2kmachine}. This methodology may be accurate, but
+it requires modifying the benchmarks, which is error prone and time consuming;
+it can be extremely difficult to validate changes to a benchmark. Ideally,
+researchers can run large experimental workloads without changes, but the overheads
+of current memory management systems prevent this from being practical.
 
@@
 
-%Two threads walk into a bar. Bartender Hasnain looks up and yells, ``Hey, I want
-%don’t any conditions race like time last!''
+Two threads walk into a bar. Bartender Hasnain looks up and yells, ``Hey, I want
+don’t any conditions race like time last!''
 
-The Simics simulator uses demand paging for simulated environments.
-As long as the working set of
+The Simics simulator uses demand paging for simulated environments, which means
+that memory is only allocated when it is used. As long as the working set of
 the target system fits in the host memory, performance will be tolerable.
@@
 can be orders of magnitude slower than native execution, making it infeasible
-for studying large workloads or systems \cite{2kmachine}.
+for studying large workloads or systems \cite{2kmachine}. However, the Simics
+simulator illustrates the usefulness of memory management techniques such as
+demand paging and the potential overheads of swapping.
 

### 2017-05-05 paper.tex

@@
 
-Much research is currently being done to study how non-volatile and large memory
-    systems would change system design.  We see the need for a lightweight tool
-    to study such systems at scale with real workloads.  The first step towards
-    building such a tool is to study the scalability of existing memory
-    management systems. 
-We find that in most cases, Linux kernel memory management scales well with
-    large amounts of physical memory. However, workloads that allocated many
-    virtual memory areas can accumulate significant memory and computational
-    overhead due to kernel metadata.     
-    Furthermore, in certain situations, transparent hugepages can nullify the
-    benefits from better TLB reaches by inducing memory pressure earlier, even
-    for workloads that would otherwise not experience it.
-Finally, we see that the 99.99-th percentile of disk latency during swapping
-    can account for over 70\% of runtime for the workloads we tested.
+Much research is currently being done to study how non-volatile and large
+memory systems would change system design.  We see the need for a lightweight
+tool to study such systems at scale with real workloads.  The first step
+towards building such a tool is to study the scalability of existing memory
+management systems.  We find that in most cases, Linux kernel memory management
+scales well with large amounts of physical memory. However, workloads that
+allocated many virtual memory areas can accumulate significant memory and
+computational overhead due to kernel metadata.     Furthermore, in certain
+situations, transparent hugepages can nullify the benefits from better TLB
+reaches by inducing memory pressure earlier, even for workloads that would
+otherwise not experience it.  Finally, we see that the 99.99-th percentile of
+disk latency during swapping can account for over 70\% of runtime for the
+workloads we tested.
 
@@
 
-Two threads walk into a bar. Bartender Hasnain looks up and yells, ``Hey, I want
-don’t any conditions race like time last!''
+%Two threads walk into a bar. Bartender Hasnain looks up and yells, ``Hey, I want
+%don’t any conditions race like time last!''
 
-The Simics simulator uses demand paging for simulated environments, which means
-that memory is only allocated when it is used. As long as the working set of
+The Simics simulator uses demand paging for simulated environments.
+As long as the working set of
 the target system fits in the host memory, performance will be tolerable.
@@
 memory due to frequent writes to this data structure. Patches proposed to the
-Linux kernel have proposed getting rid of the struct page altogether, but the
+Linux kernel have proposed getting rid of the \texttt{struct page} altogether, but the
 complexity associated with such radical changes have hindered their acceptance
-by the community. Other works such as RadixVM \cite{radixvm} have also
-approached scalability with respect to the number of cores in the system.
+by the community.
 

### 2017-05-05 paper.tex
restructure related works

--- 2017-05-05 paper.tex
@@
 The emergence of new technology has in the past often required operating systems
-to change. Traditionally, OS researchers and developers benchmark and test their
+to change. A study of recent changes to memory management in Linux has found that there
+has been a lot of focus on improving the scalability of data structures used in
+the Linux memory manager \cite{huang2016evolutionary}. Since systems with
+persistent memory are expected to have large amounts of memory, there have been
+efforts to reduce the overheads of \texttt{struct page}. This is because these
+structures can consume several gigabytes of memory for systems with terabytes
+of persistent memory \cite{corbet_persistent_progress}. These structures need
+to be stored in volatile memory not only because they need to be accessed
+quickly for performance reasons, but also to avoid wearing out of persistent
+memory due to frequent writes to this data structure. Patches proposed to the
+Linux kernel have proposed getting rid of the \texttt{struct page} altogether.
+
+Traditionally, OS researchers and developers would benchmark and test such
 solutions in new environments to demonstrate their efficacy before using them in
@@
 
-(TODO: where to put this paragraph?)
-A study of recent changes to memory management in Linux has found that there
-has been a lot of focus on improving the scalability of data structures used in
-the Linux memory manager \cite{huang2016evolutionary}. Since systems with
-persistent memory are expected to have large amounts of memory, there have been
-efforts to reduce the overheads of \texttt{struct page}. This is because these
-structures can consume several gigabytes of memory for systems with terabytes
-of persistent memory \cite{corbet_persistent_progress}. These structures need
-to be stored in volatile memory not only because they need to be accessed
-quickly for performance reasons, but also to avoid wearing out of persistent
-memory due to frequent writes to this data structure. Patches proposed to the
-Linux kernel have proposed getting rid of the \texttt{struct page} altogether, but the
-complexity associated with such radical changes have hindered their acceptance
-by the community.
-
 \section{Conclusion}

### 2017-05-05 paper.tex
Remove apology paragraph

@@
 
-Initially, we had intended to build a tool for studying large memory systems.
-We found that the first step in doing so is studying the scalability of current
-memory management systems themselves. Thus, our related work section, while
-still relevant to our overall goal, is more directed at the second step of
-building the tool.
+%Initially, we had intended to build a tool for studying large memory systems.
+%We found that the first step in doing so is studying the scalability of current
+%memory management systems themselves. Thus, our related work section, while
+%still relevant to our overall goal, is more directed at the second step of
+%building the tool.
 

### 2017-05-05 paper.tex
Remove reference to deleted figure
Remove a line

@@
 A process may use the \texttt{mmap} system call to request that the kernel
-allocate part of its virtual address space. Accesses to memory addresses
-outside of allocated regions result in a segmentation fault.
+allocate part of its virtual address space.
 The Linux kernel keeps track of which portions of a process's address space
@@
 
-Under memory pressure, the kernel may choose to write some pages of memory to
-disk and then free those pages for use. Memory mapped files are written back to
-their backing files, but anonymous memory is written to portion of the disk
-called swap space. We focus on anonymous memory in our experiments.
+Under memory pressure, the kernel may free some pages for use by writing some
+pages to disk. Memory mapped files are written back to their backing files,
+but anonymous memory is written to portion of the disk called swap space. We
+focus on anonymous memory in our experiments.
 
@@
 
-The \texttt{cont} benchmark has more surprising results. Notice that Figure
-\ref{fig:swap_time_cont} has a logarithmic y-scale. The system only begins to
+The \texttt{cont} benchmark has more surprising results.
+The system only begins to
 swap after about 4M operations, as before. Unlike the \texttt{rand} benchmark, we see a

### 2017-05-05 paper.tex
Remove a line
Remove some lines

@@
 Recent advances in memory technologies are making it possible for machines with
-large amounts of memory to become commonplace. Many current systems already
-have dozens or hundreds of gigabytes of memory. These technology advances can
+large amounts of memory to become commonplace.
+These technology advances can
 enable systems with terabytes of memory in the not very distant future. For
@@
 subsystem. Choosing which pages to swap out is important because if the system
-swaps out pages that are about to be used, thrashing may be induced, in which
-the system wastes lots of time swapping out a page just swap it back in.
+swaps out pages that are about to be used, thrashing may be induced.
 
@@
 that memory has gone below a threshold. kswapd will continue to run until enough
-pages are freed to bring the free memory pool above some threashold. Then,
-kswapd goes back to sleep \cite{utlk}.  One concern with systems that have a large memory
+pages are freed to bring the free memory pool above some threshold before going to
+sleep \cite{utlk}.  One concern with systems that have a large memory
 usage is that kswapd may expend significant CPU time scanning through hundreds

### 2017-05-05 paper.tex
Further shortening of related works

--- 2017-05-05 paper.tex
@@
 through phases of moving pages to the inactive list and reclaiming pages from
-the inactive list. This repeated until the amount of free memory increases
+the inactive list. This repeats until the amount of free memory increases
 beyond a threshold. kswapd does not do any particularly expensive operations
@@
 The emergence of new technology has in the past often required operating systems
-to change. A study of recent changes to memory management in Linux has found that there
-has been a lot of focus on improving the scalability of data structures used in
-the Linux memory manager \cite{huang2016evolutionary}. Since systems with
-persistent memory are expected to have large amounts of memory, there have been
-efforts to reduce the overheads of \texttt{struct page}. This is because these
-structures can consume several gigabytes of memory for systems with terabytes
-of persistent memory \cite{corbet_persistent_progress}. These structures need
-to be stored in volatile memory not only because they need to be accessed
-quickly for performance reasons, but also to avoid wearing out of persistent
-memory due to frequent writes to this data structure. Patches proposed to the
-Linux kernel have proposed getting rid of the \texttt{struct page} altogether.
+to change. New non-volatile memory technologies are making it possible to have
+terabytes of memory. Recent changes to memory management in Linux demonstrate a
+lot of focus on improving the scalability of its data structures
+\cite{huang2016evolutionary}.
 
-Traditionally, OS researchers and developers would benchmark and test such
-solutions in new environments to demonstrate their efficacy before using them in
+There have been efforts to reduce the
+overheads of \texttt{struct page}s, which can comprise several gigabytes of
+overhead and must be kept in volatile memory both for performance reasons and to
+reduce wear on persistent memory cells \cite{corbet_persistent_progress,
+struct_page}. Other studies have suggested that page tables can also have high
+overhead \cite{simics}.
+
+RadixVM tries to overcome performance issues in highly
+concurrent workloads due to serialization of memory management operations on
+kernel data structures \cite{radixvm}. This work demonstrates that many parts of
+existing memory management schemes are not scalable to larger systems.
+Similarly, in our work, we examine scalability limitations of memory
+management in the Linux kernel.
+
+Traditionally, OS researchers and developers would benchmark and test 
+designs in new environments to demonstrate their efficacy before using them in
 real systems. However, researchers do not always have access to new technology
@@
 for most research budgets. This is not the first time that researchers have been
-forced to study systems in environments they do not have access to. This section
-examines past approaches to studying these systems and motivates the need for
-our work.
+forced to study systems in environments they do not have access to.
 
 The emergence of the internet required businesses to buy expensive machines to
-keep up with demand for their services, but systems researchers did not have
+keep up with demand for their services, but researchers did not have
 access to these machines. Alameldeen et al. describe how they simulate a
@@
 to fit in the 1GB memory of the machine they use and increase the number of
-threads to improve parallelism \cite{2kmachine}. This methodology may be
-accurate, but it requires modifying the benchmarks, which is error prone and
+threads to improve parallelism \cite{2kmachine}. This methodology requires modifying the benchmarks, which is error prone and
 time consuming.  Ideally, researchers can run large experimental workloads
@@
 
-Likewise, the Quartz emulator studies the behaviour of applications on
-Non-volatile Memory (NVM) while actually running on top of DRAM. It uses
-existing hardware capabilities to emulate the higher latency and lower bandwidth
-of most NVM technologies. While Quartz does not address the fact that the actual
-memory available to the applications on NVM might be orders of magnitude higher
-than on DRAM, it demonstrates that new technologies may be emulated using
-existing technology \cite{quartz}.
+%Likewise, the Quartz emulator studies the behaviour of applications on
+%Non-volatile Memory (NVM) while actually running on top of DRAM. It uses
+%existing hardware capabilities to emulate the higher latency and lower bandwidth
+%of most NVM technologies. While Quartz does not address the fact that the actual
+%memory available to the applications on NVM might be orders of magnitude higher
+%than on DRAM, it demonstrates that new technologies may be emulated using
+%existing technology \cite{quartz}.
 
@@
 the target system fits in the host memory, performance will be tolerable.
-However, in our work, we wish to study OS behavior under workloads that
-actually use all available memory. Thus, the usefulness of demand paging is
-greatly reduced because swapping causes disk latency to become the main
+However, in our work, we study OS behavior under workloads that
+actually use all available memory, where swapping causes disk latency to become the main
 performance bottleneck. Moreover, the performance overhead of simulators like
@@
 drives and large distributed storage systems. David is a system that allows
-storage and big data researchers to run large benchmarks requiring terabytes of
-storage using off the shelf storage devices (which at that time were too small).
-David creates a compressed version of the file system by physically storing only
-metadata and discarding the contents of files. Reads to the disk causes David to
-generate data on the fly. This decision choice is based on the observation that
+storage and big data researchers to run benchmarks requiring terabytes of
+storage using much smaller storage devices.
+David physically stores only
+metadata and discards the contents of files. Reads to the disk causes David to
+generate data on the fly. This design choice is based on the observation that
 most benchmarking frameworks do not care about the actual content of the files,
@@
 the contents of a process’s heap and only storing kernel data structures and a
-process’s code and stack segments. However, generating heap contents on the fly
-is more difficult than generating disk contents because of the common use of
-custom data structures. Moreover, on Linux, the kernel data structures for 1TB
+process’s code and stack segments. However, the common use of custom data
+structures makes generating heap contents on the fly
+more difficult than generating disk. Moreover, on Linux, the kernel data structures for 1TB
 of memory may also exceed the size of physical memory on current systems
@@
 virtual machine uses the whole amount of memory allocated to it, the host can
-swap pages to disk. The resulting poor performance can cause inaccurate
-performance measurements when running benchmarks. It can also make large
+swap pages to disk, resulting in poor performance, which can cause inaccurate
+performance measurements and make large
 benchmarks impractical.
@@
 
-Gupta et al. built an emulator for high speed networks using a technique called
-time dilation, which slows down the OS clock to make it appear that external
-events are occurring faster. This allows the system to emulate network links
-with speeds that are currently not available. The implementation is based on
-the VMs and Xen hypervisor; Xen delivers the timer interrupts to the guest at a
-lower rate than hardware hence  slowing down the guest’s clock \cite{timedil}.
-A similar approach may be applied for large memory system studies to slow down
-guest time while paging in and out large portions of memory. This will allows
-the system to believe that it is reading data from the memory while actually
-most of that data is being read from the disk.
-
-Finally, there have been studies which look at the performance overheads
-associated with current implementations of virtual memory and suggest mechanisms
-to mitigate them. RadixVM tries to overcome performance issues in highly
-concurrent workloads due to serialization of memory management operations on
-kernel data structures \cite{radixvm}. This work demonstrates that many parts of
-existing memory management schemes are not scalable to larger systems.
-Similarly, in our work, we wish to examine scalability limitations of memory
-management in the Linux kernel. Other studies have suggested that \texttt{struct
-page}, \texttt{struct vm\_area\_struct}, and page tables tend to comprise a
-large portion of memory management overhead \cite{simics, struct_page}.
+%Gupta et al. built an emulator for high speed networks using a technique called
+%time dilation, which slows down the OS clock to make it appear that external
+%events are occurring faster. This allows the system to emulate network links
+%with speeds that are currently not available. The implementation is based on
+%the VMs and Xen hypervisor; Xen delivers the timer interrupts to the guest at a
+%lower rate than hardware hence  slowing down the guest’s clock \cite{timedil}.
+%A similar approach may be applied for large memory system studies to slow down
+%guest time while paging in and out large portions of memory. This will allows
+%the system to believe that it is reading data from the memory while actually
+%most of that data is being read from the disk.
 

### 2017-05-05 paper.tex
Restore deleted line; it's important

@@
 subsystem. Choosing which pages to swap out is important because if the system
-swaps out pages that are about to be used, thrashing may be induced.
+swaps out pages that are about to be used, thrashing may be induced, in which
+the system wastes lots of time swapping out a page just swap it back in.
 

### 2017-05-05 paper.tex
Remove fake TODO
added missing figure

--- 2017-05-05 paper.tex
@@
 reaches by inducing memory pressure earlier, even for workloads that would
-otherwise not experience it.  Finally, we see that the 99.99-th percentile of
-disk latency during swapping can account for over 70\% of runtime for the
+otherwise not experience it.  Finally, we see that the tail
+swapping latencies can account for over a majority of runtime for the
 workloads we tested.
@@
     \includegraphics[width=\columnwidth]{figures/vm_area_struct_count}
-    \caption{Number of \texttt{vm\_area\_struct}s for benchmarks}
+    \caption{Number of \texttt{vm\_area\_struct}s for benchmarks.}
     \label{fig:vm_area_struct_count}
@@
 
+\begin{figure}[t]
+    \includegraphics[width=\columnwidth]{figures/kswapd_jiffies}
+    \caption{Number of jiffies spent by kswapd.\label{fig:kswapd_jiffies}}
+\end{figure}
+
 \subsection{kswapd}
 
-Figure (TODO) shows the total number of jiffies spent running kswapd as the
+Figure \ref{fig:kswapd_jiffies} shows the total number of jiffies spent running kswapd as the
 number of operations performed increases. Notice that one jiffy corresponds to

### 2017-05-05 paper.tex

@@
 
-% TODO: should we mention slab allocator?
-% (MM) I don't thinks so...
 To measure memory usage, we use the procfs's reporting on current memory usage.

### 2017-05-05 paper.tex
addressed most of Karan's concerns

@@
 For demand paging to work efficiently, the correct \texttt{vm\_area\_struct}
-needs to be efficient to find. Linux caches the most recently used region and
+needs to be found quickly. Linux caches the most recently used region and
 organizes all regions into a red-black tree for fast lookup.  Moreover,
@@
 
-In our first set of experiments, we measure the memory overhead and latency of
-the Linux Memory Management system's use of \texttt{vm\_area\_struct}, which
-represents a region of a process's virtual address space. All the physical
-pages in the system are managed using the \texttt{struct page} descriptors in
-the kernel.  We do not measure the overhead of \texttt{struct page} because it
-is known that there is one struct for each physical page. We do not measure the
-overhead of page tables because this can be calculated from the size of the
-virtual memory space used by the process.
+%In our first set of experiments, we measure the memory overhead and latency of
+%the Linux Memory Management system's use of \texttt{vm\_area\_struct}, which
+%represents a region of a process's virtual address space. All the physical
+%pages in the system are managed using the \texttt{struct page} descriptors in
+%the kernel.  We do not measure the overhead of \texttt{struct page} because it
+%is known that there is one struct for each physical page. We do not measure the
+%overhead of page tables because this can be calculated from the size of the
+%virtual memory space used by the process.
 
@@
 The specifications of the machine we used to run our experiments are listed in
-Figure \ref{fig:specs}. We disabled Intel SpeedStep and TurboBoost and ensure
+Figure \ref{fig:specs}. We disabled Intel SpeedStep and TurboBoost and set the scaling governor to ensure
 that the processor frequency stays constant. All experiments are pinned to the
@@
 running beyond services that start when the system is booted and a single ssh
-session. We are careful to avoid opening extra ssh sessions, running screen or
-tmux, or executing extra commands while a benchmark is running since these may
+session. We are careful to avoid running any additional processes, which could 
 impact memory usage and, thus, our results.
@@
 
+\subsubsection{Memory}
+
 \begin{figure}[t]
-    \includegraphics[width=\columnwidth]{figures/mmap_latency}
-    \caption{Latency of memory operations. A running average with window of
-    width 5000 operations is used to reduce jitter.}
-    \label{fig:mmap_latency}
+    \includegraphics[width=\columnwidth]{figures/mmap_mem_usage}
+    \caption{$\Delta_i$ adjusted with \texttt{control} for benchmarks. Note that
+    graphs of \texttt{frag} and \texttt{stride} overlap. Also, \texttt{cont} and
+    \texttt{control} overlap.}
+    \label{fig:mmap_mem_usage}
 \end{figure}
 
-\subsubsection{Memory}
+\begin{figure}[t]
+    \includegraphics[width=\columnwidth]{figures/vm_area_struct_count}
+    \caption{Number of \texttt{vm\_area\_struct}s for benchmarks.}
+    \label{fig:vm_area_struct_count}
+\end{figure}
+
 Figure \ref{fig:mmap_mem_usage} shows the growth in memory usage of the system
@@
 number of mapped regions when these regions are separated by unused memory
-addresses. When a workload maps a large number of regions, this overhead can be
+addresses.
+In the worst case, where each memory region is only one page long, \texttt{vm\_area\_struct} would represent a 5\% memory overhead per page.
+When a workload maps a large number of regions, this overhead can be
 on the order of hundreds of megabytes to gigabytes for large memory systems.  To
@@
 \subsubsection{Time}
+
+\begin{figure}[t]
+    \includegraphics[width=\columnwidth]{figures/mmap_latency}
+    \caption{Latency of memory operations. A running average with window of
+    width 5000 operations is used to reduce jitter.}
+    \label{fig:mmap_latency}
+\end{figure}
 
@@
 and expand memory regions should avoid this pattern.
-
-\begin{figure}[t]
-    \includegraphics[width=\columnwidth]{figures/mmap_mem_usage}
-    \caption{$\Delta_i$ adjusted with \texttt{control} for benchmarks. Note that
-    graphs of \texttt{frag} and \texttt{stride} overlap. Also, \texttt{cont} and
-    \texttt{control} overlap.}
-    \label{fig:mmap_mem_usage}
-\end{figure}
-
-\begin{figure}[t]
-    \includegraphics[width=\columnwidth]{figures/vm_area_struct_count}
-    \caption{Number of \texttt{vm\_area\_struct}s for benchmarks.}
-    \label{fig:vm_area_struct_count}
-\end{figure}
 
@@
     latency.\label{fig:swap_time_cdf}}
+\end{figure}
+
+\begin{figure}[t]
+    \centering
+    \begin{tabular}{c||c|c}
+        & \% of ops & Time spent \\ \hline \hline
+        Latency $\le$ 50ms & 99.97\% & 7.39s \\
+        Latency $>$ 50ms & 0.03\% & 18.9s \\
+    \end{tabular}
 \end{figure}

### 2017-05-05 paper.tex
fix abstract

@@
 
-Much research is currently being done to study how non-volatile and large
+Much research is currently being done on how non-volatile and large
 memory systems would change system design.  We see the need for a lightweight
-tool to study such systems at scale with real workloads.  The first step
+tool to study such systems at full scale with real workloads. The first step
 towards building such a tool is to study the scalability of existing memory
@@
 scales well with large amounts of physical memory. However, workloads that
-allocated many virtual memory areas can accumulate significant memory and
-computational overhead due to kernel metadata.     Furthermore, in certain
-situations, transparent hugepages can nullify the benefits from better TLB
-reaches by inducing memory pressure earlier, even for workloads that would
-otherwise not experience it.  Finally, we see that the tail
-swapping latencies can account for over a majority of runtime for the
+allocate many virtual memory areas can accumulate significant memory and
+computational overhead due to kernel metadata. Furthermore, in some cases, 
+transparent hugepages can nullify the benefits from better TLB
+reach by inducing memory pressure earlier. Finally, we see that the tail
+swapping latencies can account for a majority of runtime for the
 workloads we tested.

### 2017-05-05 paper.tex
Fix introduction on page 1

--- 2017-05-05 paper.tex
@@
 
-Recent advances in memory technologies are making it possible for machines with
-large amounts of memory to become commonplace.
-These technology advances can
-enable systems with terabytes of memory in the not very distant future. For
-example, non-volatile memory technologies, which have denser physical
-structures than traditional DRAM are making such large memories available and
-more accessible \cite{xpoint}.
+Recent advances in memory technologies can enable systems with terabytes of
+memory to become commonplace in the not very distant future. For example,
+non-volatile memory technologies, which have denser physical structures than
+traditional DRAM are making large memories available and more accessible
+\cite{xpoint}.
 
-Much research is currently being done to study how non-volatile and large
+Much research is currently being done on how non-volatile and large
 memory systems would change system design. HP has been working on a system
-which it calls ``The Machine'', in which hundreds of cores have access to a
+called ``The Machine'', in which hundreds of cores have access to a
 multi-petabyte shared memory pool \cite{hp_machine}. Intel and
@@
 In particular, we look at the case of a single process using a large amount of
-memory in a current Linux system. We run experiments to identify the scalability
-of the Linux memory management system across multiple dimensions. First, we
-measure the overhead of \texttt{vm\_area\_struct}, a common bookkeeping data
-structure in the Linux kernel. Then, we measure the memory and processing
-overheads of swapping. From these data, we extrapolate to systems running with
-more memory; many of the behaviors we identify would be aggravated by having or
-using more memory.
+memory in a current Linux system. We run experiments to identify the
+scalability of the Linux memory management system across multiple dimensions.
+First, we measure the overhead of \texttt{vm\_area\_struct}, a common
+bookkeeping data structure in the Linux kernel. Then, we measure the memory and
+processing overheads of swapping.  Many of the behaviors we identify would be
+aggravated by having or using more memory.
 

### 2017-05-05 paper.tex
Fix page 3

@@
 
-The PFRA runs primarily as a kernel thread called kswapd. kswapd runs at regular
-intervals and sleeps in between. Also, it may be invoked if the system detects
-that memory has gone below a threshold. kswapd will continue to run until enough
-pages are freed to bring the free memory pool above some threshold before going to
-sleep \cite{utlk}.  One concern with systems that have a large memory
-usage is that kswapd may expend significant CPU time scanning through hundreds
-of thousands of pages.
+The PFRA runs primarily as a kernel thread called kswapd. kswapd runs at
+regular intervals and sleeps in between. Also, it is invoked if the system
+detects that memory has gone below a threshold such as during a page fault.
+kswapd will continue to run until enough pages are freed to bring the free
+memory pool above some threshold before going to sleep \cite{utlk}.  This may
+cause the faulting process to block for a while. One concern with systems that
+have a large memory usage is that kswapd may expend significant CPU time
+scanning through hundreds of thousands of pages.
 
@@
 
-Assume the memory usage of the rest of the system is $R$. We denote $R_i$ as the
-$i$-th measurement of $R$ during a run ($i = 0, 1, ..., 2^{20}$). System memory
-usage is known to jitter in Linux \cite{jitter}. To adjust for this jitter, we
-define $\Delta_i = R_i - R_0$. For each operation, we take the median value of
+Assume the memory usage of the rest of the system is $R$. We denote $R_i$ as
+the $i$-th measurement of $R$ during a run ($i = 0, 1, ..., 2^{20}$). $R_0$
+corresponds to the measurement before the first operation. System memory usage
+is known to jitter in Linux \cite{jitter}. To adjust for this jitter, we define
+$\Delta_i = R_i - R_0$. For each operation, we take the median value of
 $\Delta_i$ across the 5 runs to reduce jitter further.  This is the value

### 2017-05-05 paper.tex
modify page2

--- 2017-05-05 paper.tex
@@
 In our methodology, we run several experiments to measure the scalability of
-the Linux memory management system. Each experiment is designed to probe a
-different aspect of the system. This section gives some background knowledge
+the Linux memory management system. This section gives some background knowledge
 necessary to understanding the experiments.
@@
 and the processor traps to a page fault handler, which allocates memory and
-sets up a page table entry for a valid access. This is called demand paging.
+sets up a page table entry for a valid access. This is called \texttt{demand paging}.
 
@@
 
-Under memory pressure, the kernel may free some pages for use by writing some
-pages to disk. Memory mapped files are written back to their backing files,
-but anonymous memory is written to portion of the disk called swap space. We
+Under memory pressure, the kernel may free some pages for use by writing
+them to disk. Memory mapped files are written back to their backing files,
+but anonymous memory is written to a portion of the disk called swap space. We
 focus on anonymous memory in our experiments.
@@
 Before swapping a page out to disk, the kernel must reserve a slot for the page
-in the swap space, unmap it from the process's address space by removing its
-page table entries and flushing the TLB, and record which swap slot corresponds
+in the swap space; unmap it from the process's address space by removing its
+page table entries and flushing the TLB; and record which swap slot corresponds
 to the page.
 
-On Linux, the system may have multiple swap spaces in the form of a swap disk
-partition or of swap files on a filesystem. Both have a similar format on disk,
+On Linux, the system may have multiple swap spaces in the form of swap disk
+partitions or swap files on a filesystem. Both have a similar format on disk,
 in which the space is broken into a number of swap slots large enough for one
-page. For each of these swap spaces, the kernel keeps a swap map, which records
+4K page. For each of these swap spaces, the kernel keeps a swap map, which records
 the usage of each swap slot so that it can find and allocate them quickly.
@@
 swaps out pages that are about to be used, thrashing may be induced, in which
-the system wastes lots of time swapping out a page just swap it back in.
+the system wastes time swapping out a page just to swap it back in.
 
@@
 twice. Roughly speaking, if a page is not used between scans, then it is a
-condidate for reclamation.
+candidate for reclamation.
 

### 2017-05-05 paper.tex
change page 4

@@
 \begin{itemize} \item \textit{Control}. This workload is only run for the memory
-overhead experiments. It simply prints the $\Delta_i$ for each $i$. It's graph
+overhead experiments. It simply prints the $\Delta_i$ for each $i$ while performing nops. It's graph
 is a straight line with positive slope. This memory usage results from the
@@
 
-\item \textit{Continuous} (\texttt{cont} for short). This workload allocates $2^{20}$
+\item \textit{Continuous} (\texttt{cont}). This workload allocates $2^{20}$
 adjacent 16KB blocks.
 
-\item \textit{Strided}. This workload allocates $2^{20}$ untouching 16KB regions
+\item \textit{Strided} (\texttt{stride}). This workload allocates $2^{20}$ untouching 16KB regions
 in order of increasing memory address.
 
-\item \textit{Random}. This workload randomly allocates the pages in a
+\item \textit{Random} (\texttt{rand}). This workload randomly allocates the pages in a
 $2^{20}$ page region.
 
-\item \textit{Fragmented} (\texttt{frag} for short). This workload first allocates
+\item \textit{Fragmented} (\texttt{frag}). This workload first allocates
     $2^{21}$ contiguous 4KB blocks. Then, it resizes every other block to 16KB.
@@
 benchmark completes. This prevents the writing of the results from interacting
-with paging in the kernel page cache. These RSS values are graphed in our
+with paging in the kernel page cache during the experiment. These RSS values are graphed in our
 results section.
@@
 These experiments are designed to measure the processing overhead of kswapd when
-the system is under high memory pressure.
+the system is under memory pressure.
 

### 2017-05-05 paper.tex
Fix figure captions
section 4.1.1

--- 2017-05-05 paper.tex
@@
 \texttt{cont} workload coincides with the \texttt{control} workload, as only the
-first \texttt{mmap} operation allocates a \texttt{vm\_area\_struct}. The subsequent
-operations are subsumed by the initial \texttt{vm\_area\_struct} because the mapped
-regions are adjacent, and hence the memory usage does not increase despite the
-large number of \texttt{mmap} operations.
+first \texttt{mmap} operation allocates a \texttt{vm\_area\_struct}. Subsequent
+operations expand this \texttt{vm\_area\_struct} because the mapped
+regions are adjacent, avoiding extra memory usage.
 
@@
 allocations in the virtual address space. As a result, the memory usage
-increases significantly as more memory is consumed with each \texttt{mmap} operation.
-
+increases linearly with the number of \texttt{mmap}s.
 This is also the case for the \texttt{frag} workload which breaks up a single
-\texttt{vm\_area\_struct} for a large area in the virtual address space into
-multiple smaller areas, which results in the allocation of a new structure for
-every resize operation.
+\texttt{vm\_area\_struct} for a large area into
+multiple smaller areas.
 
@@
 allocations form larger contiguous allocations when two or more regions are
-mapped next to each other, resulting in a single \texttt{vm\_area\_struct} being
-used for that range. As a result, the memory usage begins to taper off and
+mapped next to each other, resulting in a single \texttt{vm\_area\_struct}. As a result, the memory usage begins to taper off and
 even decreases slightly as multiple \texttt{vm\_area\_struct}s are collapsed to
 single structures. This is made clear by Figure \ref{fig:vm_area_struct_count}
-which shows the growth \texttt{vm\_area\_struct} counts for each workload.
+which shows the number of \texttt{vm\_area\_struct}s for each workload.
 

### 2017-05-05 paper.tex
Fix some part of 4.1.2

--- 2017-05-05 paper.tex
@@
     \includegraphics[width=\columnwidth]{figures/mmap_mem_usage}
-    \caption{$\Delta_i$ adjusted with \texttt{control} for benchmarks. Note that
+    \caption{$\Delta_i$ as discussed in section \ref{ss_vm_area_struct}. Note that
     graphs of \texttt{frag} and \texttt{stride} overlap. Also, \texttt{cont} and
@@
     \includegraphics[width=\columnwidth]{figures/vm_area_struct_count}
-    \caption{Number of \texttt{vm\_area\_struct}s for benchmarks.}
+    \caption{Number of \texttt{vm\_area\_struct}s. The graphs of
+    \texttt{control} and \texttt{cont} workloads are almost coincident with the
+    x-axis.}
     \label{fig:vm_area_struct_count}
@@
 
-It can be seen that for continuous \texttt{mmap}s the time is not only the
-lowest but is also invariant with the total number of \texttt{mmap}s done so
-far. There are two reasons for this.
+For continuous \texttt{mmap}s the time is not only the lowest of any benchmark
+but is also invariant with the total number of \texttt{mmap}s done so far.
+There are two reasons for this.
 
@@
 For the other workloads, a search through the red-black tree is often needed on
-most accesses. This operation takes $O(log n)$ time, where $n$ is the number of
+most accesses. This operation takes $O(\log n)$ time, where $n$ is the number of
 virtual memory areas. The logarithmic behavior is visible in the plots for the

### 2017-05-05 paper.tex
section 4.1.2

@@
 
-The \texttt{frag} benchmark takes the most time of any benchmark. When a region of the
-address space is being expanded, the kernel is forced to unmap it and map a new
-area elsewhere in the address space. This operation needs creation of a new
-\texttt{vm\_area\_struct}.  The kernel also must search through the red-black
-tree, though it never finds a node which can be expanded to include the new
-mapping. Then, an insertion on the tree is needed, possibly including
-rebalancing. Moreover, the existing \texttt{ vm\_area\_struct} needs to be
-removed from the tree.
+The \texttt{frag} benchmark takes the most time of any benchmark. When a region
+of the address space is being expanded, the kernel is forced to unmap it and
+map a new area elsewhere in the address space.  The kernel must search through
+the red-black tree, though it never finds a node which can be expanded to
+include the new mapping.  Then, the existing \texttt{ vm\_area\_struct} needs
+to be removed from the tree.  Finally, a new \texttt{vm\_area\_struct} is
+created and inserted into the tree.
 
 The \texttt{stride} workload is very similar to \texttt{frag} because it will also force the
-kernel to always create a new structure and insert it into the tree after
-traversing the tree first. However, it does not need to delete an existing
+kernel to always create a new structure and insert it after
+searching the tree. However, it does not need to delete an existing
 structure, which causes it to be faster than the \texttt{frag} benchmark.

### 2017-05-05 paper.tex
modified page 7

--- 2017-05-05 paper.tex
@@
     \includegraphics[width=\columnwidth]{figures/mmap_mem_usage}
-    \caption{$\Delta_i$ as discussed in section \ref{ss_vm_area_struct}. Note that
+    \caption{$\Delta_i$ adjusted with \texttt{control} for benchmarks. Note that
     graphs of \texttt{frag} and \texttt{stride} overlap. Also, \texttt{cont} and
@@
     \includegraphics[width=\columnwidth]{figures/vm_area_struct_count}
-    \caption{Number of \texttt{vm\_area\_struct}s. The graphs of
-    \texttt{control} and \texttt{cont} workloads are almost coincident with the
-    x-axis.}
+    \caption{Number of \texttt{vm\_area\_struct}s for benchmarks.}
     \label{fig:vm_area_struct_count}
@@
 
-For continuous \texttt{mmap}s the time is not only the lowest of any benchmark
-but is also invariant with the total number of \texttt{mmap}s done so far.
-There are two reasons for this.
+It can be seen that for continuous \texttt{mmap}s the time is not only the
+lowest but is also invariant with the total number of \texttt{mmap}s done so
+far. There are two reasons for this.
 
@@
 For the other workloads, a search through the red-black tree is often needed on
-most accesses. This operation takes $O(\log n)$ time, where $n$ is the number of
+most accesses. This operation takes $O(log n)$ time, where $n$ is the number of
 virtual memory areas. The logarithmic behavior is visible in the plots for the
@@
 
-The \texttt{rand} workload, on the other hand, writes to pages that are likely not
-adjacent and likely not subpages of a single huge page. As a result, a huge page
+The \texttt{rand} workload, on the other hand, writes to pages that are 
+not subpages of the same huge page. As a result, a huge page
 is allocated with almost every write, and system memory is exhausted and begins
@@
     \end{tabular}
+    \label{fig:latency_table}
 \end{figure}
@@
 
-The overall behavior is as expected. In the \texttt{rand} benchmark, as before, all
-memory is exhausted quickly, and the benchmark induces swapping soon after the
-benchmark begins. Many memory accesses thus take milliseconds to complete
+In the \texttt{rand} benchmark, as before, all
+memory is exhausted, and swapping begins very soon. Many memory accesses 
+thus take milliseconds to complete
 because they require paging in or out, which is slow due to disk access.
@@
 \ref{fig:swap_time_cdf} shows that these few accesses form a majority of run
-time; they impact performance significantly.
+time; they impact performance significantly. Figure \ref{fig:latency_table} summarizes
+this trend.
 

### 2017-05-05 paper.tex
Fix page 9

--- 2017-05-05 paper.tex
@@
 
-The \texttt{rand} workload, on the other hand, writes to pages that are 
-not subpages of the same huge page. As a result, a huge page
+The \texttt{rand} workload, on the other hand, writes to pages that are likely not
+adjacent and likely not subpages of a single huge page. As a result, a huge page
 is allocated with almost every write, and system memory is exhausted and begins
@@
     \end{tabular}
-    \label{fig:latency_table}
 \end{figure}
@@
 
-In the \texttt{rand} benchmark, as before, all
-memory is exhausted, and swapping begins very soon. Many memory accesses 
-thus take milliseconds to complete
+The overall behavior is as expected. In the \texttt{rand} benchmark, as before, all
+memory is exhausted quickly, and the benchmark induces swapping soon after the
+benchmark begins. Many memory accesses thus take milliseconds to complete
 because they require paging in or out, which is slow due to disk access.
@@
 \ref{fig:swap_time_cdf} shows that these few accesses form a majority of run
-time; they impact performance significantly. Figure \ref{fig:latency_table} summarizes
-this trend.
+time; they impact performance significantly.
 
@@
 
-As expected, the \texttt{rand} benchmark with hugepages on incurs the highest
-total overhead since it begins swapping very early. For this benchmark we
-observe an overhead of roughly 0.0002 jiffies per page memory touched after
-swapping begins. Also, kswapd runs for longer periods of time with longer
-intervals of sleep in between. This is most likely because the benchmark has
-some probability of retouching a page, so free memory is used more slowly, but
-kswapd also has to do more work to find pages that are inactive.
+In the \texttt{rand} benchmark with hugepages enabled, we observe an overhead
+of roughly 0.0002 jiffies per page of memory touched after swapping begins.
+Also, kswapd runs for longer periods of time with longer intervals of sleep in
+between. This is most likely because the benchmark has some probability of
+retouching a page, so free memory is used more slowly, but kswapd also has to
+do more work to find pages that have not been recently used.
 
-In contrast, the \texttt{cont} benchmark with hugepages disabled uses
-computational time at roughly 0.0572 jiffies per page of memory touched.
+In the \texttt{cont} benchmark with hugepages disabled, kswapd uses
+0.0572 jiffies per page of memory touched.
 kswapd runs more frequently and for shorter periods of time because new pages
-are touched very quickly, requiring faster reclamation, but because all pages
+are touched very quickly, requiring faster reclamation. But because all pages
 are only touched once, kswapd has to do very little work to choose pages to
@@
 
-Notice also that the \texttt{cont} benchmark with hugepages enabled spends
+The \texttt{cont} benchmark with hugepages enabled spends
 surprisingly little time in kswapd. We suspect that kswapd runs once and choose
@@
 wait for I/O. This seems consistent with our previous results, in which some
-operations take on the order of seconds to complete, but we have not yet
+operations take hundreds of milliseconds to complete, but we have not yet
 verified it.
@@
 swapping behavior. Moreover, we see that under memory pressure, kswapd tends to
-run more often during page faults than in its own context.
+run more often during page faults than in its own context, causing very poor
+workload performance.
 
@@
 
-The emergence of new technology has in the past often required operating systems
-to change. New non-volatile memory technologies are making it possible to have
-terabytes of memory. Recent changes to memory management in Linux demonstrate a
-lot of focus on improving the scalability of its data structures
+New memory technologies are making it possible to have
+terabytes of memory. Recent changes to memory management in Linux demonstrate
+focus on improving the scalability of its data structures
 \cite{huang2016evolutionary}.
@@
 
-RadixVM tries to overcome performance issues in highly
+RadixVM tries to mitigate performance issues in highly
 concurrent workloads due to serialization of memory management operations on
@@
 existing memory management schemes are not scalable to larger systems.
-Similarly, in our work, we examine scalability limitations of memory
-management in the Linux kernel.
 

### 2017-05-05 paper.tex
Related works

--- 2017-05-05 paper.tex
@@
 However, in our work, we study OS behavior under workloads that
-actually use all available memory, where swapping causes disk latency to become the main
+actually use all available memory and swapping causes disk latency to become the main
 performance bottleneck. Moreover, the performance overhead of simulators like
@@
 metadata and discards the contents of files. Reads to the disk causes David to
-generate data on the fly. This design choice is based on the observation that
+generate data on the fly. This design is based on the observation that
 most benchmarking frameworks do not care about the actual content of the files,
@@
 a promising direction for large-memory system studies. One may consider ignoring
-the contents of a process’s heap and only storing kernel data structures and a
-process’s code and stack segments. However, the common use of custom data
+the contents of a process's heap and only storing kernel data structures and
+process code and stack segments. However, the common use of custom data
 structures makes generating heap contents on the fly
-more difficult than generating disk. Moreover, on Linux, the kernel data structures for 1TB
+difficult. Moreover, on Linux, the kernel data structures for 1TB
 of memory may also exceed the size of physical memory on current systems
@@
 \section{Conclusion}
-Terabyte-scale memory systems are poised to become commonplace in not too distant future.
-Developers and researchers need tools to study the behavior of such systems even before the actual
-deployments are avaialable.
+Terabyte-scale memory systems are poised to become commonplace in the not too distant future.
+Developers and researchers need tools to study the behavior of such systems before they are available.
 
-In this paper, we study the scalability of memory management 
-system of Linux Kernel which we believe is the first step toward building such tools. In particular, 
-we look at the scalability of \texttt{vm\_area\_struct},  
-swapping subsystem, and the effect of hugepages across a wide range of varying parameters
+In this paper, we study the scalability of the memory management 
+system of Linux, which we believe is the first step toward building such tools. In particular, 
+we look at the scalability of \texttt{vm\_area\_struct}  
+and the swapping subsystem across a wide range of varying parameters
 and workloads.
@@
 Our findings indicate that Linux's use of
-\texttt{vm\_area\_struct} tends to be sclable in the general case, but
+\texttt{vm\_area\_struct} tends to be scalable in the general case, but
 fragmented allocation patterns can cause increased computational and memory

### 2017-05-05 paper.tex
fix page 8

--- 2017-05-05 paper.tex
@@
     \centering
+    \caption {Tail latency of memory accesses}
     \begin{tabular}{c||c|c}
@@
 Surprisingly, the majority of this latency comes from kswapd itself. kswapd goes
-through phases of moving pages to the inactive list and reclaiming pages from
-the inactive list. This repeats until the amount of free memory increases
+through phases of maintaining LRU lists and reclaiming pages from
+them. This repeats until the amount of free memory increases
 beyond a threshold. kswapd does not do any particularly expensive operations
@@
 nested loops. Meanwhile, the workload continues to use up any memory reclaimed
-by kswapd. Whenever kswapd is invoked by the page fault handler, it blocks the
-faulting process until enough pages are reclaimed.  Moreover, in some cases,
+by kswapd. Moreover, in some cases,
 during the \texttt{cont} workload kswapd chooses pages to reclaim much faster
@@
 also reduce the degree of disk locality. Thus, large swap files should be
-created with disk locality in mind. Some Linux distributions, such as
-Ubuntu, have switched to using swap files rather than swap partitions by default
-\cite{swap_ubuntu}. Our results suggest that for some workloads this may have a
-negative impact on performance if the swap file ends up in non-continuous disk
-locations.
-
+created with disk locality in mind. 
 % Mark was telling Suhas a TCP joke the other day. He had to keep repeating it slower
@@
 
-Our results also show that in some cases, kswapd induces more overhead to choose
-a good page to reclaim than would be accrued by choosing the wrong page to
-reclaim. Most of this overhead seems to be due to the poor asymptotic complexity
-of PFRA. Moreover, the swapping subsystem should be aware of disk bandwidth and
-latency, since they fundamentally limits how much memory can be reclaimed per
-unit time.
+Our results also show that in some cases, kswapd induces more overhead choosing 
+a good page to reclaim than would occur by reclaiming the wrong page. Most of this 
+overhead seems to be due to the repitition fo cheap operations millions of times.
 
 Systems with larger memory may conceivably support more memory overcommitment,
-so improving the performance of swapping may be highly relavent moving forward.
-Additionally, if a system has orders of magnitude more pages, kswapd may spend
-significantly more time scanning the active and inactive lists.
+so improving the performance of swapping may be highly relevant moving forward.
+Additionally, in systems with orders of magnitude of more pages, kswapd may spend
+significantly more time maintaining LRU lists.
 
@@
 
-In the \texttt{rand} benchmark with hugepages enabled, we observe an overhead
-of roughly 0.0002 jiffies per page of memory touched after swapping begins.
-Also, kswapd runs for longer periods of time with longer intervals of sleep in
-between. This is most likely because the benchmark has some probability of
-retouching a page, so free memory is used more slowly, but kswapd also has to
-do more work to find pages that have not been recently used.
+As expected, the \texttt{rand} benchmark with hugepages on incurs the highest
+total overhead since it begins swapping very early. For this benchmark we
+observe an overhead of roughly 0.0002 jiffies per page memory touched after
+swapping begins. Also, kswapd runs for longer periods of time with longer
+intervals of sleep in between. This is most likely because the benchmark has
+some probability of retouching a page, so free memory is used more slowly, but
+kswapd also has to do more work to find pages that are inactive.
 
-In the \texttt{cont} benchmark with hugepages disabled, kswapd uses
-0.0572 jiffies per page of memory touched.
+In contrast, the \texttt{cont} benchmark with hugepages disabled uses
+computational time at roughly 0.0572 jiffies per page of memory touched.
 kswapd runs more frequently and for shorter periods of time because new pages
-are touched very quickly, requiring faster reclamation. But because all pages
+are touched very quickly, requiring faster reclamation, but because all pages
 are only touched once, kswapd has to do very little work to choose pages to
@@
 
-The \texttt{cont} benchmark with hugepages enabled spends
+Notice also that the \texttt{cont} benchmark with hugepages enabled spends
 surprisingly little time in kswapd. We suspect that kswapd runs once and choose
@@
 wait for I/O. This seems consistent with our previous results, in which some
-operations take hundreds of milliseconds to complete, but we have not yet
+operations take on the order of seconds to complete, but we have not yet
 verified it.
@@
 swapping behavior. Moreover, we see that under memory pressure, kswapd tends to
-run more often during page faults than in its own context, causing very poor
-workload performance.
+run more often during page faults than in its own context.
 
@@
 
-New memory technologies are making it possible to have
-terabytes of memory. Recent changes to memory management in Linux demonstrate
-focus on improving the scalability of its data structures
+The emergence of new technology has in the past often required operating systems
+to change. New non-volatile memory technologies are making it possible to have
+terabytes of memory. Recent changes to memory management in Linux demonstrate a
+lot of focus on improving the scalability of its data structures
 \cite{huang2016evolutionary}.
@@
 
-RadixVM tries to mitigate performance issues in highly
+RadixVM tries to overcome performance issues in highly
 concurrent workloads due to serialization of memory management operations on
@@
 existing memory management schemes are not scalable to larger systems.
+Similarly, in our work, we examine scalability limitations of memory
+management in the Linux kernel.
 

### 2017-05-05 paper.tex
a few more changes

--- 2017-05-05 paper.tex
@@
 
-\title{Journey: A Study of Scalabilty of Memory Management in Linux}
+\title{Journey: A Study of Scalability of Memory Management in Linux}
 \author{Mark Mansi, Suhas Pai, Hasnain Ali Pirzada\\\texttt{\{markm, spai, hp\}@cs.wisc.edu}}
@@
 of memory may also exceed the size of physical memory on current systems
-\cite{simics}.
+\cite{simics}. As our results show, in some cases, the overheads of
+\texttt{vm\_area\_struct} alone could use significant amounts of memory.
 
@@
 performance measurements and make large
-benchmarks impractical.
+benchmarks impractical. Our findings indicate several areas of improvement that
+are needed before such an approach could be practical.
 
@@
 
-While priliminary, we believe that our study is a first step towards building tools 
-to study large memory systems in a principled and informed manner.
+While priliminary, we believe that our study is a first step towards building
+tools to study large memory systems in a principled and informed manner. As our
+results show, approaches similar to David \cite{david, exalt} will need to deal
+with high kernel data structure overheads. Meanwhile, virtualization-based tools
+would need to deal with the poor scalability of swapping.
 

### 2017-05-06 paper.tex
a few minor fixes

@@
 and the processor traps to a page fault handler, which allocates memory and
-sets up a page table entry for a valid access. This is called \texttt{demand paging}.
+sets up a page table entry for a valid access. This is called \textit{demand paging}.
 
@@
 in which the space is broken into a number of swap slots large enough for one
-4K page. For each of these swap spaces, the kernel keeps a swap map, which records
+4KB page. For each of these swap spaces, the kernel keeps a swap map, which records
 the usage of each swap slot so that it can find and allocate them quickly.
@@
 regular intervals and sleeps in between. Also, it is invoked if the system
-detects that memory has gone below a threshold such as during a page fault.
+detects that memory has gone below a threshold, such as during a page fault.
 kswapd will continue to run until enough pages are freed to bring the free
@@
 \begin{itemize} \item \textit{Control}. This workload is only run for the memory
-overhead experiments. It simply prints the $\Delta_i$ for each $i$ while performing nops. It's graph
+overhead experiments. It simply prints the $\Delta_i$ for each $i$ while performing nops. Its graph
 is a straight line with positive slope. This memory usage results from the
@@
     \includegraphics[width=\columnwidth]{figures/mmap_mem_usage}
-    \caption{$\Delta_i$ as discussed in section \ref{ss_vm_area_struct}. Note that
+    \caption{$\Delta_i$ as defined in section \ref{ss_vm_area_struct}. Note that
     graphs of \texttt{frag} and \texttt{stride} overlap. Also, \texttt{cont} and
@@
     \centering
-    \caption {Tail latency of memory accesses}
     \begin{tabular}{c||c|c}
@@
     \end{tabular}
-    \label{fig:latency_table}
+    \caption {Tail latency of memory accesses.\label{fig:latency_table}}
 \end{figure}
